

```{r}

# load libraries
library(magrittr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(stringr)
library(pheatmap)

# determine the server path
if (file.exists("/home/amovas/")){
    print("Remote HPC Connection!")
    wd <- "/home/amovas/data/genome-evo-proj/"
} else {
    print("Local PC Connection!")
    wd <- "/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/"
}



# Visual settings
source(paste0(wd, "src/analysis/decelerating-adaptation/plot_custom.R"))
exp_line_factor <- c("MT-2_1", "MT-2_2", "MT-4_1", "MT-4_2")
line_col_palette <- c("#ff00ff", "#ff2400", "#6600cc", "#0000ff")
names(line_col_palette) <- exp_line_factor
```

# -------------------------------------------------------------------
# Single sample LD analysis
# -------------------------------------------------------------------

```{r}


ld_df <- read.table(paste0(wd, "results/tables/ld/iii/15/360.tsv"), sep = "\t", header = TRUE)
ld_df <- ld_df[!is.na(ld_df$r2), ]  # remove rows with NA r2 values


View(ld_df)


```


````{r}

# How many pairs are in strong LD? (e.g., r2 > 0.8)
strong_ld <- sum(ld_df$r2 > 0.8, na.rm = TRUE)
ld_df[ld_df$pos2 == 8973,]
ld_df[ld_df$r2 > 0.8,]
total_pairs <- sum(!is.na(ld_df$r2))
prop_strong_ld <- strong_ld / total_pairs
prop_strong_ld


```

```{r}

# summarize by distance between SNPs

ld_df$distance <- abs(ld_df$pos2 - ld_df$pos1)

# Average r2 per distance bin (e.g., every 50 bp)

ld_summary <- ld_df %>%
    filter(!is.na(r2)) %>%
    mutate(bin = cut(distance, breaks = seq(0, max(distance), by = 50))) %>%
    group_by(bin) %>%
    summarise(mean_r2 = mean(r2), median_r2 = median(r2), n_pairs = n())


```

```{r}
# r² vs distance


ggplot(ld_df, aes(x = distance, y = r2)) +
    geom_point(alpha = 0.3) +
    geom_smooth(method = "loess", color = "blue") +
    theme_minimal() +
    labs(x = "Distance between SNPs (bp)",
        y = expression(r^2),
        title = "Linkage Disequilibrium decay with distance")

```

```{r}

# Heatmap of LD

# make matrix
ld_mat <- reshape2::acast(ld_df, pos1 ~ pos2, value.var = "r2")

# visualize

pheatmap(ld_mat, cluster_rows = FALSE, cluster_cols = FALSE,
            color = colorRampPalette(c("white", "red"))(50),
            main = "LD Heatmap (r²)")

```


```{r}

# Histogram of r²

ggplot(ld_df, aes(x = r2)) +
    geom_histogram(bins = 50, fill = "steelblue", color = "black") +
    theme_minimal() +
    labs(x = expression(r^2),
        y = "Number of SNP pairs",
        title = "Distribution of LD across SNP pairs")

```

# -------------------------------------------------------------------
# Dataset LD analysis
# -------------------------------------------------------------------



```{r}

all_ld <- data.frame()

for (exp_line in 13:16) {
    for (psg in seq(10,500,10)){
        ld_df <- read.table(paste0(wd, "results/tables/ld/iii/", exp_line, "/", psg, ".tsv"), sep = "\t", header = TRUE)
        # ld_df <- ld_df[!is.na(ld_df$r2), ]  # remove rows with NA r2 values
        ld_df$exp_line <- exp_line
        ld_df$psg <- psg
        all_ld <- rbind(all_ld, ld_df)
    }
}

all_ld %<>% mutate(distance = abs(all_ld$pos2 - all_ld$pos1),
    mid_pos = (all_ld$pos1 + all_ld$pos2) / 2,
    exp_line = case_when(exp_line==13 ~ "MT-2_1", exp_line==14 ~ "MT-2_2", exp_line==15 ~ "MT-4_1", exp_line==16 ~ "MT-4_2"), # update exp line labels
)



```


```{r LD over time}


all_ld_summary <- all_ld %>% group_by(exp_line, psg) %>%
    filter(r2 >= 0.01) %>%
    summarise(mean_r2 = mean(r2, na.rm = TRUE), 
    median_r2 = median(r2, na.rm = TRUE), 
    n_pairs = n(),
    prop_high = mean(r2 > 0.5, na.rm = TRUE)) %>%
    arrange(exp_line, psg) %>%
    mutate_at(2, as.integer)

fig <- all_ld_summary %>% ggplot(aes(x = psg, y = mean_r2, color = as.factor(exp_line))) +
    geom_line() +
    geom_point() +
    theme_minimal() +
    scale_x_continuous(breaks = seq(0, 500, by = 50)) +
    labs(x = "Transfer",
        y = expression("Mean "~(r^2)),
        title = "Mean LD across PSGs and Experimental Lines",
        color = "Population") +
    scale_color_manual(labels = exp_line_factor, values = line_col_palette) +
    custom_plot_theme(application = "publication")

# save figure
custom_plot_save(fig, "ld_per_transfer.pdf", height = 14, width = 14)



t.test(all_ld_summary$mean_r2[all_ld_summary$exp_line == 13 & all_ld_summary$psg <= 50],
    all_ld_summary$mean_r2[all_ld_summary$exp_line == 13 & all_ld_summary$psg > 50])

```

```{r LD decay loess}



all_ld_summary <- all_ld %>% group_by(exp_line, distance) %>%
    filter(!is.na(r2)) %>%
    summarise(mean_r2 = mean(r2, na.rm = TRUE), 
    median_r2 = median(r2, na.rm = TRUE), 
    n_pairs = n(),
    prop_high = mean(r2 > 0.5, na.rm = TRUE)) %>%
    arrange(exp_line, psg)

fig <- all_ld_summary %>% ggplot(aes(x = distance, y = mean_r2, color = as.factor(exp_line))) +
    geom_point() +
    geom_smooth(se = FALSE, method = "loess", span = 0.3) +  # loess smoothing
    labs(x = "Distance between SNPs (bp)",
        y = expression(mean(r^2)),
        title = "Mean LD decay with distance across Experimental Lines",
        color = "Population") +
    scale_color_manual(labels = exp_line_factor, values = line_col_palette) +
    custom_plot_theme(application = "publication")

# save figure
custom_plot_save(fig, "ld_decay_per_distance.pdf", height = 14, width = 14)



# Reshape into wide format
wide_ld <- all_ld_summary %>%
    filter(exp_line %in% c(13, 14)) %>%
    select(distance, exp_line, mean_r2) %>%
    pivot_wider(names_from = exp_line, values_from = mean_r2, names_prefix = "line_")

# Drop rows where either is NA
wide_ld <- na.omit(wide_ld)

# Paired t-test (distance is the pairing factor)
t.test(wide_ld$line_13, wide_ld$line_14, paired = TRUE)


```



```{r LD over genome}



all_ld_summary <- all_ld %>% group_by(exp_line, mid_pos) %>%
    filter(!is.na(r2)) %>%
    filter(r2 > 0.2) %>%
    summarise(mean_r2 = mean(r2, na.rm = TRUE), 
    median_r2 = median(r2, na.rm = TRUE), 
    n_pairs = n(),
    prop_high = mean(r2 > 0.5, na.rm = TRUE)) %>%
    arrange(exp_line, psg)

fig <- all_ld_summary %>% ggplot(aes(x = mid_pos, y = mean_r2, color = as.factor(exp_line))) +
    geom_point(alpha = 0.5) +
    labs(x = "\n Distance between SNPs (bp)",
        y = "r^2 > 0.1",
        title = "Mean LD decay with distance across Experimental Lines",
        color = "Line") +
    scale_color_manual(labels = exp_line_factor, values = line_col_palette) +
    scale_x_continuous(breaks = seq(0,10000, by = 400)) +
    scale_y_continuous(breaks = seq(0,1, by = 0.1)) +
    custom_plot_theme(application = "publication")

# save figure
custom_plot_save(fig, "ld_over_genome.pdf", height = 14, width = 14)



```



Hill and Weir (1988) that describes the relationship between 
linkage disequilibrium (LD) (measured as r² or other correlation coefficients) and
the physical or genetic distance between loci, often observed in a population. 
This model is used to predict how LD decays as the distance between loci increases,
reflecting the influence of genetic drift and recombination. 
The Hill-Weir model typically uses a non-linear regression to fit the observed LD decay, 
which can then be used to infer population parameters like effective population size (Ne). 






```{r LD decay basic model}

# bring data into suitable format

all_ld_summary <- all_ld %>% group_by(exp_line, distance) %>%
    filter(!is.na(r2)) %>%
    summarise(mean_r2 = mean(r2, na.rm = TRUE), 
    median_r2 = median(r2, na.rm = TRUE), 
    n_pairs = n(),
    prop_high = mean(r2 > 0.5, na.rm = TRUE)) %>%
    arrange(exp_line, psg) %>%
    mutate(
    host = ifelse(exp_line %in% c("MT-2_1","MT-2_2"), "MT2", "MT4"),
    pop_size_est = case_when(exp_line== "MT-2_1" ~ 10^4.37, exp_line== "MT-2_2" ~ 10^4.37, exp_line== "MT-4_1" ~ 10^5.06, exp_line== "MT-4_2" ~ 10^5.06)) # update exp line labels


# Define the Hill-Weir model formula
model_formula <- as.formula("mean_r2 ~ 1 / (1 + (2* r * pop_size_est * (distance)^(0.7)))")




# Function to fit Hill–Weir and return predicted values
fit_hw_fun <- function(df) {
    fit <- tryCatch(
        nls(model_formula,
            data = df,
            start = list(r = 1e-4)),
        error = function(e) NULL
    )
    
    if (!is.null(fit)) {
        df$fit_r2 <- predict(fit, newdata = df)
        df$r_est  <- coef(fit)["r"]
    } else {
        df$fit_r2 <- NA
        df$r_est  <- NA
    }
    return(df)
}

# define function to extract fit statistics
get_fit_stats <- function(fit, df) {
    preds <- predict(fit, newdata = df)
    obs   <- df$mean_r2
    
    ss_res <- sum((obs - preds)^2, na.rm = TRUE)   # residual sum of squares
    ss_tot <- sum((obs - mean(obs, na.rm = TRUE))^2, na.rm = TRUE)  # total sum of squares
    
    r2 <- 1 - (ss_res / ss_tot)
    
    data.frame(
        r2 = r2,
        RMSE = sqrt(mean((obs - preds)^2, na.rm = TRUE)),
        AIC = AIC(fit),
        r_est = coef(fit)[["r"]]
    )
}


# get the modelling results
fit_results <- all_ld_summary %>%
    mutate(host = factor(host, levels = c("MT2", "MT4"))) %>%
    group_by(host) %>%
    group_map(~ {
        fit <- nls(model_formula,
                data = .x,
                start = list(r = 1e-4))
        get_fit_stats(fit, .x)
    }) %>%
    bind_rows(.id = "host") %>% 
    mutate(host = case_when(host == "1" ~ "MT2", host == "2" ~ "MT4"),
        label = paste0(
            "\n Est. recombination rate \n per nucl. per gen. = ", signif(r_est, 3),
            "\n Goodness of fit (R²) = ", round(r2, 2)
        )
    ) 



# choose annotation positions (adjust as needed)
fit_results <- fit_results %>%
    mutate(xpos = 150,   # distance value where label appears
            ypos = c(0.15, 0.10))  # y position for MT2 / MT4 separately

# make the figure
fig <- ggplot(all_ld_fits, aes(x = distance, y = mean_r2, color = host)) +
    geom_point(alpha = 0.5) +
    geom_line(aes(y = fit_r2), size = 1.2) + 
    geom_text(data = fit_results,
        aes(x = xpos, y = ypos, label = label, color = host),
        inherit.aes = FALSE,
        hjust = 0, size = 8, fontface = "bold", show.legend = FALSE ) +
    labs(x = "\n Distance between SNPs (bp)",
        y = expression(mean(r^2)),
        title = "Mean LD decay with distance across Experimental Lines",
        color = "Line") +
    scale_color_manual(values = c("#ff8b08", "#00bbff"),) +
    custom_plot_theme(application = "publication")

# save the figure
custom_plot_save(fig, "ld_decay_per_distance_model.pdf", height = 14, width = 14)

```

