


Prepare annotated table of fixated variants for karin:

```{r}

# first read in the annotated variant data from appropriate script



head(variants_ann)
for (virus_line in unique(variants_ann$exp_line)){
  print(virus_line)
  for (mut in unique(variants_ann$mut_info[variants_ann$mut_cat == "Fixated" & variants_ann$exp_line == virus_line])){
      emergence_psg <- min(variants_ann$passage[variants_ann$exp_line == virus_line & variants_ann$mut_info == mut & variants_ann$allele_freq >= 0.01])
      fixation_psg <- min(variants_ann$passage[variants_ann$exp_line == virus_line & variants_ann$mut_info == mut & variants_ann$allele_freq >= 0.99])

      variants_ann$emergence_passage[variants_ann$mut_info == mut & variants_ann$exp_line == virus_line] <- emergence_psg
      variants_ann$fixation_passage[variants_ann$mut_info == mut & variants_ann$exp_line == virus_line] <- fixation_psg
  }

}
variants_ann_fixated <- 
            variants_ann %>%
            filter(allele_freq >= 0.99) %>%
            filter(genomic_pos != 3086) %>%
            mutate(allele_freq = round(allele_freq,3))





variants_ann_fixated <- variants_ann_fixated[,-match(c("chrom", "loss_of_function", "nonsense_mediated_decay", "mut_3_context", "mut_5_context", "mut_info"), colnames(variants_ann))]
variants_ann_fixated <- variants_ann_fixated[,c("exp", "host_line", "exp_line", "passage", "coverage", "variant_id", "genomic_pos", "ref_allele", 
                                                "alt_allele", "mut_type", "allele_freq", "mut_cat", "feature_id", "feature", "feature_range", "ORF", "annotation_id", "impact", "effect", "aa_change", "loc_in_cds", "loc_in_protein", "mut_context", "emergence_passage", "fixation_passage")]

sum(variants_ann_fixated$coverage < 500)
datatable(variants_ann, filter = "bottom")

write.table(variants_ann_fixated, file = "/Users/alimos313/Desktop/fixated_variants_ann_2.csv", sep = ",", row.names = F)




```


fixated variants that are missed by different coverage thresholds

Based on the analysis below, there are mainly to fixated-label mutations that are missed due to coverage below 
threshold 500, both in the end of virus sample from line 15 passage 500. (positions 9521 and 9558).

Another observation, specifically for mutation at position 483 in line 15 and passage 530, is that when the coverage is low,
a fixated mutation's allele frequency arguably has more variations and might fall under the 0.99 threshold. For this mutation
and in this line this phenomenon happens multiple times. On the other side from passage 370 to 400 it seems to be
correctly identified as majority as the coverage is above 2000 but allele freq around 0.93. One solution to address this
issue could be setting 0.95 as fixated threshold. Should be tested and decided upon.

Last but not least, doing this check another problem came to my attention and that is the mutations that happen to be in the regions that
later on in viral evolution will be removed due to LD or small deletions. Here two points are noteworthy:

1. First, the LD are not called based on the current mutation calling script. Therefore, while they exist 
almost homogeneously in certain populations (), they are not called by the pipeline and are absent in the variant tables.

2. Some mutations will inevitably disappear from the population due to small or large deletions. When comparing the rate of mutation disappearance
I have to keep that in mind. 


```{r}


# read in the data
variants <- read.table(file = "/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/tables/2-p/all_variants.csv.gz", 
sep = ",", stringsAsFactors = FALSE, header = TRUE)



variants_fixated <- 
            variants %>%
            filter(fraction >= 0.95) %>%
            filter(start != 3086) 



View(variants_fixated[variants_fixated$coverage < 500 & ((variants_fixated$start >= 454 & variants_fixated$start <= 550) | (variants_fixated$start >= 9500 & variants_fixated$start <= 9626)),])
View(variants_fixated[variants_fixated$start == 483 & variants_fixated$line == 15,][order(variants_fixated$passage[variants_fixated$start == 483 & variants_fixated$line == 15]),])



```



## Genetic drift simulation



```{r}

population_size <- 100
transfer_size <- population_size/2
starting_freq <- 50/100
generation_number <- log2(population_size/(population_size-transfer_size))

nr_line <- 5
nr_passage <- 200

allele_freq_records <- list()

set.seed(625)
i <- 0
while (i < nr_line){
  i <- i + 1
  print(i)
  
  alleles <- c(rep("A", times = population_size*starting_freq), rep("B", times = population_size*(1-starting_freq)))
  allele_freq_record <- c(starting_freq)

  j <- 1
  while (j < nr_passage){
    j <- j + 1
    alleles <- sample(alleles, size = transfer_size, replace = F)
    if ("A" %in% alleles){
      alleles <- rep(alleles, each = 2^generation_number)
      current_freq <- as.numeric((table(alleles)/population_size)["A"])
      allele_freq_record <- append(allele_freq_record, current_freq)
    } else {
      allele_freq_record <- append(allele_freq_record, 0)
    }
  }
  
  allele_freq_records <- append(allele_freq_records, list(allele_freq_record))
}


allele_freq_records_df <- as.data.frame(allele_freq_records)
colnames(allele_freq_records_df) <- 1:nr_line
allele_freq_records_df$passage <- 1:nr_passage

allele_freq_records_df2 <- gather(allele_freq_records_df, key = "line", value = "allele_freq", 1:nr_line)

fig <- allele_freq_records_df2 %>% 
          ggplot(aes(x = passage, y = allele_freq, color = line)) +
          geom_line(size=2) +
          labs(x="\nPassage", y= "Allele Frequency\n", color = "Virus Line", title = paste0("Simulation of Genetic Drift (Population Size: ", population_size,")\n"))+
          scale_y_continuous(breaks = seq(0,1,0.1), limits = c(0,1)) +
          geom_hline(yintercept = 0.5, color = "black", size = 2) +
          theme_bw() +
          theme(plot.title = element_text(size = 50, hjust = 0.5),
          axis.title = element_text(size = 45),
          axis.text.x = element_text(size = 40, angle = 90),
          axis.text.y = element_text(size = 40),
          legend.title = element_text(size=35),
          legend.text = element_text(size = 30))



png(paste0("/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/figs/presentation/301123/genetic_drift_sim_popsize_", population_size, ".png")
    , width = 1500, height = 1500)
fig
dev.off()


### ========================================================================================================================================================

# What are the chances of a new mutation with starting allele frequency of 1/2N to reach fixation? How long will it take for all potential positions to have the chance to reach fixation?


mu <- 3/10000
genome_length <- 10000




population_size <- 50000
transfer_size <- population_size/4
starting_freq <- 1/100
generation_number <- log2((population_size/transfer_size))

nr_cycles <- population_size - transfer_size

nr_mutations_per_genome_per_generation <- nr_cycles * mu


likelihood_of_fixation <- 1/(population_size)


nr_passage  <- genome_length/(likelihood_of_fixation * nr_mutations_per_genome_per_generation)











```






```{r G>A in their context}

threshold <- 0.01
experiment <- "EXPIII"


mapping_vec <- c(A = "T", T = "A", C = "G", G = "C")

variants_sub <- variants %>%
                    filter(allele_freq >= threshold) %>%
                    filter(exp == experiment)


for (i in 1:nrow(variants_sub)){
    if (variants_sub$ref_allele[i] %in% c("G", "A")){
        context <- variants_sub$mut_context[i]
        rev_context <- paste0(as.character(mapping_vec[str_sub(context,3,3)]),
                                        as.character(mapping_vec[str_sub(context,2,2)]), 
                                        as.character(mapping_vec[str_sub(context,1,1)]))
        variants_sub$mut_context_6[i] <- rev_context
    } else {
        variants_sub$mut_context_6[i] <- variants_sub$mut_context[i]
    }
}
head(variants_sub)

variants_sub2 <- variants_sub %>%
                    filter((ref_allele == "G" & alt_allele == "A") | (ref_allele == "C" & alt_allele == "T")) %>% 
                    select(genomic_pos, exp_line, exp, mut_context_6) %>%
                    distinct() %>%
                    group_by(exp_line) %>%
                    count(mut_context_6) %>%
                    dplyr::rename(count = n) %>%
                    mutate(perc = count/sum(count))



variants_sub2 %>% ggplot(aes(x = mut_context_6, y = perc)) +
        geom_bar(stat = "identity") +
        facet_wrap(~exp_line)


variants_sub3 <- variants_sub %>%
                    filter((ref_allele == "G" & alt_allele == "A")) %>% 
                    select(genomic_pos, exp_line, exp, mut_3_context) %>%
                    distinct() %>%
                    group_by(exp_line) %>%
                    count(mut_3_context) %>%
                    dplyr::rename(count = n) %>%
                    mutate(perc = count/sum(count))

variants_sub3 %>% ggplot(aes(x = mut_3_context, y = perc)) +
        geom_bar(stat = "identity") +
        facet_wrap(~exp_line)

```


