---
title:  
    "Shannon Entropy"
date: 
    "Feb 15th, 2024"
author:  
    "Ali Movasati"  
output:  
    html_document:
        toc: true
        toc_depth: 3
        toc_float: true
        number_sections: true
        theme: lumen
        highlight: zenburn
        code_folding: hide
---




```{css, echo=F}

#TOC {
  position: fixed;
  left: 0;
  top: 0;
  width: 250px;
  height: 100%;
  overflow:auto;
}

body, td {
  font-family: sans-serif;
  background-color: white;
  font-size: 16px;
}


h1, .h1, h2, .h2, h3, .h3 {
    margin-top: 30px;
}


p.caption {
  font-size: 0.9em;
  font-style: italic;
  color: grey;
  margin-right: 10%;
  margin-left: 10%;  
  text-align: justify;
}


/* scrollable code chunks, limit code chunk window height */

pre {
  max-height: 600px;
  overflow-y: auto;
}

pre[class] {
  max-height: 300px;
}


/* scrollable output of code chunks */
/* add class.output="scroll-100" to code chunks */

.scroll-100 {
  max-height: 300px;
  overflow-y: auto;
  background-color: inherit;
}
```



```{r setting global code chunk parameters, include=FALSE}

# options for code chunk and output inclusion 
knitr::opts_chunk$set(
  message = FALSE, echo = TRUE, warning=FALSE, results = "hide"
)

# options for figures
knitr::opts_chunk$set(
  fig.width = 8, fig.height = 8, fig.align = "center"
)


# options(scipen=999)

# margin of plotly
m <- list(
  l = 150,
  r = 150,
  b = 100,
  t = 100,
  pad = 4
)


```


```{r load libraries and set variables}

library(tidyr)
library(magrittr)
library(dplyr)
library(ggplot2)
library(plotly)



line_col_palette <- c("#ff00ff", "#ff2400", "#6600cc", "#0000ff", "#66ff66", "#009900", "#00cc99", "#558000")
names(line_col_palette) <- c("MT-2_1", "MT-2_2", "MT-4_1", "MT-4_2", "MT-2_i", "MT-2_ii", "MT-2_iii", "MT-2_iv")
```


# Introduction

The Shannon entropy is a standard measure for the order state of symbol sequences, such as, for example, DNA sequences. By calculating Shannon entropy of sample sequencing data we can measure the 
genomic diversity within each population. 

Shannon entropy would take a value from 0 (when there is no variation at a site) to 2 (when the frequency
of each four bases are equal at a site). Therefore, the more heterogenous a site is the higher its entropy. 


```{r read in the data}


lines <- c(13:20)

shannon_df <- data.frame()

for (line in lines){
    shannon_df <- rbind(shannon_df, read.table(file = paste0("/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/tables/misc/", line, "_shannon_entropy.tsv"), sep = "\t", header = T))
}

# set the correct viral line names and  add host info

shannon_df %<>% mutate(host_line = ifelse(exp_line %in% c("13", "14","17", "18", "19", "20"), "MT2", "MT4")) %>% # add a field containing pos, ref, and alt info of muts
                mutate(exp_line = case_when(exp_line==13 ~ "MT-2_1", exp_line==14 ~ "MT-2_2", exp_line==15 ~ "MT-4_1", exp_line==16 ~ "MT-4_2",
                                            exp_line==17 ~ "MT-2_i", exp_line==18 ~ "MT-2_ii", exp_line==19 ~ "MT-2_iii", exp_line==20 ~ "MT-2_iv"))

```


> First let's take a look at total and mean entropy per sample. 


```{r entropy per sample, results=TRUE}




d <- shannon_df %>% group_by(experiment,host_line,exp_line,passage) %>% summarise (sum_ent = sum(shannon_entropy, na.rm = T), mean_ent = mean(shannon_entropy, na.rm = T), .groups = 'drop')

d %<>% mutate(exp_line = factor(exp_line, levels = names(line_col_palette)))

fig <- ggplot(d, aes(x = passage, y = sum_ent, color = exp_line)) +
  geom_line(linewidth=2) + 
  labs(x="\nTransfer", y= "Total Entropy\n", color = "Virus Line", title = paste0("Sum of Entropy per Sample")) +
  scale_color_manual(values = line_col_palette) +
  theme_bw() +
        theme(plot.title = element_text(size = 25, hjust = 0.5),
        axis.title = element_text(size = 20),
        axis.text.x = element_text(size = 15, angle = 90),
        axis.text.y = element_text(size = 15))

ggplotly(fig)


fig <- ggplot(d, aes(x = passage, y = mean_ent, color = exp_line)) +
  geom_line(linewidth=2) + 
  labs(x="\nTransfer", y= "Mean\n", color = "Virus Line", title = paste0("Mean of Entropy per Sample")) +
  scale_color_manual(values = line_col_palette) +
  theme_bw() +
        theme(plot.title = element_text(size = 20, hjust = 0.5),
        axis.title = element_text(size = 15),
        axis.text.x = element_text(size = 10, angle = 90),
        axis.text.y = element_text(size = 10),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 10))

ggplotly(fig)


```


As it is evident from the plots, they very very similar to minority plots generated before. However,
with these plots we can better see that diversity for viral lines MT-4_1 and MT-4_2 is also reaching its
saturation level after passage 500.


> Now let's take a look at total and mean entropy per position. 


```{r entropy per position, results=TRUE}

dd <- shannon_df %>% group_by(experiment,host_line,exp_line,genomic_position) %>% summarise (sum_ent = sum(shannon_entropy, na.rm = T), mean_ent = mean(shannon_entropy, na.rm = T), var_ent = var(shannon_entropy, na.rm = T),.groups = 'drop')


fig <- ggplot(dd, aes(x=genomic_position, y = mean_ent)) + facet_wrap(~exp_line, ncol=4) +
    geom_point() +
    labs(x="\nPosition", y= "Entropy Mean \n", title = paste0("Mean of Entropy per Genomic Position")) +
    theme_bw() +
        theme(plot.title = element_text(size = 20, hjust = 0.5),
        axis.title = element_text(size = 15),
        axis.text.x = element_text(size = 10, angle = 90),
        axis.text.y = element_text(size = 10))



ggplotly(fig)

```



> What can be more interesting is to look at is the site entropy differences over time. Below you'll see the mean and variance of entropy changes from one sample to the next.



```{r entropy change per position, results=TRUE}


dd <- shannon_df %>% group_by(experiment,host_line,exp_line,genomic_position) %>% summarise (sum_ent = sum(shannon_entropy, na.rm = T), mean_ent = mean(abs(diff(shannon_entropy)), na.rm = T), var_ent = var(abs(diff(shannon_entropy)), na.rm = T),.groups = 'drop')

dd %<>% mutate(exp_line = factor(exp_line, levels = names(line_col_palette)))



fig <- ggplot(dd, aes(x=genomic_position, y = mean_ent)) + facet_wrap(~exp_line, ncol=4) +
    geom_point() +
    labs(x="\nPosition", y= "Mean of Entropy Difference\n", title = paste0("Mean of Entropy Difference per Genomic Position")) +
    theme_bw() +
        theme(plot.title = element_text(size = 25, hjust = 0.5),
        axis.title = element_text(size = 20),
        axis.text.x = element_text(size = 15, angle = 90),
        axis.text.y = element_text(size = 15),
        strip.text = element_text(size = 20))




ggplotly(fig)


fig <- ggplot(dd, aes(x=genomic_position, y = var_ent)) + facet_wrap(~exp_line, ncol=4) +
    geom_point() +
    labs(x="\nPosition", y= "Variance of Entropy Difference\n", title = paste0("Variance of Entropy Difference per Genomic Position")) +
    theme_bw() +
        theme(plot.title = element_text(size = 20, hjust = 0.5),
        axis.title = element_text(size = 15),
        axis.text.x = element_text(size = 10, angle = 90),
        axis.text.y = element_text(size = 10))


ggplotly(fig)



```


What can be observed from these data is that overall the rate of entropy change for 3 and 5 prime untranslated regions is higher. Moreover,


Additionally, despite lower entropy in EXPIV viral lines, we see higher variance of entropy changes in those lines.




> let's calculate the correlation coefficients (Spearman's rho) for mean difference of entropy levels for each position.

!! The reason I chose Spearman's rho over Pearson's r is that the data is highly skewed and zero-inflated.
In such scenario's Pearson's r will be biased, unless the data is appropriately transformed.
An alternative is to use a rank-based correlation coefficient (such as Spearman's rho or Kendall's tau),
which will be a a subjective transformation for the data. 


```{r distribution of mean entropy difference, results=TRUE}


fig <- dd[dd$exp_line == "MT-2_1",] %>% ggplot(aes(x=mean_ent)) +
          geom_histogram(aes(y = (..count..)/sum(..count..)), binwidth=0.01) +
          scale_y_continuous(labels = scales::percent) +
          labs(x="\nMean Entropy Difference", y= "Percentage\n", title = "Distribution of Mean Entropy Difference for MT-2_1") +
          theme_bw() +
                theme(plot.title = element_text(size = 20, hjust = 0.5),
                axis.title = element_text(size = 15),
                axis.text.x = element_text(size = 10, angle = 90),
                axis.text.y = element_text(size = 10))


ggplotly(fig)


```

```{r}

cor_coefs <- data.frame()

for (i in levels(dd$exp_line)){
  for (j in levels(dd$exp_line)[levels(dd$exp_line) != i]){
    cor_res <- cor.test(dd$mean_ent[dd$exp_line == i], dd$mean_ent[dd$exp_line == j], method = 'spearman', exact = FALSE)
    cor_coefs <- rbind(cor_coefs, c(i , j, cor_res$estimate))
    print(paste0("Spearman's Rho between ", i, " and ", j, " is ", cor_res$estimate, " and p.value is ", cor_res$p.value))
  }
}


cor_coefs <- as.data.frame(cor_coefs) %>% 
                        `colnames<-`(c("line1", "line2", "Rho")) %>%
                        mutate(line1 = factor(line1, levels = names(line_col_palette))) %>%
                        mutate(line2 = factor(line2, levels = names(line_col_palette))) %>%
                        mutate(Rho = round(as.numeric(Rho), 2))




plt <- ggplot(data = as.data.frame(cor_coefs), aes(x=line1, y=line2, fill=100*Rho)) + 
        geom_tile() +
        geom_text(aes(line1, line2, label = Rho), color = "black", size = 4) +
        labs(title = paste0("Spearman's Rho Coorelation Coefficient for Positional Entropy")) +
        scale_fill_gradient2(low = "white", high = "red", 
        midpoint = 1, limit = c(0,100), space = "Lab", 
        name="Shared %") +
        theme(
        axis.title = element_blank(),
        plot.title = element_text(size = 20, hjust = 0.5)
        )

plt

```

```{r test repeatability of high entropy positions in the sequence}


# fig <- plot_ly(as.data.frame(dd), x = ~genomic_position, y = ~sum_ent, 
#             type = 'scatter') + facet_grid(exp_line ~ .)
# fig <- fig %>% layout(title = paste0("Sum of Entropy per Sample") ,
#                       xaxis = list(title = 'Passage'),
#                       yaxis = list(title = 'Sum'))

# fig



# plot(x = d$passage, y = d$sum_ent, col = d$exp_line)
# plot(x = d$passage, y = d$mean_ent, col = d$exp_line)

# line <- 15
# plot(dd$genomic_position[dd$exp_line == line], dd$mean_ent[dd$exp_line == line])

# a <- dd$genomic_position[dd$exp_line == 13 & dd$mean_ent >= 0.4]

# b <- dd$genomic_position[dd$exp_line == 14 & dd$mean_ent >= 0.4]

# c <- dd$genomic_position[dd$exp_line == 15 & dd$mean_ent >= 0.1]

# d <- dd$genomic_position[dd$exp_line == 16 & dd$mean_ent >= 0.1]

# length(intersect(c,d))

# length(unique(c(c,d)))

# aa <- sample(seq(1,9000), 1161)
# bb <- sample(seq(1,9000), 1161)

# length(intersect(aa,bb))


# den <- density(shannon_df$shannon_entropy[shannon_df$passage == 500], na.rm =T)
# plot(den)

# length(shannon_df$shannon_entropy[shannon_df$passage == 570])


# shannon_df$shannon_entropy[shannon_df$passage == 90]

```