---
title:  
    "Overview of mutation composition of EXPIII"
date: 
    "Mar 25th, 2024"
author:  
    "Ali Movasati"  
output:  
    html_document:
        #css: style.css
        toc: true
        toc_depth: 2
        toc_float: true
        number_sections: true
        theme: lumen
        highlight: zenburn
        code_folding: hide
        code_download: true # to make the Rmd source code available
---


```{css, echo=F}

#TOC {
  position: fixed;
  left: 0;
  top: 0;
  width: 250px;
  height: 100%;
  overflow:auto;
}

body, td {
  font-family: sans-serif;
  background-color: white;
  font-size: 16px;
}


h1, .h1, h2, .h2, h3, .h3 {
    margin-top: 30px;
}


p.caption {
  font-size: 0.9em;
  font-style: italic;
  color: grey;
  margin-right: 10%;
  margin-left: 10%;  
  text-align: justify;
}


/* scrollable code chunks, limit code chunk window height */

pre {
  max-height: 600px;
  overflow-y: auto;
}

pre[class] {
  max-height: 300px;
}


/* scrollable output of code chunks */
/* add class.output="scroll-100" to code chunks */

.scroll-100 {
  max-height: 300px;
  overflow-y: auto;
  background-color: inherit;
}
```



```{r setting global code chunk parameters, include=FALSE}

# options for code chunk and output display 
knitr::opts_chunk$set(
  message = FALSE, echo = TRUE, warning=FALSE, results = "hide"
)

# options for figure display
knitr::opts_chunk$set(
  fig.width = 8, fig.height = 8, fig.align = "center"
)


# options(scipen=999)

# margin of plotly
m <- list(
  l = 150,
  r = 150,
  b = 100,
  t = 100,
  pad = 4
)



```




# Introduction

In this document I will characterize the mutational profile that we observe in EXPIII evolutionary lines. One part of the result section of the manuscript will be dedicated to mutation statistics. I will try
to compose those paragraphs in this document using in-text coding feature of R Markdown. My goal is to later copy-paste the whole paragraphs from this document to the manuscript.


```{r loading libraries, echo=FALSE}

# loading libraries
library("ggplot2")
library("plotly")
library("gganimate")
library("DT")
library("dplyr")
library("tidyr")
library("Biostrings")
library("stringr")
library("magrittr")
library("hrbrthemes")

```


```{r functions and operators}
`%notin%` <- Negate(`%in%`)
```


```{r read-in the data}

# determine the working directory path based on server

if (file.exists("/home/amovas/")){
  print("Remote HPC Connection!")
  wd <- "/home/amovas/data/genome-evo-proj/"
} else {
  print("Local PC Connection!")
  wd <- "/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/"
}

variants_ann <- read.table(file = paste0(wd, "results/tables/pipeline-outputs/all_annotated_variants.tsv.gz"), 
sep = "\t", stringsAsFactors = FALSE, header = TRUE)


# only keep the EXPIII data
exp <- "EXPIII"
variants_ann_expiii <- variants_ann[variants_ann$exp_line %in% 13:16,]


# correct for the switched experimental lines
variants_ann_expiii$exp_line[variants_ann_expiii$exp_line == 15 & variants_ann_expiii$passage > 360] <- "A"
variants_ann_expiii$exp_line[variants_ann_expiii$exp_line == 16 & variants_ann_expiii$passage > 360] <- "B"

variants_ann_expiii$exp_line[variants_ann_expiii$exp_line == "A"] <- "16"
variants_ann_expiii$exp_line[variants_ann_expiii$exp_line == "B"] <- "15"


# facors

host_factor <- c("MT2", "MT4")
exp_line_factor <- c("MT-2_1", "MT-2_2", "MT-4_1", "MT-4_2")

translational_impact_factor <- c("U", "S", "N")
gene_factor <- c("gag", "pol", "vif", "vpr", "tat", "rev", "vpu", "envelope", "nef")
feature_factor <- c("5utr", "gag", "pol", "vif", "vpr", "tat1", "rev1", "vpu", "envelope", "tat2", "rev2", "nef", "3utr")


# format the dataframe
variants_ann_expiii <- variants_ann_expiii %>%
                            mutate(host_line = ifelse(exp_line %in% c("13", "14"), "MT2", "MT4"), # set host cell line info
                                exp_line = case_when(exp_line==13 ~ "MT-2_1", exp_line==14 ~ "MT-2_2", exp_line==15 ~ "MT-4_1", exp_line==16 ~ "MT-4_2"), # update exp line labels
                                mut_info = paste(exp_line, genomic_pos, ref_allele, alt_allele, sep = "_")) %>% 
                            mutate(effect_simplified = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
                            mutate(effect_simplified = replace_na(effect_simplified, "N")) %>%
                            mutate(exp_line = factor(exp_line, levels = exp_line_factor),
                            host_line = factor(host_line, levels = host_factor),
                            effect_simplified = factor(effect_simplified, levels = translational_impact_factor)) 

# remove duplicate records for the same variants
variants_expiii <- variants_ann_expiii %>% filter(!duplicated(variant_id))


# last passage 

end_passage <- 560


# Visual settings


if (exp == "EXPIII"){
    line_palette <- c("#ff00ff", "#ff2400", "#6600cc", "#0000ff")
    breaks_step <- 50
    line_labels <- c("MT-2_1", "MT-2_2", "MT-4_1", "MT-4_2")
} else if ((exp == "EXPIV")) {
    line_palette <- c("#66ff66", "#009900", "#00cc99", "#558000")
    breaks_step <- 5
    line_labels <- c("MT-2_i", "MT-2_ii", "MT-2_iii", "MT-2_iv")
}


```




## Allele Frequency distribution

Let's first take a look at the allele frequency distribution of all mutations.



```{r, results=TRUE}


gg_fig <- ggplot(variants_expiii[variants_expiii$allele_freq >= 0.01,], aes(x=allele_freq)) + 
    geom_histogram(aes(y=..count..),      # Histogram with density instead of count on y-axis
                binwidth=.01,
                colour="black", fill="white") +
    geom_density(aes(y = ..count.. * .01), alpha=.3, fill="#FF6666") +
    facet_wrap(~exp_line, ncol = 2) +
    transition_states(passage, transition_length = 1, state_length = 1, wrap = TRUE) + # to create animation 
    labs(title = "Allele Freq. Distribution \n(Passage : {closest_state})", x = "Allele Freq.", y = "Count") +
    theme_bw() +
    theme(
    plot.title = element_text(size = 25, hjust = 0.5),
    axis.title = element_text(size = 20),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    strip.text.x = element_text(size = 25),
    legend.title = element_text(size=15),
    legend.text = element_text(size = 10),
    plot.margin = unit(c(1,1,1,1), "cm"))



gg_fig

if (!file.exists(paste0(wd, "results/figs/gif/allele-freq-accumulation.gif"))){
  anim_save(gg_fig, file = paste0(wd, "results/figs/gif/allele-freq-accumulation.gif"))
}


# in-line variables
a <- 100*nrow(variants_expiii[variants_expiii$allele_freq >= 0.01 & variants_expiii$allele_freq < 0.05 & variants_expiii$passage == end_passage,])/
nrow(variants_expiii[variants_expiii$allele_freq >= 0.01 & variants_expiii$passage == end_passage,])

b <- 100*nrow(variants_expiii[variants_expiii$allele_freq >= 0.95 & variants_expiii$passage == end_passage,])/
nrow(variants_expiii[variants_expiii$allele_freq >= 0.01 & variants_expiii$passage == end_passage,])


```


In the beginning of the experiment we observe a skewed, uni-modal distribution of variants. As time passes,
some of the emerged variants find the chance to increase in their frequencies, either derived by positive selection or genetic drift. After a while we observe a bi-modal
distribution of the variants, with more than `r round(a, digits=1)`% of variants having an allele frequency of below 5% and `r round(b, digits=1)`% having AF of
above 95% at passage `r end_passage`.


## variants in numbers

```{r overall variant stats}

# in-line variables
tot_pos_min <- length(unique(variants_expiii$genomic_pos[variants_expiii$allele_freq >= 0.05 & variants_expiii$allele_freq < 0.5]))
tot_pos_maj <- length(unique(variants_expiii$genomic_pos[variants_expiii$allele_freq >= 0.5 & variants_expiii$allele_freq < 1.01]))


# read in ancestral reference sequence 
fastaFile <- readDNAStringSet(paste0(wd,"data/reference/provirus/provirus-consensus/hiv_provirus_ref_genome.fasta"))
seq_name = names(fastaFile)
sequence = paste(fastaFile)

# add subtype info
variants_expiii_sub <-  variants_expiii %>% mutate(mut_type_sub = ifelse(mut_type == "P", paste0(ref_allele,">",alt_allele), mut_type))


```

```{r majority variant stats}

## Majority stats
min_threshold <- 0.5
max_threshold <- 1.01

# prepare data for visualization
mut_profile_maj_unique <- variants_expiii_sub %>% filter(allele_freq >= min_threshold & allele_freq < max_threshold) %>%
                            select(mut_info, mut_type_sub) %>%
                            distinct() %>%
                            group_by(mut_type_sub) %>%
                            summarise(count = n())


# in-text variables
tot_var_maj <- sum(mut_profile_maj_unique$count)
tot_var_maj_snv <- sum(mut_profile_maj_unique$count[mut_profile_maj_unique$mut_type_sub %notin% c("I", "D")])
tot_var_maj_ins <- sum(mut_profile_maj_unique$count[mut_profile_maj_unique$mut_type_sub %in% "I"])
tot_var_maj_del <- sum(mut_profile_maj_unique$count[mut_profile_maj_unique$mut_type_sub %in% "D"])
tot_var_maj_G2A <- sum(mut_profile_maj_unique$count[mut_profile_maj_unique$mut_type_sub %in% "G>A"])
tot_var_maj_C2T <- sum(mut_profile_maj_unique$count[mut_profile_maj_unique$mut_type_sub %in% "C>T"])
tot_var_maj_T2C <- sum(mut_profile_maj_unique$count[mut_profile_maj_unique$mut_type_sub %in% "T>C"])

```

```{r minority variant stats}
## Minority stats
min_threshold <- 0.05
max_threshold <- 0.5

# prepare data for visualization
mut_profile_min_unique <- variants_expiii_sub %>% filter(allele_freq >= min_threshold & allele_freq < max_threshold) %>%
                            select(mut_info, mut_type_sub) %>%
                            distinct() %>%
                            group_by(mut_type_sub) %>%
                            summarise(count = n())


# in-text variables
tot_var_min <- sum(mut_profile_min_unique$count)
tot_var_min_snv <- sum(mut_profile_min_unique$count[mut_profile_min_unique$mut_type_sub %notin% c("I", "D")])
tot_var_min_ins <- sum(mut_profile_min_unique$count[mut_profile_min_unique$mut_type_sub %in% "I"])
tot_var_min_del <- sum(mut_profile_min_unique$count[mut_profile_min_unique$mut_type_sub %in% "D"])
tot_var_min_G2A <- sum(mut_profile_min_unique$count[mut_profile_min_unique$mut_type_sub %in% "G>A"])
tot_var_min_C2T <- sum(mut_profile_min_unique$count[mut_profile_min_unique$mut_type_sub %in% "C>T"])
tot_var_min_T2C <- sum(mut_profile_min_unique$count[mut_profile_min_unique$mut_type_sub %in% "T>C"])


```


In all four viral lines, we recorded `r tot_pos_min` &  `r tot_pos_maj` positions of the ancestral genome of length `r nchar(sequence)` as minority and majority mutations, respectively. 

Out of `r tot_var_maj` unique majority mutations, `r tot_var_maj_snv` (`r round(100*(tot_var_maj_snv/tot_var_maj), digits=1)`%) are single substitutions, while small insertions and deletions make up
`r round(100*(tot_var_maj_ins/tot_var_maj), digits=1)`% and `r round(100*(tot_var_maj_del/tot_var_maj), digits=1)`% of unique majority mutations, respectively. 
Point mutations are distributed unevenly among different substitution types, with G>A substitutions constituting `r round(100*(tot_var_maj_G2A/tot_var_maj), digits=1)`% of SNVs, 
followed by C>T (`r round(100*(tot_var_maj_C2T/tot_var_maj), digits=1)`%) and T>C (`r round(100*(tot_var_maj_T2C/tot_var_maj), digits=1)`%) (refer to pie-chart).


Out of `r tot_var_min` unique minority mutations, `r tot_var_min_snv` (`r round(100*(tot_var_min_snv/tot_var_min), digits=1)`%) are single substitutions, while small insertions and deletions make up
`r round(100*(tot_var_min_ins/tot_var_min), digits=1)`% and `r round(100*(tot_var_min_del/tot_var_min), digits=1)`% of unique minority mutations, respectively. 
Point mutations are distributed unevenly among different substition types, with G>A substitutions constituting `r round(100*(tot_var_min_G2A/tot_var_min), digits=1)`% of SNVs, 
followed by C>T (`r round(100*(tot_var_min_C2T/tot_var_min), digits=1)`%) and T>C (`r round(100*(tot_var_min_T2C/tot_var_min), digits=1)`%) (refer to pie-chart).



## Mutational composition

Despite the substitution type imbalance, the patterns seems to be well-preserved throughout the experiment and among the 4 evolutionary lines (refer to the figure). 
This indicates that the source of mutagenesis is operating similarly between the two hosts. In addition when we look at the spectra of newly appeared mutations (refer to the figure), we see that they maintain very similar profiles over time,
indicating that the source of mutagenesis is remaining unchanging throughout the experiment.

Therefore, even though we see that environment impacts the kinetics of mutation accumulation in HIV-1 genome, environment does not substantially effect the source of mutations.


```{r min mutation spectrum, results=TRUE}

## Minority stats
min_threshold <- 0.05
max_threshold <- 0.5

# prepare data for visualization
counts <- variants_expiii %>%
            filter(allele_freq >= min_threshold & allele_freq < max_threshold) %>%
            filter(mut_type=="P") %>%
            mutate(mut_subtype=paste0(ref_allele, ">", alt_allele)) %>%
            group_by(exp_line, passage, mut_subtype) %>%
            count() %>%
            dplyr::rename(count=n) %>%
            group_by(exp_line,passage) %>%
            mutate(pct =  100 *count/sum(count)) %>%
            ungroup() %>%
            mutate_at(2:3, as.factor)

# visualize using ggplot
fig <- as.data.frame(counts) %>% ggplot(aes(x=passage, y = pct, fill = mut_subtype)) +
                                geom_bar(position="stack", stat="identity") + 
                                facet_wrap(~exp_line) +
                                labs(x="Passage", y= "Percentage", fill = "Base Substitution", title = "Substitution Spectra of Accumulating Variants")+
                                theme_bw() +
                                theme(plot.title = element_text(size = 25, hjust = 0.5),
                                axis.title = element_text(size = 20),
                                axis.text.x = element_blank(),
                                axis.text.y = element_text(size = 10),
                                strip.text.x = element_text(size = 25),
                                legend.title = element_text(size=15),
                                legend.text = element_text(size = 10))


fig

```



```{r prepare difference dataframe for majority, results=TRUE}

## Majority stats
min_threshold <- 0.05
max_threshold <- 0.5


passages <- seq(10,570, 10)
diff_df <- data.frame()



for (i in 1:56){
    for (exp_line in c("MT-2_1", "MT-2_2", "MT-4_1", "MT-4_2")){
        prev_in <- passages[i]
        pres_in <- passages[i+1]

        prev <- variants_expiii[variants_expiii$exp_line == exp_line & variants_expiii$passage == prev_in & variants_expiii$allele_freq >= min_threshold & variants_expiii$allele_freq < max_threshold,c("genomic_pos", "ref_allele", "alt_allele", "mut_type", "exp_line")]

        pres <- variants_expiii[variants_expiii$exp_line == exp_line & variants_expiii$passage == pres_in & variants_expiii$allele_freq >= min_threshold & variants_expiii$allele_freq < max_threshold,c("genomic_pos", "ref_allele", "alt_allele", "mut_type", "exp_line")]

        addition <- dplyr::setdiff(pres, prev)
        nrow(prev)

        diff_df <- rbind(diff_df, cbind(addition, passage = passages[i+1]))

    }
    
}


counts <- diff_df %>%
            filter(mut_type=="P") %>%
            mutate(mut_subtype=paste0(ref_allele, ">", alt_allele)) %>%
            group_by(exp_line, passage, mut_subtype) %>%
            count() %>%
            dplyr::rename(count=n) %>%
            group_by(exp_line,passage) %>%
            mutate(pct =  100 *count/sum(count)) %>%
            ungroup() %>%
            mutate_at(2:3, as.factor)


fig <- as.data.frame(counts) %>% ggplot(aes(x=passage, y = pct, fill = mut_subtype)) +
                                geom_bar(position="stack", stat="identity") + 
                                facet_wrap(~exp_line) +
                                labs(x="Passage", y= "Percentage", fill = "Base Substitution", title = "Substitution Spectra of Newly Added Variants")+
                                theme_bw() +
                                theme(plot.title = element_text(size = 25, hjust = 0.5),
                                axis.title = element_text(size = 20),
                                axis.text.x = element_blank(),
                                axis.text.y = element_text(size = 10),
                                strip.text.x = element_text(size = 25),
                                legend.title = element_text(size=15),
                                legend.text = element_text(size = 10))


fig

```



## Fixation dynamics

```{r draw af of once fixed mutations, results=TRUE}

tot_pos_fixed <- length(unique(variants_expiii$genomic_pos[variants_expiii$allele_freq >= 0.99 & variants_expiii$allele_freq < 1.01]))
tot_fixed <- length(unique(paste0(variants_expiii$genomic_pos[variants_expiii$allele_freq >= 0.99 & variants_expiii$allele_freq < 1.01],
variants_expiii$ref_allele[variants_expiii$allele_freq >= 0.99 & variants_expiii$allele_freq < 1.01],
variants_expiii$alt_allele[variants_expiii$allele_freq >= 0.99 & variants_expiii$allele_freq < 1.01])))

fixed_muts <- unique(variants_expiii$mut_info[variants_expiii$allele_freq >= 0.99])

tot_fixed_4lines <- length(fixed_muts)


fig <- variants_expiii[variants_expiii$mut_info %in% fixed_muts,] %>% ggplot(aes(x = passage, y = allele_freq, group = mut_info, color = mut_info)) +
  geom_line() +
  facet_wrap(~exp_line, ncol = 1) +
  theme_bw() +
  theme(plot.title = element_text(size = 25, hjust = 0.5),
  axis.title = element_text(size = 20),
  axis.text.x = element_blank(),
  axis.text.y = element_text(size = 10),
  strip.text.x = element_text(size = 25),
  legend.position="none")

fig

pdf(file = paste0(wd, "results/figs/pdf/frequency_dynamics_fixed_mutations.pdf"), width = 56, height = 21)
fig
dev.off()

```

Overall `r tot_pos_fixed` positions (`r tot_fixed` unique mutations) throughout the HIV-1 genome has undergone a fixation at some point in the experiment. Number of recorded fixed mutations in all 4 lines is `r tot_fixed_4lines`.



```{r, results=TRUE}

# record at what passages the fixed mutations first appeared and reached fixation (AF above 0.99)
df <- data.frame()

for (exp_line in line_labels){
  fixed_muts <- unique(variants_expiii$mut_info[variants_expiii$exp_line == exp_line & variants_expiii$allele_freq >= 0.99])
  for (mut in fixed_muts){
    appearance <- min(variants_expiii$passage[variants_expiii$mut_info == mut & variants_expiii$allele_freq >= 0.01])
    fixation <- min(variants_expiii$passage[variants_expiii$mut_info == mut & variants_expiii$allele_freq >= 0.99])
    translational_impact <- unique(as.character(variants_expiii$effect_simplified[variants_expiii$mut_info == mut]))
    df <- rbind(df, c(exp_line, mut, translational_impact, appearance, fixation, fixation-appearance))
  }
}

# set column names and right data types
df %<>% set_colnames(c("exp_line", "mut_info", "effect_simplified", "appearance", "fixation", "time_to_fixation")) %>% 
        mutate_at(4:6, as.integer) %>%
        mutate(exp_line = factor(exp_line, levels = exp_line_factor),
        effect_simplified = factor(effect_simplified, levels = translational_impact_factor))


# view the frequency dynamics of one fixed mutation
df[df$mut_info == "MT-2_1_566_G_A",]

fig <- variants_expiii[variants_expiii$mut_info %in% "MT-2_1_566_G_A",] %>% ggplot(aes(x = passage, y = allele_freq)) +
  geom_line(group = 1) +
  theme(legend.position="none")

fig

# summarize time to fixation per passage of appearance
df_summary <- df %>% group_by(exp_line, appearance) %>%
                    summarize(Mean = mean(time_to_fixation, na.rm=TRUE),
                    n = n())


# density of time_to_fixation distributions
fig <- df_summary %>% ggplot(aes(x = Mean, color = exp_line)) +
                    geom_density() +
                    theme_ipsum()

fig

# get inline variables
a <- min(df$time_to_fixation)
b <- nrow(df[df$time_to_fixation==10,])


# visualize time to fixation summary
fig <- df_summary %>% ggplot(aes(x = appearance, y = Mean, group = exp_line, color = exp_line)) +
  geom_line() +
  geom_point(aes(size = n)) +
  facet_wrap(~exp_line, ncol = 1) 
  
fig

# summarize time to fixation per passage of appearance
df_summary <- df %>% group_by(effect_simplified, appearance) %>%
                    summarize(Mean = mean(time_to_fixation, na.rm=TRUE),
                    n = n())


# density of time_to_fixation distributions
fig <- df_summary %>% ggplot(aes(x = Mean, color = effect_simplified)) +
                    geom_density() +
                    theme_ipsum()

fig



```



Overall we see fixed mutations that appeared earlier in the experiment have a shorter time-to-fixation
compared to late-emerged mutations. For `r b` mutations it only took them 10 or fewer passages to emerge
and reach fixation.

```{r, results=TRUE}
# number of once fixated mutations that fall under 50% af later in the experiment

df2 <- data.frame()
variants_expiii_del <- data.frame()

for (exp_line in line_labels){
  fixed_muts <- unique(variants_expiii$mut_info[variants_expiii$exp_line == exp_line & variants_expiii$allele_freq >= 0.99])
  for (mut in fixed_muts){
    appearance <- min(variants_expiii$passage[variants_expiii$mut_info == mut & variants_expiii$allele_freq >= 0.01])
    fixation <- min(variants_expiii$passage[variants_expiii$mut_info == mut & variants_expiii$allele_freq >= 0.99])
    deleted <- seq(fixation,570,10)[seq(fixation,570,10) %notin% variants_expiii$passage[variants_expiii$mut_info == mut & variants_expiii$passage >= fixation]]
    deleted_bool <- length(deleted) > 0
    for (d in deleted){
      cc <- variants_expiii[variants_expiii$mut_info == mut & variants_expiii$passage == fixation,]
      cc$passage <- d
      cc$allele_freq <- NA
      variants_expiii_del <- rbind(variants_expiii_del, cc)
    }
    disappearance <- min(variants_expiii$passage[variants_expiii$mut_info == mut & variants_expiii$allele_freq < 0.8 & variants_expiii$passage > fixation])
    disappearance <- ifelse(is.infinite(disappearance), 0, disappearance)
    df2 <- rbind(df2, c(exp_line, mut, appearance, fixation, fixation-appearance,deleted_bool,disappearance))
  }
}
colnames(df2) <- c("exp_line", "mut_info", "appearance", "fixation", "time_to_fixation", "deleted", "reversion")


variants_expiii1<- rbind(variants_expiii, variants_expiii_del)
dddd <- df2$mut_info[df2$reversion != 0]

jjjj <- df2$mut_info[df2$deleted == "TRUE"]

fig <- variants_expiii[variants_expiii$mut_info %in% dddd,] %>% ggplot(aes(x = passage, y = allele_freq, group = mut_info)) +
  geom_line() +
  facet_wrap(~exp_line, ncol = 1) +
  theme(legend.position="none")

ggplotly(fig)




fig <- variants_expiii[variants_expiii$mut_info %in% "MT-4_1_9154_A_T",] %>% ggplot(aes(x = passage, y = allele_freq)) +
  geom_line(group = 1) +
  geom_point() +
  facet_wrap(~exp_line, ncol = 1) +
  theme(legend.position="none")

fig
```



```{r}


# for each fixed mutations compute the fold increase in allele frequency from the passage they appeared to the next 10th passage
head(df2)

df2 <- data.frame()


for (exp_line in line_labels){
  fixed_muts <- unique(variants_expiii$mut_info[variants_expiii$exp_line == exp_line & variants_expiii$allele_freq >= 0.5])
  for (mut in fixed_muts){
    appearance <- min(variants_expiii$passage[variants_expiii$mut_info == mut & variants_expiii$allele_freq >= 0.01])

    af_appearance <- variants_expiii$allele_freq[variants_expiii$mut_info == mut & variants_expiii$passage == appearance]
    af_next <- variants_expiii$allele_freq[variants_expiii$mut_info == mut & variants_expiii$passage == appearance+20]
    df2 <- rbind(df2, c(exp_line, mut, appearance, ifelse(length(af_next/af_appearance) == 0, NA, af_next/af_appearance)))
  }
}

colnames(df2) <- c("exp_line", "mut_info", "appearance", "fold_inc")

head(df2)

df_summary2 <- df2 %>%
                    mutate_at(4, as.numeric) %>%
                    group_by(exp_line, appearance) %>%
                    summarize(Mean = mean(fold_inc, na.rm=TRUE),
                    n = n()) %>%
                    mutate(appearance = factor(appearance, levels = seq(10,570,10)))


fig <- df_summary2 %>% ggplot(aes(x = appearance, y = Mean, group = exp_line, color = exp_line)) +
  geom_line() +
  geom_point(aes(size = n)) +
  facet_wrap(~exp_line, ncol = 1) 
  
fig
```