---
title:  
    "Jeffrey Barrick et al. 2009 Analysis"
date: 
    "Nov 15th, 2023"
author:  
    "Ali Movasati"  
output:  
    html_document:
        toc: true
        toc_depth: 3
        toc_float: true
        number_sections: true
        theme: lumen
        highlight: zenburn
        code_folding: hide
---




```{r setting global chunk parameters, include=FALSE}

# options for code chunk and output inclusion 
knitr::opts_chunk$set(
  message = FALSE, echo = TRUE, warning=FALSE, results = "hide"
)

# options for figures
knitr::opts_chunk$set(
  fig.width = 8, fig.height = 8, fig.align = "center"
)

# options(scipen=999)

# margin of plotly
m <- list(
  l = 150,
  r = 150,
  b = 100,
  t = 100,
  pad = 4
)


```




```{r loading libraries, echo=FALSE}

# loading libraries
library("plotly")
library("ggplot2")
library("DT")
library("dplyr")
library("tidyr")
library("Biostrings")
library("stringr")
library("magrittr")
library("ggvenn")
library("VennDiagram")
library("ggVennDiagram")


```


# Introduction


In this document I will try to reproduce some of the analysis provided and ideas presented by Jeffrey Barrick et al. (2019).

[Jeffrey Barrick](https://www.nature.com/articles/nature08480)




```{r read-in the data}

# read in the data
variants_ann <- read.table(file = "/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/tables/pipeline-outputs/all_annotated_variants.tsv.gz", 
sep = "\t", stringsAsFactors = FALSE, header = TRUE)


# account for the switched experimental lines
variants_ann$exp_line[variants_ann$exp_line == 15 & variants_ann$passage > 360] <- "A"
variants_ann$exp_line[variants_ann$exp_line == 16 & variants_ann$passage > 360] <- "B"

variants_ann$exp_line[variants_ann$exp_line == "A"] <- "16"
variants_ann$exp_line[variants_ann$exp_line == "B"] <- "15"



variants_ann$exp_line[variants_ann$exp_line == 18 & variants_ann$passage > 350] <- "A"
variants_ann$exp_line[variants_ann$exp_line == 19 & variants_ann$passage > 350] <- "B"

variants_ann$exp_line[variants_ann$exp_line == "A"] <- "19"
variants_ann$exp_line[variants_ann$exp_line == "B"] <- "18"


## prepare data

variants_ann %<>% 
                filter(coverage >= 500) %>%
                mutate(exp = case_when(exp_line %in% 13:16 ~ "EXPIII", exp_line %in% 17:20 ~ "EXPIV")) %>% # add experiment number info
                mutate(host_line = ifelse(exp_line %in% c("13", "14","17", "18", "19", "20"), "MT2", "MT4")) %>% # add a field containing pos, ref, and alt info of muts
                mutate(mut_info = paste(genomic_pos, ref_allele, alt_allele, sep = "_")) %>% # add host cell line info
                mutate(exp_line = case_when(exp_line==13 ~ "MT-2_1", exp_line==14 ~ "MT-2_2", exp_line==15 ~ "MT-4_1", exp_line==16 ~ "MT-4_2",
                                            exp_line==17 ~ "MT-2_i", exp_line==18 ~ "MT-2_ii", exp_line==19 ~ "MT-2_iii", exp_line==20 ~ "MT-2_iv")) %>%
                mutate(mut_cat = case_when((allele_freq < 0.01) ~ "Unreliable",
                                        (allele_freq >= 0.01 & allele_freq < 0.05) ~ "Sporadic", 
                                        (allele_freq >= 0.05 & allele_freq < 0.5) ~ "Minority",
                                        (allele_freq >= 0.5 & allele_freq < 0.99) ~ "Majority", 
                                        (allele_freq >= 0.99) ~ "Fixated"))

variants <- variants_ann %>% filter(!duplicated(variant_id))


# set the ggplot variables

gg_var_list <- list(EXPIII = list(line_palette = c("#ff00ff", "#ff2400", "#6600cc", "#0000ff"), break_step = 50),
                    EXPIV = list(line_palette = c("#68FF00", "#22E9AA", "#0DD100", "#009163"), break_step = 5))

gg_size_list <- list(ggplot = list(title.size = 50, axis.title.size = 45, axis.text.size = 40, legend.title.size = 35, legend.text.size = 30, strip.text.x.size = 35),
                    plotly = list(title.size = 25, axis.title.size = 20, axis.text.size = 10, legend.title.size = 20, legend.text.size = 15, strip.text.x.size = 20))

# reordering the variant dataframe

# variants_ann <- variants_ann[,-match(c("chrom", "loss_of_function", "nonsense_mediated_decay", "mut_3_context", "mut_5_context", "mut_info"), colnames(variants_ann))]
# variants_ann <- variants_ann[,c("exp", "host_line", "exp_line", "passage", "variant_id", "genomic_pos", "ref_allele", "alt_allele", "mut_type", "allele_freq", "mut_cat", "feature_id", "feature", "feature_range", "ORF", "annotation_id", "impact", "effect", "aa_change", "loc_in_cds", "loc_in_protein", "mut_context")]


# facors

exp_line_factor <- c("MT-2_1", "MT-2_2", "MT-4_1", "MT-4_2", "MT-2_i", "MT-2_ii", "MT-2_iii", "MT-2_iv")
mut_cat_factor <- c("Sporadic", "Minority", "Majority", "Fixated")

gene_factor <- c("gag", "pol", "vif", "vpr", "tat", "rev", "vpu", "envelope", "nef")
feature_factor <- c("5utr", "gag", "pol", "vif", "vpr", "tat1", "rev1", "vpu", "envelope", "tat2", "rev2", "nef", "3utr")


```



In this paper, four line of evidence is suggested to reject the drift hypothesis outlined by the author. The first evidence suggests that under the drift hypothesis we would expect to see higher number of synonymous than
non-synonymous mutations. First let's plot the number of these mutations types for each experiment and each mutation category.

## Raw data 

Let's first take a look at the accumulation of synonymous and nonsynonymous mutations over time.

```{r mutation accumulation plot stratified by translational impact, results=T}


plt_type <- "plotly"

for (experiment in unique(variants$exp)){

    counts <- variants %>%
        filter(exp == experiment) %>%
        mutate(effect_simplified = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
        mutate(effect_simplified = replace_na(effect_simplified, "N")) %>%
        filter(effect_simplified != "U") %>%
        group_by(exp_line, passage, mut_cat, effect_simplified) %>%
        count() %>%
        mutate(host_line = ifelse(exp_line %in% c("MT-4_1","MT-4_2"), "MT4", "MT2")) %>%
        dplyr::rename(mut_count = n) %>%
        mutate_at(c(1,3,4,6), as.factor) %>%
        mutate(mut_cat = factor(mut_cat, levels = c("Sporadic", "Minority", "Majority", "Fixated"))) %>%
        mutate_at(c(2,5), as.numeric) 
    
    
    fig <- counts %>%  
        ggplot(aes(x=passage, y = mut_count, color = exp_line, linetype=effect_simplified)) +
        geom_line(size=1) +
        facet_wrap(~mut_cat, scales = "free")+
        labs(x="\nPassage", y= "Mutation Count\n", title = paste0("Mutation Accumulation in ", experiment, "\n"))+
        scale_color_manual(name = "Virus Line", values = gg_var_list[[experiment]][["line_palette"]]) +
        scale_linetype_manual(name = "Translational Effect", , values = c("dashed", "solid")) +
        theme_bw() +
        theme(plot.title = element_text(size = gg_size_list[[plt_type]][["title.size"]], hjust = 0.5),
        axis.title = element_text(size = gg_size_list[[plt_type]][["axis.title.size"]]),
        axis.text.x = element_text(size = gg_size_list[[plt_type]][["axis.text.size"]], angle = 90),
        axis.text.y = element_text(size = gg_size_list[[plt_type]][["axis.text.size"]]),
        legend.title = element_text(size=gg_size_list[[plt_type]][["legend.title.size"]]),
        legend.text = element_text(size = gg_size_list[[plt_type]][["legend.text.size"]]))

    print(fig)

    # save pngs

    pdf(paste0("/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/figs/pdf/jeffrey-barrick/syn_vs_nonsyn_", experiment, ".pdf")
    , width = 21, height = 21)
    print(fig)
    dev.off()
}

```


## Characterization of HIV-1 genome 

Let's see what are the chances of a point mutation in the protein-coding regions of HIV-1 to be synonymous vs nonsynonymous
and subsequently normalize all these counts base on that.


```{r number of all possible mutations + percentage of synonymous ones, results=T}

# read in ancestral reference sequence 
fastaFile <- readDNAStringSet("/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/data/reference/plasmid/hiv_plasmid_ref_genome.fasta")
seq_name = names(fastaFile)
sequence = paste(fastaFile)


# read in feature coordinate data

features <- read.table(file = "/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/tables/pipeline-outputs/cds_feature_list.tsv", sep = "\t", header = T)

# by looking at the features list:

features2 <- data.frame(name = c("gag", "pol", "vif", "vpr", "tat1", "rev1", "vpu", "envelope", "tat2", "rev2", "nef"),
                        start = c(790, 2085, 5041, 5559, 5830, 5969, 6061, 6221, 8369+2-1, 8369+3-1, 8787),
                        end = c(2292, 5096, 5619, 5849, 6044-2, 6044-1, 6306, 8785, 8414, 8643, 9407))

# check
# (features2$end-features2$start+1)/3



# make objects containing bases and aas
bases <- c("A", "T", "C", "G")
aa <- unique(GENETIC_CODE)


gene_summary <- data.frame()
aa_summary <- data.frame()

for (gene in features2$name){ 
    # counters and value containers
    syn_count <- 0
    nonsyn_count <- 0
    aa_ref_composition <- c()
    aa_alt_composition <- c()

    start_pos <- features2$start[features2$name == gene]
    end_pos <- features2$end[features2$name == gene]
    gene_length <- end_pos - start_pos + 1
    codon_count <- gene_length/3

    # get the codons
    codons <- substring(substring(sequence,start_pos, end_pos), seq(1, end_pos-start_pos+1, 3), seq(3, end_pos-start_pos+1, 3))

    # iterate over the codons
    for (i in 1:length(codons)){
        
        ref_codon = codons[i]
        # GENETIC_CODE from biostrings
        ref_aa <- GENETIC_CODE[[ref_codon]]

        aa_ref_composition <- append(aa_ref_composition,ref_aa)

        # iterate over possible mutations for each codon
        for (j in 1:3){
            mutation <- bases != str_sub(ref_codon, start = j, end = j)
            for (k in bases[mutation]){
                
                alt_codon <- ref_codon
                substring(alt_codon,j, j) <- k
                alt_aa <- GENETIC_CODE[[alt_codon]]
                
                if (alt_aa == ref_aa){
                    syn_count <- syn_count + 1
                } else {
                    nonsyn_count <- nonsyn_count + 1
                    aa_alt_composition <- append(aa_alt_composition,alt_aa)
                }
            }   
        }
    } 
    all_count <- syn_count + nonsyn_count
    gene_summary <- rbind(gene_summary, c(gene, gene_length, syn_count, syn_count/all_count, nonsyn_count, nonsyn_count/all_count, all_count))
    
    # counts amino acids in reference and their possibilities to be produced
    aa_ref_composition <- factor(aa_ref_composition, levels = aa)
    aa_ref_composition_tbl <- table(aa_ref_composition)

    aa_alt_composition <- factor(aa_alt_composition, levels = aa)
    aa_alt_composition_tbl <- table(aa_alt_composition)

    aa_summary <- rbind(aa_summary, c(gene, "ref_composition", as.integer(aa_ref_composition_tbl)))
    aa_summary <- rbind(aa_summary, c(gene, "alt_composition", as.integer(aa_alt_composition_tbl)))
}

colnames(gene_summary) <- c("gene", "gene_length", "synonymous_count", "synonymous_percentage", "nonsynonymous_count", "nonsynonymous_percentage", "all_count")
colnames(aa_summary) <- c("gene", "mode", aa)



datatable(gene_summary, filter = "bottom", caption = "Gene-level summary of Positional Synonymity")


gene_summary %<>% mutate_at(2:ncol(gene_summary), as.numeric)



datatable(as.data.frame(colMeans(gene_summary[,c(4,6)])), filter = "bottom", caption = "Genome-level summary of Positional Synonymity", colnames = "Mean")



aa_summary %<>% 
    mutate_at(3:ncol(aa_summary), as.numeric)

aa_summary$sum <- rowSums(aa_summary[,3:ncol(aa_summary)])



mypaste <- function(x,y) paste(x, "(%", y, ")", sep="")

# percentage of aa throughout the genome
datatable(as.data.frame(round(colMeans(round(aa_summary[,3:(ncol(aa_summary)-1)]/aa_summary[,ncol(aa_summary)]*100, 1)),2)),
        filter = "bottom", caption = "Genome_level Amino Acid Composition", colnames = "Percent")


# percentage and count of aa per gene
aa_summary[,3:(ncol(aa_summary)-1)] <- mapply(mypaste, aa_summary[,3:(ncol(aa_summary)-1)], round(aa_summary[,3:(ncol(aa_summary)-1)]/aa_summary[,ncol(aa_summary)]*100, 1))

datatable(aa_summary, filter = "bottom", caption = "Gene_level Amino Acid Composition")



```




## dN/dS Analysis


### Per Gene Plot

In this analysis I try to obtain and comapre the dN/dS measurements for each gene separately. We would like to see how they change over time, how they compare between the two environments or the two experiments.

!! Due to high number of plots view them as pdf in the designated direcotry. !!
```{r gene dn/ds,results=T}

# plot the difference of synonymous and non-synonymous mutations
plt_type <- "plotly"


# prepare the synonymous and nonsynonymous likelihoods data frame for processing
gene_summary %<>%
            select(!c("synonymous_percentage", "nonsynonymous_percentage"))


gene_summary <- 
            rbind(gene_summary, 
            c("tat", colSums(gene_summary[gene_summary$gene %in% c("tat1", "tat2"),2:ncol(gene_summary)])),
            c("rev", colSums(gene_summary[gene_summary$gene %in% c("rev1", "rev2"),2:ncol(gene_summary)])))

gene_summary <- gene_summary[!(gene_summary$gene %in% c("tat1", "tat2", "rev1", "rev2")),]

gene_summary %<>%
        mutate(gene = factor(gene, levels = gene_factor)) %>%
        mutate_at(2:ncol(gene_summary), as.numeric)


# iterate over genes and experiments and plot

for (gene in gene_summary$gene){
    
    
    syn_pos <- gene_summary$synonymous_count[gene_summary$gene == gene]
    nonsyn_pos <- gene_summary$nonsynonymous_count[gene_summary$gene == gene]

    for (experiment in unique(variants$exp)){
        counts <- variants %>%
            mutate(mut_cat = factor(mut_cat, levels = mut_cat_factor)) %>%
            filter(exp == experiment) %>%
            mutate(effect_simplified = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
            mutate(effect_simplified = replace_na(effect_simplified, "N")) %>%
            filter(feature == gene) %>%
            mutate(effect_simplified = factor(effect_simplified, levels = c("N", "S"))) %>%
            group_by(exp_line, passage, mut_cat, effect_simplified, .drop=F) %>%
            count() %>%
            dplyr::rename(mut_count = n) %>%
            mutate(host_line = ifelse(exp_line %in% c("MT-4_1","MT-4_2"), "MT4", "MT2")) %>%
            mutate(exp_line = factor(exp_line, levels = exp_line_factor)) %>%
            mutate(effect_simplified = factor(effect_simplified, levels = c("N", "S"))) %>%
            mutate(host_line = factor(host_line, levels = c("MT2", "MT4")))
        
        counts_plt <- 
            counts %>%
            spread(key = effect_simplified, mut_count) %>%
            mutate(S = case_when(S == 0 ~ S+0.01 , S !=0 ~ S)) %>%
            mutate(`dN/dS` = log((N/nonsyn_pos)/(S/syn_pos))) 
        
        counts_plt$`dN/dS`[is.infinite(counts_plt$`dN/dS`) | is.nan(counts_plt$`dN/dS`)] <- 0
        
        
        fig <- counts_plt %>%  
            ggplot(aes(x=passage, y = `dN/dS`, color = exp_line)) +
            geom_line(size=1) +
            facet_wrap(~mut_cat, scales = "free")+
            labs(x="\nPassage", y= "Log2(dN/dS)\n", color = "Virus Line", title = paste0("dN/dS Over Time\n in ", experiment, " for ", gene, "\n"))+
            scale_color_manual(values = gg_var_list[[experiment]][["line_palette"]]) +
            scale_y_continuous() +
            geom_hline(yintercept = 0, color = "black", size = 2) +
            theme_bw() +
            theme(plot.title = element_text(size = gg_size_list[[plt_type]][["title.size"]], hjust = 0.5),
            axis.title = element_text(size = gg_size_list[[plt_type]][["axis.title.size"]]),
            axis.text.x = element_text(size = gg_size_list[[plt_type]][["axis.text.size"]], angle = 90),
            axis.text.y = element_text(size = gg_size_list[[plt_type]][["axis.text.size"]]),
            legend.title = element_text(size=gg_size_list[[plt_type]][["legend.title.size"]]),
            legend.text = element_text(size = gg_size_list[[plt_type]][["legend.text.size"]]))

        # save pdfs

        pdf(paste0("/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/figs/pdf/jeffrey-barrick/dn_ds_", experiment, "_", gene, ".pdf")
        , width = 21, height = 21)
        print(fig)
        dev.off()
    }
}





```






### Per Gene Statistical Significance

#### positive vs purifying selection in each virus line

```{r, results=T}

t_test_res_all <- data.frame()
for (gene in gene_summary$gene){
    
    
    syn_pos <- gene_summary$synonymous_count[gene_summary$gene == gene]
    nonsyn_pos <- gene_summary$nonsynonymous_count[gene_summary$gene == gene]

    for (experiment in unique(variants$exp)){
        counts <- variants %>%
            mutate(mut_cat = factor(mut_cat, levels = mut_cat_factor)) %>%
            filter(exp == experiment) %>%
            mutate(effect_simplified = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
            mutate(effect_simplified = replace_na(effect_simplified, "N")) %>%
            filter(feature == gene) %>%
            mutate(effect_simplified = factor(effect_simplified, levels = c("N", "S"))) %>%
            group_by(exp_line, passage, mut_cat, effect_simplified, .drop=F) %>%
            count() %>%
            dplyr::rename(mut_count = n) %>%
            mutate(host_line = ifelse(exp_line %in% c("MT-4_1","MT-4_2"), "MT4", "MT2")) %>%
            mutate(exp_line = factor(exp_line, levels = exp_line_factor)) %>%
            mutate(effect_simplified = factor(effect_simplified, levels = c("N", "S"))) %>%
            mutate(host_line = factor(host_line, levels = c("MT2", "MT4"))) %>%
            spread(key = effect_simplified, mut_count) %>%
            mutate(S = case_when(S == 0 ~ S+0.01 , S !=0 ~ S)) %>%
            mutate(`dN/dS` = log((N/nonsyn_pos)/(S/syn_pos)))
         
        
        counts$`dN/dS`[is.infinite(counts$`dN/dS`) | is.nan(counts$`dN/dS`)] <- 0
        
        for (mut_cat in mut_cat_factor){
            for (host_line in unique(counts$host_line)){
                dnds <- counts$`dN/dS`[counts$mut_cat == mut_cat & counts$host_line == host_line]
                
                t_test_res <- t.test(dnds, mu = 0, alternative = "two.sided")
                
                t_test_res_all <- rbind(t_test_res_all, c(experiment, host_line, gene, mut_cat, round(mean(dnds),3), t_test_res$p.value))
            
        }
        }
        

    }
}

colnames(t_test_res_all) <- c("exp", "host", "gene", "mut_cat", "mean(dN/dS)", "p-value")

datatable(t_test_res_all, filter = "bottom")




```



#### dN/dS level tests between each virus line group

```{r, results = T}





t_test_res_all <- data.frame()
for (gene in gene_summary$gene){
    print(gene)
    
    syn_pos <- gene_summary$synonymous_count[gene_summary$gene == gene]
    nonsyn_pos <- gene_summary$nonsynonymous_count[gene_summary$gene == gene]

    
    counts <- variants %>%
        mutate(mut_cat = factor(mut_cat, levels = mut_cat_factor)) %>%
        mutate(effect_simplified = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
        mutate(effect_simplified = replace_na(effect_simplified, "N")) %>%
        filter(feature == gene) %>%
        mutate(effect_simplified = factor(effect_simplified, levels = c("N", "S"))) %>%
        filter(exp != "EXPIII" | passage <= 510) %>%
        mutate(passage = factor(passage)) %>%
        group_by(exp_line, passage, mut_cat, effect_simplified, .drop=F) %>%
        count() %>%
        dplyr::rename(mut_count = n) %>%
        mutate(host_line = ifelse(exp_line %in% c("MT-4_1","MT-4_2"), "MT4", "MT2")) %>%
        mutate(exp_line = factor(exp_line, levels = exp_line_factor)) %>%
        mutate(effect_simplified = factor(effect_simplified, levels = c("N", "S"))) %>%
        mutate(host_line = factor(host_line, levels = c("MT2", "MT4"))) %>%
        spread(key = effect_simplified, mut_count) %>%
        mutate(S = case_when(S == 0 ~ S+0.01 , S !=0 ~ S)) %>%
        mutate(`dN/dS` = log((N/nonsyn_pos)/(S/syn_pos)))
    
    
    counts$`dN/dS`[is.infinite(counts$`dN/dS`) | is.nan(counts$`dN/dS`)] <- 0
    
    # counts_deformed[counts_deformed$exp_line == "III_MT-2" & counts_deformed$passage == 30 & counts_deformed$mut_cat == "Sporadic",]
    
    counts_deformed <- counts %>% 
                select(!c("N", "S")) %>%
                mutate(exp_line = case_when(exp_line %in% exp_line_factor[1:2] ~ "III_MT-2",
                                            exp_line %in% exp_line_factor[3:4] ~ "III_MT-4",
                                            exp_line %in% exp_line_factor[4:8] ~ "IV_MT-2")) %>%
                mutate(exp_line = factor(exp_line, levels = c("III_MT-2", "III_MT-4", "IV_MT-2"))) %>%
                select(-"host_line") %>%
                group_by(exp_line, passage, mut_cat) %>%
                summarise(`dN/dS mean` = mean(`dN/dS`)) %>%
                ungroup()

    for (mut_cat in mut_cat_factor){
        
        dnds1 <- counts_deformed$`dN/dS mean`[counts_deformed$mut_cat == mut_cat & counts_deformed$exp_line == "III_MT-2"]
        dnds2 <- counts_deformed$`dN/dS mean`[counts_deformed$mut_cat == mut_cat & counts_deformed$exp_line == "III_MT-4"]
        dnds3 <- counts_deformed$`dN/dS mean`[counts_deformed$mut_cat == mut_cat & counts_deformed$exp_line == "IV_MT-2"]

        t_test_res1 <- t.test(dnds1, dnds2, paired = T, alternative = "two.sided")
        
        t_test_res_all <- rbind(t_test_res_all, c("III_MT-2 vs III_MT-4", gene, mut_cat, round(mean(dnds1),3), round(mean(dnds2),3), t_test_res1$p.value))

        t_test_res2 <- t.test(dnds1, dnds3, paired = T, alternative = "two.sided")
        
        t_test_res_all <- rbind(t_test_res_all, c("III_MT-2 vs IV_MT-2", gene, mut_cat, round(mean(dnds1),3), round(mean(dnds3),3), t_test_res2$p.value))
    
        t_test_res3 <- t.test(dnds2, dnds3, paired = T, alternative = "two.sided")
        
        t_test_res_all <- rbind(t_test_res_all, c("III_MT-4 vs IV_MT-2", gene, mut_cat, round(mean(dnds2),3), round(mean(dnds3),3), t_test_res3$p.value))
    
    }
    


}

colnames(t_test_res_all) <- c("comparison", "gene", "mut_cat", "mean(dN/dS) 1st", "mean(dN/dS) 2nd", "p-value")

datatable(t_test_res_all, filter = "bottom")





```




### Whole Genome Plot

let's plot the dN/dS trends over time for the whole HIV-1 genome

```{r whole genome dn/ds,results=T}

# plot the difference eof synonymous and non-synonymous mutations
plt_type <- "plotly"

# get the synonymous and nonsynonymous likelihoods number for processing entire genome



syn_pos <- as.numeric(sum(gene_summary[,3]))
nonsyn_pos <- as.numeric(sum(gene_summary[,4]))

# iterate over experiments and plot

for (experiment in unique(variants$exp)){
    counts <- variants %>%
        mutate(mut_cat = factor(mut_cat, levels = mut_cat_factor)) %>%
        filter(exp == experiment) %>%
        mutate(effect_simplified = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
        mutate(effect_simplified = replace_na(effect_simplified, "N")) %>%
        filter(effect_simplified != "U") %>%
        mutate(effect_simplified = factor(effect_simplified, levels = c("N", "S"))) %>%
        group_by(exp_line, passage, mut_cat, effect_simplified, .drop=F) %>%
        count() %>%
        dplyr::rename(mut_count = n) %>%
        mutate(host_line = ifelse(exp_line %in% c("MT-4_1","MT-4_2"), "MT4", "MT2")) %>%
        mutate(exp_line = factor(exp_line, levels = exp_line_factor)) %>%
        mutate(effect_simplified = factor(effect_simplified, levels = c("N", "S"))) %>%
        mutate(host_line = factor(host_line, levels = c("MT2", "MT4")))
    
    counts_plt <- 
        counts %>%
        spread(key = effect_simplified, mut_count) %>%
        mutate(S = case_when(S == 0 ~ S+0.01 , S !=0 ~ S)) %>%
        mutate(`dN/dS` = log((N/nonsyn_pos)/(S/syn_pos))) 
    
    counts_plt$`dN/dS`[is.infinite(counts_plt$`dN/dS`) | is.nan(counts_plt$`dN/dS`)] <- 0
    
    
    fig <- counts_plt %>%  
        ggplot(aes(x=passage, y = `dN/dS`, color = exp_line)) +
        geom_line(size=1) +
        facet_wrap(~mut_cat, scales = "free")+
        labs(x="\nPassage", y= "Log2(dN/dS)\n", color = "Virus Line", title = paste0("dN/dS Over Time\n in ", experiment, " for the Whole Genome", "\n"))+
        scale_color_manual(values = gg_var_list[[experiment]][["line_palette"]]) +
        scale_y_continuous() +
        geom_hline(yintercept = 0, color = "black", size = 2) +
        theme_bw() +
        theme(plot.title = element_text(size = gg_size_list[[plt_type]][["title.size"]], hjust = 0.5),
        axis.title = element_text(size = gg_size_list[[plt_type]][["axis.title.size"]]),
        axis.text.x = element_text(size = gg_size_list[[plt_type]][["axis.text.size"]], angle = 90),
        axis.text.y = element_text(size = gg_size_list[[plt_type]][["axis.text.size"]]),
        legend.title = element_text(size=gg_size_list[[plt_type]][["legend.title.size"]]),
        legend.text = element_text(size = gg_size_list[[plt_type]][["legend.text.size"]]))

    print(fig)

    # save pngs

    pdf(paste0("/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/figs/pdf/jeffrey-barrick/dn_ds_", experiment, "_whole_genome", ".pdf")
    , width = 21, height = 21)
    print(fig)
    dev.off()
    
}





```





### Whole Genome Statistical Significance

#### positive vs purifying selection in each virus line

```{r, results=T}




syn_pos <- as.numeric(sum(gene_summary[,3]))
nonsyn_pos <- as.numeric(sum(gene_summary[,4]))


t_test_res_all <- data.frame()

for (experiment in unique(variants$exp)){
    counts <- variants %>%
        mutate(mut_cat = factor(mut_cat, levels = mut_cat_factor)) %>%
        filter(exp == experiment) %>%
        mutate(effect_simplified = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
        mutate(effect_simplified = replace_na(effect_simplified, "N")) %>%
        mutate(effect_simplified = factor(effect_simplified, levels = c("N", "S"))) %>%
        group_by(exp_line, passage, mut_cat, effect_simplified, .drop=F) %>%
        count() %>%
        dplyr::rename(mut_count = n) %>%
        mutate(host_line = ifelse(exp_line %in% c("MT-4_1","MT-4_2"), "MT4", "MT2")) %>%
        mutate(exp_line = factor(exp_line, levels = exp_line_factor)) %>%
        mutate(effect_simplified = factor(effect_simplified, levels = c("N", "S"))) %>%
        mutate(host_line = factor(host_line, levels = c("MT2", "MT4"))) %>%
        spread(key = effect_simplified, mut_count) %>%
        mutate(S = case_when(S == 0 ~ S+0.01 , S !=0 ~ S)) %>%
        mutate(`dN/dS` = log((N/nonsyn_pos)/(S/syn_pos)))
        
    
    counts$`dN/dS`[is.infinite(counts$`dN/dS`) | is.nan(counts$`dN/dS`)] <- 0
    
    for (mut_cat in mut_cat_factor){
        for (host_line in unique(counts$host_line)){
            dnds <- counts$`dN/dS`[counts$mut_cat == mut_cat & counts$host_line == host_line]
            
            t_test_res <- t.test(dnds, mu = 0, alternative = "two.sided")
            
            t_test_res_all <- rbind(t_test_res_all, c(experiment, host_line, mut_cat, round(mean(dnds),3), t_test_res$p.value))
        
    }
    }
    

}


colnames(t_test_res_all) <- c("exp", "host", "mut_cat", "mean(dN/dS)", "p-value")

datatable(t_test_res_all, filter = "bottom")




```



#### dN/dS level tests between each virus line group

```{r, results = T}





t_test_res_all <- data.frame()

counts <- variants %>%
    filter(exp != "EXPIII" | passage <= 510) %>%
    mutate(mut_cat = factor(mut_cat, levels = mut_cat_factor)) %>%
    mutate(effect_simplified = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
    mutate(effect_simplified = replace_na(effect_simplified, "N")) %>%
    mutate(effect_simplified = factor(effect_simplified, levels = c("N", "S"))) %>%
    group_by(exp_line, passage, mut_cat, effect_simplified, .drop=F) %>%
    count() %>%
    dplyr::rename(mut_count = n) %>%
    mutate(host_line = ifelse(exp_line %in% c("MT-4_1","MT-4_2"), "MT4", "MT2")) %>%
    mutate(exp_line = factor(exp_line, levels = exp_line_factor)) %>%
    mutate(effect_simplified = factor(effect_simplified, levels = c("N", "S"))) %>%
    mutate(host_line = factor(host_line, levels = c("MT2", "MT4"))) %>%
    spread(key = effect_simplified, mut_count) %>%
    mutate(S = case_when(S == 0 ~ S+0.01 , S !=0 ~ S)) %>%
    mutate(`dN/dS` = log((N/nonsyn_pos)/(S/syn_pos)))


counts$`dN/dS`[is.infinite(counts$`dN/dS`) | is.nan(counts$`dN/dS`)] <- 0

# counts_deformed[counts_deformed$exp_line == "III_MT-2" & counts_deformed$passage == 30 & counts_deformed$mut_cat == "Sporadic",]

counts_deformed <- counts %>% 
            select(!c("N", "S")) %>%
            mutate(exp_line = case_when(exp_line %in% exp_line_factor[1:2] ~ "III_MT-2",
                                        exp_line %in% exp_line_factor[3:4] ~ "III_MT-4",
                                        exp_line %in% exp_line_factor[4:8] ~ "IV_MT-2")) %>%
            mutate(exp_line = factor(exp_line, levels = c("III_MT-2", "III_MT-4", "IV_MT-2"))) %>%
            select(-"host_line") %>%
            group_by(exp_line, passage, mut_cat) %>%
            summarise(`dN/dS mean` = mean(`dN/dS`)) %>%
            ungroup()

for (mut_cat in mut_cat_factor){

    dnds1 <- counts_deformed$`dN/dS mean`[counts_deformed$mut_cat == mut_cat & counts_deformed$exp_line == "III_MT-2"]
    dnds2 <- counts_deformed$`dN/dS mean`[counts_deformed$mut_cat == mut_cat & counts_deformed$exp_line == "III_MT-4"]
    dnds3 <- counts_deformed$`dN/dS mean`[counts_deformed$mut_cat == mut_cat & counts_deformed$exp_line == "IV_MT-2"]

    t_test_res1 <- t.test(dnds1, dnds2, paired = T, alternative = "two.sided")
    
    t_test_res_all <- rbind(t_test_res_all, c("III_MT-2 vs III_MT-4", mut_cat, round(mean(dnds1),3), round(mean(dnds2),3), t_test_res1$p.value))

    t_test_res2 <- t.test(dnds1, dnds3, paired = T, alternative = "two.sided")
    
    t_test_res_all <- rbind(t_test_res_all, c("III_MT-2 vs IV_MT-2", mut_cat, round(mean(dnds1),3), round(mean(dnds3),3), t_test_res2$p.value))

    t_test_res3 <- t.test(dnds2, dnds3, paired = T, alternative = "two.sided")
    
    t_test_res_all <- rbind(t_test_res_all, c("III_MT-4 vs IV_MT-2", mut_cat, round(mean(dnds2),3), round(mean(dnds3),3), t_test_res3$p.value))

}
    




colnames(t_test_res_all) <- c("comparison", "mut_cat", "mean(dN/dS) 1st", "mean(dN/dS) 2nd", "p-value")

datatable(t_test_res_all, filter = "bottom")





```




Next step would be to profile the nonsynonymous mutations in terms of the amino acid changes that they cause and profile the synonymous mutations 
in terms of the codon usage changes (one can test this with permutation of mutations and simulating data)