---
title:  
    "Quality of EXPIII"
date: 
    "July 8th, 2024"
author:  
    "Ali Movasati"  
output:  
    html_document:
        #css: style.css
        toc: true
        toc_depth: 2
        toc_float: true
        number_sections: true
        theme: lumen
        highlight: zenburn
        code_folding: hide
        code_download: true # to make the Rmd source code available
---



```{r loading libraries, echo=FALSE}

library("ggplot2")
library("stringr")
library("tidyr")
library("magrittr")


```



```{r read-in the data}

# determine the server path

if (file.exists("/home/amovas/")){
  print("Remote HPC Connection!")
  wd <- "/home/amovas/data/genome-evo-proj/"
} else {
  print("Local PC Connection!")
  wd <- "/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/"
}



# read in quality table
quals = read.table(file = paste0(wd, "results/tables/pipeline-outputs/qc/all_quals.tsv.gz"), sep = "\t", header = FALSE, stringsAsFactors = FALSE)
colnames(quals) = c("file","average","cover","average_5","average_a","average_b","average_c","average_d","average_e","average_3","cover_all","mapped", "mapquality")
quals$full_sample_name = as.character(lapply(str_split(lapply(str_split(quals$file, pattern = "/"), "[[", 12), pattern = "_"), "[[", 1))
quals = quals[,-1]


# read in the metadata
meta = read.table(file = "/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/data/metadata/NGS_samples_list_processed_vlast.csv", sep = ",", header = TRUE, stringsAsFactors = FALSE)
meta = meta[meta$included == TRUE,]


# merge the two dataset
quals_meta <- merge(quals, meta, by = "full_sample_name")


# check merged worked fine
nrow(quals_meta)
quals[quals$full_sample_name %notin% quals_meta$full_sample_name,]

sum(quals$average_3<= 200)

##### plot

# keep relevant fields
quals <- quals_meta[,c("full_sample_name", "average_5", "average_a", "average_b", "average_c", "average_d", "average_e", "average_3", "virus_line_no", "transfer_no", "cover_all")]

# reshape the dataset and prepare for plotting
quals_gathered <- gather(quals, key = "amplicon", value = "average_depth", 2:8)

# convert certain fields to factor
quals_gathered$virus_line_no <- factor(quals_gathered$virus_line_no, levels = c(13,14,15,16,17,18,19,20))
quals_gathered$amplicon <- factor(quals_gathered$amplicon, levels = c("average_5","average_a", "average_b", "average_c", "average_d", "average_e","average_3"))
quals_gathered$full_sample_name <- factor(quals_gathered$full_sample_name)

# check the data type of dataset fields
lapply(quals_gathered,class)

# sort the dataset (optional)
quals_gathered <- quals_gathered[order(quals_gathered$full_sample_name),]

# logarithmize the depth value for better visualization
quals_gathered$average_depth <- log10(quals_gathered$average_depth)


```{r sequencing depth summary plots}


d <- quals_gathered %>% filter(virus_line_no %in% 13:16) %>% group_by(amplicon, virus_line_no) %>% summarise (mean_coverage = mean(average_depth, na.rm = T), .groups = 'drop') %>%
    mutate(virus_line_no = case_when(virus_line_no==13 ~ "MT-2_1", virus_line_no==14 ~ "MT-2_2", virus_line_no==15 ~ "MT-4_1", virus_line_no==16 ~ "MT-4_2")) %>%
    mutate(virus_line_no = factor(virus_line_no, levels = exp_line_factor))



# render the ggplot object

depth_summary <- ggplot(data=d, aes(x = amplicon, y = mean_coverage,group=virus_line_no, color=virus_line_no))+
    geom_line( linewidth=1.2) +
    geom_point(color="black", size=3) +
    scale_color_manual(values = line_col_palette) +
    labs(title = "Sequencing Depth\n", x = "\nAmplicons", y = "Average Sequencing Depth (log10)\n", color = "Virus Line") +
    scale_x_discrete(labels = c("Amp-A (5utr)", "Amp_A","Amp_B","Amp_C","Amp_D","Amp_E", "Amp-E (3utr)")) +
    scale_y_continuous(limit = c(-0.5,5), breaks = 0:5) +
    geom_hline(yintercept = log10(200), linetype="dashed") +
    theme_bw() +
    theme(plot.title = element_text(size = 40, face = "bold", hjust = 0.5),
        axis.title.x = element_text(size = 25),
        axis.title.y = element_text(size = 25),
        axis.text = element_text(size = 15))


# save figure
date <- str_split(Sys.time(), pattern = " ")[[1]][1]
fig_dir <- paste0(wd, "results/figs/publication/expiii/provisional/", date, "/")
ifelse(!dir.exists(file.path(fig_dir)), dir.create(file.path(fig_dir)), FALSE)
pdf(file = paste0(fig_dir, "Sequencing_depth_line_plot.pdf"), height = 14, width = 14)
depth_summary
dev.off()


# box plot

quals_gathered_iii <- quals_gathered %>% filter(virus_line_no %in% 13:16) %>%
        mutate(virus_line_no = case_when(virus_line_no==13 ~ "MT-2_1", virus_line_no==14 ~ "MT-2_2", virus_line_no==15 ~ "MT-4_1", virus_line_no==16 ~ "MT-4_2")) %>%
        mutate(virus_line_no = factor(virus_line_no, levels = exp_line_factor))

depth_summary <- ggplot(data=quals_gathered_iii, aes(x = amplicon, y = average_depth,fill=virus_line_no))+
    geom_boxplot()+
    scale_fill_manual(values = line_col_palette) +
    labs(title = "Sequencing Depth\n", x = "\nAmplicons", y = "Average Sequencing Depth (log10)\n", fill = "Virus Line") +
    scale_x_discrete(labels = c("Amp-A (5utr)", "Amp_A","Amp_B","Amp_C","Amp_D","Amp_E", "Amp-E (3utr)")) +
    scale_y_continuous(limit = c(-0.5,5), breaks = 0:5) +
    geom_hline(yintercept = log10(200), linetype="dashed") +
    theme_bw() +
    theme(plot.title = element_text(size = 40, face = "bold", hjust = 0.5),
        axis.title.x = element_text(size = 25),
        axis.title.y = element_text(size = 25),
        axis.text = element_text(size = 15))


# save figure
date <- str_split(Sys.time(), pattern = " ")[[1]][1]
fig_dir <- paste0(wd, "results/figs/publication/expiii/provisional/", date, "/")
ifelse(!dir.exists(file.path(fig_dir)), dir.create(file.path(fig_dir)), FALSE)
pdf(file = paste0(fig_dir, "Sequencing_depth_box_plot.pdf"), height = 14, width = 14)
depth_summary
dev.off()



```



```{r}




```