---
title:  
    "Jeffrey Barrick et al. 2009 Analysis"
date: 
    "Nov 15th, 2023"
author:  
    "Ali Movasati"  
output:  
    html_document:
        toc: true
        toc_depth: 3
        toc_float: true
        number_sections: true
        theme: lumen
        highlight: zenburn
        code_folding: hide
---




```{r setting global chunk parameters, include=FALSE}

# options for code chunk and output inclusion 
knitr::opts_chunk$set(
  message = FALSE, echo = TRUE, warning=FALSE, results = "hide"
)

# options for figures
knitr::opts_chunk$set(
  fig.width = 8, fig.height = 8, fig.align = "center"
)

# options(scipen=999)

# margin of plotly
m <- list(
  l = 150,
  r = 150,
  b = 100,
  t = 100,
  pad = 4
)


```




```{r loading libraries, echo=FALSE}

# loading libraries
library("plotly")
library("ggplot2")
library("DT")
library("dplyr")
library("tidyr")
library("Biostrings")
library("stringr")
library("magrittr")
library("ggvenn")
library("VennDiagram")
library("ggVennDiagram")


```


# Introduction


In this document I will try to reproduce some of the analysis provided and ideas presented by Jeffrey Barrick et al. (2019).

[Jeffrey Barrick](https://www.nature.com/articles/nature08480)




```{r read-in the data}

# read in the data
variants_ann <- read.table(file = "/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/tables/2-p/all_annotated_variants.tsv.gz", 
sep = "\t", stringsAsFactors = FALSE, header = TRUE)

# set correct data types
variants_ann$exp_line <- as.integer(variants_ann$exp_line)


# set host cell line info
variants_ann$host_line <- ifelse(variants_ann$exp_line %in% c("13", "14","17", "18", "19", "20"), "MT2", "MT4")


# account for the switched experimental lines
variants_ann$exp_line[variants_ann$exp_line == 15 & variants_ann$passage > 360] <- "A"
variants_ann$exp_line[variants_ann$exp_line == 16 & variants_ann$passage > 360] <- "B"

variants_ann$exp_line[variants_ann$exp_line == "A"] <- "16"
variants_ann$exp_line[variants_ann$exp_line == "B"] <- "15"



variants_ann$exp_line[variants_ann$exp_line == 18 & variants_ann$passage > 350] <- "A"
variants_ann$exp_line[variants_ann$exp_line == 19 & variants_ann$passage > 350] <- "B"

variants_ann$exp_line[variants_ann$exp_line == "A"] <- "19"
variants_ann$exp_line[variants_ann$exp_line == "B"] <- "18"


# add a field containing pos, ref, and alt info of muts
variants_ann$mut_info <- paste(variants_ann$genomic_pos, variants_ann$ref_allele, variants_ann$alt_allele, sep = "_")

## prepare data
threshold <- 0.01

variants_ann %<>% 
                mutate(exp = case_when(exp_line %in% 13:16 ~ "EXPIII", exp_line %in% 17:20 ~ "EXPIV")) %>%
                mutate(exp_line = case_when(exp_line==13 ~ "MT-2_1", exp_line==14 ~ "MT-2_2", exp_line==15 ~ "MT-4_1", exp_line==16 ~ "MT-4_2",
                                            exp_line==17 ~ "MT-2_i", exp_line==18 ~ "MT-2_ii", exp_line==19 ~ "MT-2_iii", exp_line==20 ~ "MT-2_iv")) %>%
                mutate(mut_info = paste(genomic_pos, ref_allele, alt_allele, sep = "_")) %>%
                mutate(mut_cat = case_when((allele_freq < 0.05) ~ "Sporadic", (allele_freq >= 0.05 & allele_freq < 0.5) ~ "Minority", (allele_freq >= 0.5 & allele_freq < 0.99) ~ "Majority", (allele_freq >= 0.99) ~ "Fixated")) %>% 
                filter(allele_freq >= threshold)

variants <- variants_ann %>% filter(!duplicated(variant_id))

# set the ggplot variables

gg_var_list <- list(EXPIII = list(line_palette = c("#ff00ff", "#ff2400", "#6600cc", "#0000ff"), break_step = 50),
                    EXPIV = list(line_palette = c("#68FF00", "#22E9AA", "#0DD100", "#009163"), break_step = 5))

gg_size_list <- list(ggplot = list(title.size = 50, axis.title.size = 45, axis.text.size = 40, legend.title.size = 35, legend.text.size = 30, strip.text.x.size = 35),
                    plotly = list(title.size = 25, axis.title.size = 20, axis.text.size = 10, legend.title.size = 20, legend.text.size = 15, strip.text.x.size = 20))




```



In this paper, four line of evidence is suggested to reject the drift hypothesis outlined by the author. The first evidence suggests that under the drift hypothesis we would expect to see higher number of synonymous than
non-synonymous mutations. First let's plot the number of these mutations types for each experiment and each mutation category.


```{r mutation accumulation plot stratified by translational impact, results=T}


plt_type <- "plotly"

for (experiment in unique(variants$exp)){

    counts <- variants %>%
        filter(exp == experiment) %>%
        mutate(effect_simplified = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
        mutate(effect_simplified = replace_na(effect_simplified, "N")) %>%
        filter(effect_simplified != "U") %>%
        group_by(exp_line, passage, mut_cat, effect_simplified) %>%
        count() %>%
        mutate(host_line = ifelse(exp_line %in% c("MT-4_1","MT-4_2"), "MT4", "MT2")) %>%
        dplyr::rename(mut_count = n) %>%
        mutate_at(c(1,3,4,6), as.factor) %>%
        mutate(mut_cat = factor(mut_cat, levels = c("Sporadic", "Minority", "Majority", "Fixated"))) %>%
        mutate_at(c(2,5), as.numeric) 
    
    
    fig <- counts %>%  
        ggplot(aes(x=passage, y = mut_count, color = exp_line, linetype=effect_simplified)) +
        geom_line(size=1) +
        facet_wrap(~mut_cat, scales = "free")+
        labs(x="\nPassage", y= "Mutation Count\n", title = paste0("Mutation Accumulation in ", experiment, "\n"))+
        scale_color_manual(name = "Virus Line", values = gg_var_list[[experiment]][["line_palette"]]) +
        scale_linetype_manual(name = "Translational Effect", , values = c("dashed", "solid")) +
        theme_bw() +
        theme(plot.title = element_text(size = gg_size_list[[plt_type]][["title.size"]], hjust = 0.5),
        axis.title = element_text(size = gg_size_list[[plt_type]][["axis.title.size"]]),
        axis.text.x = element_text(size = gg_size_list[[plt_type]][["axis.text.size"]], angle = 90),
        axis.text.y = element_text(size = gg_size_list[[plt_type]][["axis.text.size"]]),
        legend.title = element_text(size=gg_size_list[[plt_type]][["legend.title.size"]]),
        legend.text = element_text(size = gg_size_list[[plt_type]][["legend.text.size"]]))

    print(fig)

    # save pngs

    pdf(paste0("/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/figs/pdf/jeffrey-barrick/syn_vs_nonsyn_", experiment, ".pdf")
    , width = 21, height = 21)
    print(fig)
    dev.off()
}

```



```{r differential mutation accumulation plot based on translational impact, results=T}

# plot the difference eof synonymous and non-synonymous mutations
plt_type <- "plotly"

for (experiment in unique(variants$exp)){
    counts <- variants %>%
        filter(exp == experiment) %>%
        mutate(effect_simplified = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
        mutate(effect_simplified = replace_na(effect_simplified, "N")) %>%
        filter(effect_simplified != "U") %>%
        group_by(exp_line, passage, mut_cat, effect_simplified) %>%
        count() %>%
        mutate(host_line = ifelse(exp_line %in% c("MT-4_1","MT-4_2"), "MT4", "MT2")) %>%
        dplyr::rename(mut_count = n) %>%
        mutate_at(c(1,3,4,6), as.factor) %>%
        mutate(mut_cat = factor(mut_cat, levels = c("Sporadic", "Minority", "Majority", "Fixated"))) %>%
        mutate_at(c(2,5), as.numeric) 
    
    counts_plt <- 
        counts %>%
            spread(key = effect_simplified, mut_count) %>%
            replace(is.na(.), 0) %>%
            mutate(`S-N` = S-N)
    
    fig <- counts_plt %>%  
        ggplot(aes(x=passage, y = `S-N`, color = exp_line)) +
        geom_line(size=1) +
        facet_wrap(~mut_cat, scales = "free")+
        labs(x="\nPassage", y= "Synonymous-Nonsynonymous Count\n", color = "Virus Line", title = paste0("Differential Mutation Accumulation in ", experiment, "\n"))+
        scale_color_manual(values = gg_var_list[[experiment]][["line_palette"]]) +
        #scale_linetype_manual(values = c("dashed", "solid")) +
        theme_bw() +
        theme(plot.title = element_text(size = gg_size_list[[plt_type]][["title.size"]], hjust = 0.5),
        axis.title = element_text(size = gg_size_list[[plt_type]][["axis.title.size"]]),
        axis.text.x = element_text(size = gg_size_list[[plt_type]][["axis.text.size"]], angle = 90),
        axis.text.y = element_text(size = gg_size_list[[plt_type]][["axis.text.size"]]),
        legend.title = element_text(size=gg_size_list[[plt_type]][["legend.title.size"]]),
        legend.text = element_text(size = gg_size_list[[plt_type]][["legend.text.size"]]))

    print(fig)

    # save pngs

    pdf(paste0("/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/figs/pdf/jeffrey-barrick/difference_syn_vs_nonsyn_", experiment, ".pdf")
    , width = 21, height = 21)
    print(fig)
    dev.off()

}

```




```{r statistical tests for mutations counts stratified by translational impact, results=T}

t_test_res_all <- data.frame()

for (experiment in unique(variants$exp)){
    
    counts <- variants %>%
        filter(exp == experiment) %>%
        mutate(effect_simplified = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
        mutate(effect_simplified = replace_na(effect_simplified, "N")) %>%
        filter(effect_simplified != "U") %>%
        mutate(mut_cat = factor(mut_cat, levels = c("Sporadic", "Minority", "Majority", "Fixated"))) %>%
        mutate_at(c("exp_line", "mut_cat", "effect_simplified"), as.factor) %>%
        group_by(exp_line, passage, mut_cat, effect_simplified, .drop = FALSE) %>%
        count() %>%
        mutate(host_line = ifelse(exp_line %in% c("MT-4_1","MT-4_2"), "MT4", "MT2")) %>%
        dplyr::rename(mut_count = n) %>%
        mutate_at(c(1,3,4,6), as.factor) %>%
        mutate_at(c(2,5), as.numeric)

    for (mut_cat in unique(counts$mut_cat)){
        for (host_line in unique(counts$host_line)){
            syn_count <- counts$mut_count[counts$effect_simplified == "S" & counts$mut_cat == mut_cat & counts$host_line == host_line]
            nonsyn_count <- counts$mut_count[counts$effect_simplified == "N" & counts$mut_cat == mut_cat & counts$host_line == host_line]
            t_test_res <- t.test(syn_count, nonsyn_count, paired = T, alternative = "two.sided")
            
            t_test_res_all <- rbind(t_test_res_all, c(experiment, host_line, mut_cat, mean(syn_count-nonsyn_count), sd(syn_count-nonsyn_count), t_test_res$p.value))
            
        }
        
    }
}

colnames(t_test_res_all) <- c("exp", "host", "mut_cat", "mean(synonymous-nonsynonymous)", "sd(synonymous-nonsynonymous)", "p-value")

datatable(t_test_res_all)

```



Let's see what are the chances of a point mutation in the protein-coding regions of HIV-1 to be synonymous vs nonsynonymous
and subsequently normalize all these counts base on that.


```{r number of all possible mutations + percentage of synonymous ones, results=T}

# read in ancestral reference sequence 
fastaFile <- readDNAStringSet("/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/data/reference/plasmid/hiv_plasmid_ref_genome.fasta")
seq_name = names(fastaFile)
sequence = paste(fastaFile)


# read in feature coordinate data

features <- read.table(file = "/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/tables/2-p/cds_feature_list.tsv", sep = "\t", header = T)

# by looking at the features list:

features2 <- data.frame(name = c("gag", "pol", "vif", "vpr", "tat1", "rev1", "vpu", "env", "tat2", "rev2", "nef"),
                        start = c(790, 2085, 5041, 5559, 5830, 5969, 6061, 6221, 8369+2-1, 8369+3-1, 8787),
                        end = c(2292, 5096, 5619, 5849, 6044-2, 6044-1, 6306, 8785, 8414, 8643, 9407))

# check
# (features2$end-features2$start+1)/3



# make objects containing bases and aas
bases <- c("A", "T", "C", "G")
aa <- unique(GENETIC_CODE)


gene_summary <- data.frame()
aa_summary <- data.frame()
for (gene in features2$name){ 
    # counters and value containers
    syn_count <- 0
    nonsyn_count <- 0
    aa_ref_composition <- c()
    aa_alt_composition <- c()

    start_pos <- features2$start[features2$name == gene]
    end_pos <- features2$end[features2$name == gene]
    gene_length <- end_pos - start_pos + 1
    codon_count <- gene_length/3

    # get the codons
    codons <- substring(substring(sequence,start_pos, end_pos), seq(1, end_pos-start_pos+1, 3), seq(3, end_pos-start_pos+1, 3))

    # iterate over the codons
    for (i in 1:length(codons)){
        
        ref_codon = codons[i]
        # GENETIC_CODE from biostrings
        ref_aa <- GENETIC_CODE[[ref_codon]]

        aa_ref_composition <- append(aa_ref_composition,ref_aa)

        # iterate over possible mutations for each codon
        for (j in 1:3){
            mutation <- bases != str_sub(ref_codon, start = j, end = j)
            for (k in bases[mutation]){
                
                alt_codon <- ref_codon
                substring(alt_codon,j, j) <- k
                alt_aa <- GENETIC_CODE[[alt_codon]]
                
                if (alt_aa == ref_aa){
                    syn_count <- syn_count + 1
                } else {
                    nonsyn_count <- nonsyn_count + 1
                    aa_alt_composition <- append(aa_alt_composition,alt_aa)
                }
            }   
        }
    } 
    all_count <- syn_count + nonsyn_count
    gene_summary <- rbind(gene_summary, c(gene, gene_length, syn_count, syn_count/all_count, nonsyn_count, nonsyn_count/all_count, all_count))
    
    # counts amino acids in reference and their possibilities to be produced
    aa_ref_composition <- factor(aa_ref_composition, levels = aa)
    aa_ref_composition_tbl <- table(aa_ref_composition)

    aa_alt_composition <- factor(aa_alt_composition, levels = aa)
    aa_alt_composition_tbl <- table(aa_alt_composition)

    aa_summary <- rbind(aa_summary, c(gene, "ref_composition", as.integer(aa_ref_composition_tbl)))
    aa_summary <- rbind(aa_summary, c(gene, "alt_composition", as.integer(aa_alt_composition_tbl)))
}

colnames(gene_summary) <- c("gene", "gene_length", "synonymous_count", "synonymous_percentage", "nonsynonymous_count", "nonsynonymous_percentage", "all_count")
colnames(aa_summary) <- c("gene", "mode", aa)




datatable(gene_summary)

gene_summary %<>% mutate_at(2:ncol(gene_summary), as.numeric)



datatable(as.data.frame(colMeans(gene_summary[,c(4,6)])))



aa_summary %<>% 
    mutate_at(3:ncol(aa_summary), as.numeric)

aa_summary$sum <- rowSums(aa_summary[,3:ncol(aa_summary)])



mypaste <- function(x,y) paste(x, "(%", y, ")", sep="")

# percentage of aa throughout the genome
datatable(as.data.frame(colMeans(round(aa_summary[,3:(ncol(aa_summary)-1)]/aa_summary[,ncol(aa_summary)]*100, 1))))

# percentage and count of aa per gene
aa_summary[,3:(ncol(aa_summary)-1)] <- mapply(mypaste, aa_summary[,3:(ncol(aa_summary)-1)], round(aa_summary[,3:(ncol(aa_summary)-1)]/aa_summary[,ncol(aa_summary)]*100, 1))

datatable(aa_summary)



```


```{r normalized syn-nonsyn plots, results = T}

syn_chance <- as.data.frame(colMeans(gene_summary[,c(4,6)]))[1,1]
nonsyn_chance <- as.data.frame(colMeans(gene_summary[,c(4,6)]))[2,1]


# plot the difference eof synonymous and non-synonymous mutations
plt_type <- "plotly"

for (experiment in unique(variants$exp)){
    counts <- variants %>%
        filter(exp == experiment) %>%
        mutate(effect_simplified = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
        mutate(effect_simplified = replace_na(effect_simplified, "N")) %>%
        filter(effect_simplified != "U") %>%
        group_by(exp_line, passage, mut_cat, effect_simplified) %>%
        count() %>%
        mutate(host_line = ifelse(exp_line %in% c("MT-4_1","MT-4_2"), "MT4", "MT2")) %>%
        dplyr::rename(mut_count = n) %>%
        mutate_at(c(1,3,4,6), as.factor) %>%
        mutate(mut_cat = factor(mut_cat, levels = c("Sporadic", "Minority", "Majority", "Fixated"))) %>%
        mutate_at(c(2,5), as.numeric) 
    
    counts_plt <- 
        counts %>%
            spread(key = effect_simplified, mut_count) %>%
            replace(is.na(.), 0) %>%
            mutate(`Normalized S-N` = (1/syn_chance)*S - (1/nonsyn_chance)*N)
    
    fig <- counts_plt %>%  
        ggplot(aes(x=passage, y = `Normalized S-N`, color = exp_line)) +
        geom_line(size=1) +
        facet_wrap(~mut_cat, scales = "free")+
        labs(x="\nPassage", y= "Synonymous-Nonsynonymous Normalized Count\n", color = "Virus Line", title = paste0("Normalized Differential Mutation Accumulation\n in ", experiment, "\n"))+
        scale_color_manual(values = gg_var_list[[experiment]][["line_palette"]]) +
        #scale_linetype_manual(values = c("dashed", "solid")) +
        theme_bw() +
        theme(plot.title = element_text(size = gg_size_list[[plt_type]][["title.size"]], hjust = 0.5),
        axis.title = element_text(size = gg_size_list[[plt_type]][["axis.title.size"]]),
        axis.text.x = element_text(size = gg_size_list[[plt_type]][["axis.text.size"]], angle = 90),
        axis.text.y = element_text(size = gg_size_list[[plt_type]][["axis.text.size"]]),
        legend.title = element_text(size=gg_size_list[[plt_type]][["legend.title.size"]]),
        legend.text = element_text(size = gg_size_list[[plt_type]][["legend.text.size"]]))

    print(fig)

    # save pngs

    pdf(paste0("/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/figs/pdf/jeffrey-barrick/normalized_difference_syn_vs_nonsyn_", experiment, ".pdf")
    , width = 21, height = 21)
    print(fig)
    dev.off()

}

```


```{r normalized statistical tests for mutations counts stratified by translational impact, results=T}

t_test_res_all <- data.frame()

for (experiment in unique(variants$exp)){
    
    counts <- variants %>%
        filter(exp == experiment) %>%
        mutate(effect_simplified = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
        mutate(effect_simplified = replace_na(effect_simplified, "N")) %>%
        filter(effect_simplified != "U") %>%
        mutate(mut_cat = factor(mut_cat, levels = c("Sporadic", "Minority", "Majority", "Fixated"))) %>%
        mutate_at(c("exp_line", "mut_cat", "effect_simplified"), as.factor) %>%
        group_by(exp_line, passage, mut_cat, effect_simplified, .drop = FALSE) %>%
        count() %>%
        mutate(host_line = ifelse(exp_line %in% c("MT-4_1","MT-4_2"), "MT4", "MT2")) %>%
        dplyr::rename(mut_count = n) %>%
        mutate_at(c(1,3,4,6), as.factor) %>%
        mutate_at(c(2,5), as.numeric)

    for (mut_cat in unique(counts$mut_cat)){
        for (host_line in unique(counts$host_line)){
            syn_count <- counts$mut_count[counts$effect_simplified == "S" & counts$mut_cat == mut_cat & counts$host_line == host_line]
            syn_count_normalized <- syn_count*(1/syn_chance)
            nonsyn_count <- counts$mut_count[counts$effect_simplified == "N" & counts$mut_cat == mut_cat & counts$host_line == host_line]
            nonsyn_count_normalized <- nonsyn_count*(1/nonsyn_chance)
            t_test_res <- t.test(syn_count_normalized, nonsyn_count_normalized, paired = T, alternative = "two.sided")
            
            t_test_res_all <- rbind(t_test_res_all, c(experiment, host_line, mut_cat, mean(syn_count_normalized-nonsyn_count_normalized), sd(syn_count_normalized-nonsyn_count_normalized), t_test_res$p.value))
            
        }
        
    }
}

colnames(t_test_res_all) <- c("exp", "host", "mut_cat", "normalized mean(synonymous-nonsynonymous)", "normalized sd(synonymous-nonsynonymous)", "p-value")

datatable(t_test_res_all)

```


Next step would be to profile the nonsynonymous mutations in terms of the amino acid changes that they cause and profile the synonymous mutations in terms of the codon usage changes.