---
title:  
    "Overview of HIV-1 LTEE genomic data (EXPIII)"
date: 
    "Oct 3rd, 2023"
author:  
    "Ali Movasati"  
output:  
    html_document:
        #css: style.css
        toc: true
        toc_depth: 3
        toc_float: true
        number_sections: true
        theme: lumen
        highlight: zenburn
        code_folding: hide
---



```{r setting global chunk parameters, include=FALSE}

# options for code chunk and output inclusion 
knitr::opts_chunk$set(
  message = FALSE, echo = TRUE, warning=FALSE, results = "hide"
)

# options for figures
knitr::opts_chunk$set(
  fig.width = 8, fig.height = 8, fig.align = "center"
)

# options(scipen=999)

# margin of plotly
m <- list(
  l = 150,
  r = 150,
  b = 100,
  t = 100,
  pad = 4
)


```



# Introduction

In this document I will analyze the data for **EXPIII** of HIV-1 LTEE dataset. Some of the analyses are reproduction of what has already been published and throughout the report I will compare
our findings with them. The link to the two paper's that have already been published are below:

- Fredrick's paper: https://academic.oup.com/mbe/article/36/11/2400/5524887?login=false
- Eva's paper: https://academic.oup.com/mbe/article/36/11/2400/5524887?login=false


EXPIII dataset includes genomic sequencing data of HIV-1 NL4-3 passages up to 570. There are four independent lines that have been passaged in two different host-cell lines, MT2 and MT4.



```{r loading libraries, echo=FALSE}

# loading libraries
library("ggplot2")
library("plotly")
library("DT")
library("dplyr")
library("tidyr")
library("Biostrings")
library("stringr")
library("magrittr")


```


```{r read-in the data}

# read in the data
variants_ann <- read.table(file = "/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/tables/2-p/all_annotated_variants.tsv.gz", 
sep = "\t", stringsAsFactors = FALSE, header = TRUE)

# set correct data types
variants_ann$exp_line <- as.integer(variants_ann$exp_line)


# set host cell line info
variants_ann$host_line <- ifelse(variants_ann$exp_line %in% c("13", "14","17", "18", "19", "20"), "MT2", "MT4")


# account for the switched experimental lines
variants_ann$exp_line[variants_ann$exp_line == 15 & variants_ann$passage > 360] <- "A"
variants_ann$exp_line[variants_ann$exp_line == 16 & variants_ann$passage > 360] <- "B"

variants_ann$exp_line[variants_ann$exp_line == "A"] <- "16"
variants_ann$exp_line[variants_ann$exp_line == "B"] <- "15"

# add a field containing pos, ref, and alt info of muts
variants_ann$mut_info <- paste(variants_ann$genomic_pos, variants_ann$ref_allele, variants_ann$alt_allele, sep = "_")

## only keep the EXPIII data

variants_ann_expiii <- variants_ann[variants_ann$exp_line %in% 13:16,]
end_passage <- 560
variants_ann_expiii <- variants_ann_expiii %>%
                              mutate(exp_line = case_when(exp_line==13 ~ "MT2_1", exp_line==14 ~ "MT2_2", exp_line==15 ~ "MT4_1", exp_line==16 ~ "MT4_2"))

variants_expiii <- variants_ann_expiii %>% filter(!duplicated(variant_id))

```



# Overview

## Minority Mutations



```{r number of min mutations}

category <- "Minority"
min_threshold <- 0.05
max_threshold <- 0.50


# get numbers for in-text reporting
variants_expiii_subset <- variants_expiii %>%
                                filter(allele_freq >= min_threshold & allele_freq <= max_threshold)

a <- nrow(variants_expiii_subset)
b <- length(unique(variants_expiii_subset$mut_info))

# prepare the data for statistical test
counts <- variants_expiii_subset %>%
              group_by(passage, host_line) %>%
              count()

# store the statistical test result
t_test_result <- t.test(counts$n[counts$host_line == "MT2"], counts$n[counts$host_line == "MT4"], paired = TRUE, alternative = "two.sided")


```

We observe a total of `r a` `r category` mutations across the four evolution lines of EXP III (`r b` unique mutations).

> `r category` mutations are the mutations that became reliably detectable in the population (a.k.a, allele frequency above `r min_threshold` and below `r max_threshold`)!

Of note, we observe slight deceleration of mutation accumulation towards the end of the experiment. Also we can see an overall higher number Of
`r category` mutations in MT4 host cell lines compared to MT2 (p_val of `r format(t_test_result$p.value, scientific=TRUE)`)



```{r time-wise min mut accumulation, results=TRUE}

# prepare data for visualization
counts <- variants_expiii %>%
              filter(allele_freq >= min_threshold & allele_freq < max_threshold) %>%
              group_by(exp_line, passage) %>%
              count() %>%
              dplyr::rename(mut_count = n) %>%
              mutate(host_line = ifelse(exp_line %in% c("MT2_1","MT2_2"), "MT2", "MT4")) %>%
              mutate_at(c(1,4), as.factor) %>%
              mutate_at(2:3, as.numeric)

# visualize using plotly


fig <- plot_ly(as.data.frame(counts), x = ~passage, y = ~mut_count, 
            type = 'scatter', mode = "lines", color = ~exp_line, linetype= ~host_line)
fig <- fig %>% layout(title = paste0(category, " Mutations Accumulation"),
                      xaxis = list(title = 'Passage'),
                      yaxis = list(title = 'Count'))

fig

```



The mutations spectra of all 4 lines look similar for minority mutations! Therefore, even though it has previously 
been shown that environment impacts the mutational profile of HIV-1, that affect does not change the overall mutational spectrum.

```{r min mutation spectrum, results=TRUE}

# prepare data for visualization
counts <- variants_expiii %>%
            filter(allele_freq >= min_threshold & allele_freq <= max_threshold) %>%
            filter(mut_type=="M") %>%
            mutate(mut_subtype=paste0(ref_allele, ">", alt_allele)) %>%
            group_by(exp_line, passage, mut_subtype) %>%
            count() %>%
            dplyr::rename(count=n) %>%
            group_by(exp_line,passage) %>%
            mutate(pct =  100 *count/sum(count)) %>%
            ungroup() %>%
            mutate_at(2:3, as.factor)


# visualize using ggplot
fig <- as.data.frame(counts) %>% ggplot(aes(x=passage, y = pct, fill = mut_subtype)) +
                                  geom_bar(position="stack", stat="identity") + 
                                  facet_wrap(~exp_line) +
                                  labs(x="Passage", y= "Percentage", fill = "Base Substitution")+
                                  theme_bw() +
                                  theme(axis.title = element_text(size = 20),
                                  axis.text.x = element_blank(),
                                  axis.text.y = element_text(size = 10),
                                  strip.text.x = element_text(size = 25),
                                  legend.title = element_text(size=15),
                                  legend.text = element_text(size = 10))



fig





```

```{r, include = FALSE}

png("/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/figs/presentation/270923/spectrum-iii.png"
    , width = 1600, height = 1200)
fig
dev.off()


```


## Majority Mutations

```{r number of maj mutations}


category <- "Majority"
threshold <- 0.5

# get numbers for in-text reporting
a <- nrow(variants_expiii[variants_expiii$passage == end_passage,])
b <- length(unique(variants_expiii$mut_info[variants_expiii$passage == end_passage]))

# prepare the data for statistical test
counts <- variants_expiii %>%
              filter(allele_freq >= threshold) %>%
              group_by(passage, host_line) %>%
              count()

# store the statistical test result
t_test_result <- t.test(counts$n[counts$host_line == "MT2"], counts$n[counts$host_line == "MT4"], paired = TRUE, alternative = "two.sided")

```

We observe a total of `r a` `r category` mutations across the four evolution lines of EXP III (`r b` unique mutations).

> `r category` mutations are the mutations that became more frequent in the population than the wild-type nucleotide (a.k.a, allele frequency above `r threshold`)!

Of note, we do observe deceleration of mutation accumulation towards the end of the experiment compared to first 100 passages. Another observation is the overall higher number of
`r category` mutations in MT2 host cell lines compared to MT4 (p_val of `r format(t_test_result$p.value, scientific=TRUE)`). This is while we see opposite trend for [Minority](#minority-mutations) mutations.


```{r time-wise maj mut accumulation, results=TRUE}


# prepare data for visualization
counts <- variants_expiii %>%
              filter(allele_freq >= threshold) %>%
              group_by(exp_line, passage) %>%
              count() %>%
              dplyr::rename(mut_count = n) %>%
              mutate(host_line = ifelse(exp_line %in% c("MT2_1", "MT2_2"), "MT2", "MT4")) %>%
              mutate_at(c(1,4), as.factor) %>%
              mutate_at(2:3, as.numeric)

# visualize using plotly
fig <- plot_ly(as.data.frame(counts), x = ~passage, y = ~mut_count, 
            type = 'scatter', mode = "lines", color = ~exp_line, linetype= ~host_line)
fig <- fig %>% layout(title = paste0(category, " Mutations Accumulation") ,
                      xaxis = list(title = 'Passage'),
                      yaxis = list(title = 'Count'))

fig


```



___
___

We can also plot the number of new `r category` mutations that appear or disappear compared to the immediate previous passage.

```{r prepare difference dataframe}


passages <- seq(10,570, 10)
diff_df <- data.frame()


for (i in 1:56){
    for (exp_line in c("MT2_1", "MT2_2", "MT4_1", "MT4_2")){
        prev_in <- passages[i]
        pres_in <- passages[i+1]

        prev <- variants_expiii$mut_info[variants_expiii$exp_line == exp_line & variants_expiii$passage == prev_in & variants_expiii$allele_freq >= 0.5]

        pres <- variants_expiii$mut_info[variants_expiii$exp_line == exp_line & variants_expiii$passage == pres_in & variants_expiii$allele_freq >= 0.5]

        addition <- length(setdiff(pres, prev))
        removal <- length(setdiff(prev,pres))

        diff_df <- rbind(diff_df, c(exp_line, passages[i+1], addition, removal))

    }
    
}

diff_df %<>% set_colnames(c("exp_line", "passage", "addition", "removal")) %>%
              arrange(exp_line) %>%
              mutate(host_line = ifelse(exp_line %in% c("MT2_1", "MT2_2"), "MT2", "MT4")) %>%
              mutate_at(c(1,5), as.factor) %>%
              mutate_at(2:4, as.numeric)


```


```{r}


# prepare the data for statistical test
counts <- diff_df %>%
            group_by(passage, host_line) %>%
            summarize(addition_mean = mean(addition),
                      removal_mean = mean(removal), .groups = 'drop')

# store the statistical test result
t_test_result_addition <- t.test(counts$addition_mean[counts$host_line == "MT2"], counts$addition_mean[counts$host_line == "MT4"], paired = TRUE, alternative = "two.sided")

t_test_result_removal <- t.test(counts$removal_mean[counts$host_line == "MT2"], counts$removal_mean[counts$host_line == "MT4"], paired = TRUE, alternative = "two.sided")

```


In MT2 lines we observe average of `r mean(counts$addition_mean[counts$host_line == "MT2"])` new `r category` mutations compared to 10 passage earlier while this number for HIV-1 evolved in MT4 is
`r mean(counts$addition_mean[counts$host_line == "MT4"])`. The number of new mutations per passage in MT2 is significantly higher than MT4 (paired T-test p_val of `r format(t_test_result_addition$p.value, scientific=TRUE)`)

Interestingly enough number of mutations disappearing from population compared to 10 previous passage in MT2 is also higher than MT4 (mean of `r mean(counts$removal_mean[counts$host_line == "MT2"])` and `r mean(counts$removal_mean[counts$host_line == "MT4"])`, respectively) 
which is also significant (paired T-test p_val of `r format(t_test_result_removal$p.value, scientific=TRUE)`)


```{r plot addition plot, results=TRUE}



# visualize using plotly
fig <- plot_ly(diff_df, x = ~passage, y = ~addition, 
            type = 'scatter', mode = "lines", color = ~exp_line, linetype= ~host_line)
fig <- fig %>% layout(title = paste0(category, " Mutations Emergence") ,
                      xaxis = list(title = 'Passage'),
                      yaxis = list(title = 'Count'))

fig


```



```{r plot removal plot, results=TRUE}



# visualize using plotly
fig <- plot_ly(as.data.frame(diff_df), x = ~passage, y = ~removal, 
            type = 'scatter', mode = "lines", color = ~exp_line, linetype= ~host_line)
fig <- fig %>% layout(title = paste0(category, " Mutations Disappearance") ,
                      xaxis = list(title = 'Passage'),
                      yaxis = list(title = 'Count'))

fig


```



___
___

If we plot the accumulation of `r category` mutations based on the effect type, we can visualize
the trends for S (synonymous), N (non-synonymous), and U (untranslated) mutations.

```{r mut effect plot, results=TRUE}


# prepare data for visualization
counts <- variants_ann_expiii%>% 
              filter(allele_freq >= threshold) %>%
              mutate(effect_simplified = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
              mutate(effect_simplified = replace_na(effect_simplified, "N")) %>%
              group_by(exp_line, passage, effect_simplified) %>%
              summarise(nr_majority = n()) %>%
              mutate_at(c(2,4), as.numeric) %>%
              mutate_at(c(1,3), as.factor)
              
# visualize using plotly

fig <- plot_ly(as.data.frame(counts), x = ~passage, y = ~nr_majority, 
              #text = ~paste("Car: ", car),
              type = 'scatter', mode = "lines", color = ~effect_simplified, linetype= ~exp_line)
fig <- fig %>% layout(title = paste0(category, " Mutation Accumulation Based on Translational Effect") ,
                      xaxis = list(title = 'Passage'),
                      yaxis = list(title = 'Count'))

fig


```



Alternatively, we can plot the same accumulation for different effect types but corrected for the 
genomic length that each mutation can occur in. As you can see even though the absolute number of 
synonymous and non-synonymous mutations are higher than mutations in untranslated regions per passage,
the density of mutations in untranslated regions is initially higher than protein-coding regions,
while the difference decreases towards the end of evolution. 


```{r statistics of maj muts in genomic features}


# prepare data for visualization
counts_df <- variants_ann_expiii%>% 
              filter(allele_freq >= threshold) %>%
              mutate(effect_simplified = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
              mutate(effect_simplified = replace_na(effect_simplified, "N"))


t_test_res <- c()
for (host_line in c("MT2", "MT4")){
  u <- c()
  s_n <- c()
  for (passage in seq(10, 570, by=10)){
    counts <- table(counts_df$effect_simplified[counts_df$passage == passage & counts_df$host_line == host_line])
    u <- append(u, as.numeric(counts["U"]))
    s_n <- append(s_n, sum(as.numeric(counts["S"]),as.numeric(counts["N"])))
  }
  # normalize for length
  u <- u*100/(9626-9076+789-454)
  s_n <- s_n*100/(9407-790)
  t_test_result <- t.test(u, s_n, paired = TRUE, alternative = "two.sided")
  t_test_res <- append(t_test_res, t_test_result$p.value)
}

names(t_test_res) <- c("MT2", "MT4")


```



Overall the higher frequency of `r category` mutations in
LTRs compared to protein-coding regions of the HIV-1 genome is significant in both host cell environments of MT2 (p.value of `r format(as.numeric(t_test_res["MT2"]), scientific = TRUE)`) and
of MT4 (p.value of `r format(as.numeric(t_test_res["MT4"]), scientific = TRUE)`).


```{r, results=TRUE}

# prepare data for visualization
counts <- variants_ann_expiii %>% 
            filter(allele_freq >= threshold) %>%
            mutate(effect_simplified = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
            mutate(effect_simplified = replace_na(effect_simplified, "N")) %>%
            mutate(effect_simplified2 = ifelse(effect_simplified %in% c("S","N"), "S/N", "U")) %>%
            group_by(exp_line, passage, effect_simplified2) %>%
            summarise(nr_majority = n()) %>%
            mutate_at(c(2,4), as.numeric) %>%
            mutate_at(c(1,3), as.factor) %>%
            mutate(nr_majority_density = ifelse(effect_simplified2 == "S/N", nr_majority*100/(9407-790),nr_majority*100/(9626-9076+789-454)))

# make the visualization
fig <- plot_ly(as.data.frame(counts), x = ~passage, y = ~nr_majority_density, 
              #text = ~paste("Car: ", car),
              type = 'scatter', mode = "lines", color = ~effect_simplified2, linetype= ~exp_line,
              colors = "Set1")
fig <- fig %>% layout(title = paste0("Normalized ", category, " Mutation Accumulation Based on Translational Effect") ,
                      xaxis = list(title = 'Passage'),
                      yaxis = list(title = 'Normalized Count'))

fig
```


___
___


Next we take a look at the number of majority mutations of passage `r end_passage` in different features of the HIV-1 genome:


```{r plot majority mut counts per feature, results=TRUE}


# prepare data for visualization
counts <- variants_ann_expiii %>% 
            filter(allele_freq >= threshold) %>%
            mutate(mut_type = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
            mutate(mut_type = replace_na(mut_type, "N")) %>%
            filter(passage == end_passage) %>%
            group_by(exp_line, feature, mut_type) %>%
            summarise(count = n()) %>%
            mutate_at(c(1,2,3), as.factor)

# make the visualization
fig <- ggplot(counts) +
  geom_bar(aes(x = mut_type, y = count, fill = exp_line),
           position = "stack",
           stat = "identity") +
  labs(x = "Genomic Features", y = "Count", fill = "Line")+
  facet_grid(~ feature, switch = "x") +
  theme(strip.placement = "outside",
        strip.background = element_rect(fill = NA, color = "white"),
        panel.spacing = unit(-.01,"cm"))+
  theme(axis.title = element_text(size = 20),
  axis.text.x = element_text(size = 10),
  axis.text.y = element_text(size = 10),
  strip.text.x = element_text(size = 15, angle=90),
  legend.text=element_text(size=15),
  legend.title = element_text(size=10))


fig
```


Looking at the frequency of different mutations types in the genomic regions (normalized
for the length of the genome):


```{r plot normalized majority mut counts per feature, results=TRUE}


# based on feature range
feature_length <- data.frame(feature = c("5UTR", "gag", "pol", "vif","vpr", "tat","rev","vpu","envelope","nef","3UTR"),
                            feature_length = c(335,1503,3012,579,291,261,351,246,2565,621,550))


# prepare data for visualization
counts <- variants_ann_expiii %>% 
            filter(allele_freq >= threshold) %>%
            mutate(mut_type = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
            mutate(mut_type = replace_na(mut_type, "N")) %>%
            filter(passage == end_passage) %>%
            group_by(exp_line, feature, mut_type) %>%
            summarise(count = n()) %>%
            mutate_at(c(1,2,3), as.factor) %>%
            left_join(feature_length, by = join_by(feature == feature)) %>%
            mutate(normalized_count = count*100/feature_length)


# make the visualization
fig <- ggplot(counts) +
  geom_bar(aes(x = mut_type, y = normalized_count, fill = exp_line),
           position = "stack",
           stat = "identity") +
  labs(x = "Genomic Features", y = "Normalized Count", fill = "Line")+
  facet_grid(~ feature, switch = "x") +
  theme(strip.placement = "outside",
        strip.background = element_rect(fill = NA, color = "white"),
        panel.spacing = unit(-.01,"cm"))+
  theme(axis.title = element_text(size = 20),
  axis.text.x = element_text(size = 10),
  axis.text.y = element_text(size = 10),
  strip.text.x = element_text(size = 15, angle=90),
  legend.text=element_text(size=15),
  legend.title = element_text(size=10))

fig
```

```{r, include = FALSE}

png("/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/figs/presentation/270923/fearure-iii.png"
    , width = 1600, height = 1200)
fig
dev.off()


```


Additionally we can look at the normalized count of mutations per genomic features for previous passages
to get a sense for whether and how they are changing over time. In the following you can view the plots for
passages 100, 200, 300, and 400. (for a better visualization view the pdf version)



```{r plot normalized majority mut counts per feature for past passages, results=TRUE}


# based on feature range
feature_length <- data.frame(feature = c("5UTR", "gag", "pol", "vif","vpr", "tat","rev","vpu","envelope","nef","3UTR"),
                            feature_length = c(335,1503,3012,579,291,261,351,246,2565,621,550))


# prepare data for visualization
counts <- variants_ann_expiii %>% 
            filter(allele_freq >= threshold) %>%
            mutate(mut_type = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
            mutate(mut_type = replace_na(mut_type, "N")) %>%
            filter(passage %in% c(100,200,300,400, 560)) %>%
            group_by(exp_line, passage, feature, mut_type) %>%
            summarise(count = n()) %>%
            mutate_at(1:4, as.factor) %>%
            left_join(feature_length, by = join_by(feature == feature)) %>%
            mutate(normalized_count = count*100/feature_length)


# make the visualization
fig <- ggplot(counts) +
  geom_bar(aes(x = mut_type, y = normalized_count, fill = exp_line),
           position = "stack",
           stat = "identity") +
  labs(x = "Genomic Features", y = "Normalized Count", fill = "Line")+
  facet_grid(~ feature + passage, switch = "x") +
  theme(strip.placement = "outside",
        strip.background = element_rect(fill = NA, color = "white"),
        panel.spacing = unit(-.01,"cm"))+
  theme(axis.title = element_text(size = 20),
  axis.text.x = element_text(size = 10),
  axis.text.y = element_text(size = 10),
  strip.text.x = element_text(size = 15, angle=90),
  legend.text=element_text(size=15),
  legend.title = element_text(size=10))

fig
```



```{r, include = FALSE}

pdf("/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/figs/pdf/fearure-count-history-iii.pdf"
    , width = 24.5, height = 14)
fig
dev.off()


```







## Mutational profile


In the chart below you can see the mutation subtypes of all/unique majority and 
minority mutations of all 4 lines of EXPIII.


```{r mut_profile pie-chart, results=TRUE}


# add subtype info
variants_expiii_sub <-  variants_expiii %>% mutate(mut_type_sub = ifelse(mut_type == "M", paste0(ref_allele,">",alt_allele), mut_type))

# minority

subset_name <- "Minority"
min_threshold <- 0.05
max_threshold <- 0.50


# prepare data for visualization (unique mutations)
mut_profile_min_unique <- variants_expiii_sub %>% filter(allele_freq >= min_threshold & allele_freq <= max_threshold) %>%
                            select(mut_info, mut_type_sub) %>%
                            distinct() %>%
                            group_by(mut_type_sub) %>%
                            summarise(count = n())
#                            rename(mut_sub_type = mut_type_sub)



# prepare data for visualization (all mutations)
mut_profile_min_all <- variants_expiii_sub %>% filter(allele_freq >= min_threshold & allele_freq <= max_threshold) %>%
                            group_by(mut_type_sub) %>%
                            summarise(count = n())
#                            rename(mut_sub_type = mut_type_sub)



# make the visualization
fig <- plot_ly()
fig <- fig %>% add_pie(data = mut_profile_min_unique, labels = ~mut_type_sub, values = ~count, title = paste0("All Mutations", " (", sum(mut_profile_min_all$count),")"),
                       name = "Cut", domain = list(x = c(0, 0.4), y = c(0.4, 1.4)))
fig <- fig %>% add_pie(data = mut_profile_min_all, labels = ~mut_type_sub, values = ~count, title = paste0("Unique Mutations", " (", sum(mut_profile_min_unique$count),")"),
                       name = "Color", domain = list(x = c(0.6, 1), y = c(0.4, 1.4))) %>%
                       layout(title=paste0('Mutation Sub-Types _ ', subset_name))
fig



# majority

subset_name <- "Majority"
threshold <- 0.5

# prepare data for visualization
mut_profile_maj_unique <- variants_expiii_sub %>% filter(allele_freq >= threshold) %>%
                            select(mut_info, mut_type_sub) %>%
                            distinct() %>%
                            group_by(mut_type_sub) %>%
                            summarise(count = n())


# prepare data for visualization
mut_profile_maj_all <- variants_expiii_sub %>% filter(allele_freq >= threshold) %>%
                            group_by(mut_type_sub) %>%
                            summarise(count = n())

# make the visualization
fig <- plot_ly()
fig <- fig %>% add_pie(data = mut_profile_maj_unique, labels = ~mut_type_sub, values = ~count, title = paste0("All Mutations", " (", sum(mut_profile_maj_all$count),")"),
                       name = "Cut", domain = list(x = c(0, 0.4), y = c(0.4, 1)))
fig <- fig %>% add_pie(data = mut_profile_maj_all, labels = ~mut_type_sub, values = ~count, title = paste0("Unique Mutations", " (", sum(mut_profile_maj_unique$count),")"),
                       name = "Color", domain = list(x = c(0.6, 1), y = c(0.4, 1))) %>%
                       layout(title=paste0('Mutation Sub-Types _ ', subset_name))
fig

```


```{r genome composition}

# read in ancestral reference sequence 
fastaFile <- readDNAStringSet("/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/data/reference/plasmid/hiv_plasmid_ref_genome.fasta")
seq_name = names(fastaFile)
sequence = paste(fastaFile)

# get the guanine composition of the reference genome
g_percent <- str_count(sequence, pattern = "G")/nchar(sequence)
```

<!--
As it is evident almost `r #mut_profile_min$count[mut_profile_min$mut_sub_type == "G>A"]/nrow(variants_expiii_min)`% of minority and 
`r #mut_profile_maj$count[mut_profile_maj$mut_sub_type == "G>A"]/nrow(variants_expiii_maj)`% of majority mutations are "G>A" mutations. This is while Guanine constitutes 
`r #g_percent`% of HIV-1 genome.
-->




### APOBEC3G

```{r proportion of G2A mutations}

subset_name <- "Minority"
min_threshold <- 0.05
max_threshold <- 0.50


variants_expiii_sub <- variants_expiii %>%
                        filter(allele_freq >= min_threshold & allele_freq <= max_threshold)

mt2_g2a <- nrow(variants_expiii_sub[variants_expiii_sub$ref_allele == "G" & variants_expiii_sub$alt_allele == "A" & variants_expiii_sub$passage == end_passage & variants_expiii_sub$exp_line %in% c("MT2_1", "MT2_2"),])
mt2_g2a_percent <- mt2_g2a/nrow(variants_expiii_sub[variants_expiii_sub$passage == end_passage & variants_expiii_sub$exp_line %in% c("MT2_1", "MT2_2"),])


mt4_g2a <- nrow(variants_expiii_sub[variants_expiii_sub$ref_allele == "G" & variants_expiii_sub$alt_allele == "A" & variants_expiii_sub$passage == end_passage & variants_expiii_sub$exp_line %in% c("MT4_1", "MT4_2"),])
mt4_g2a_percent <- mt4_g2a/nrow(variants_expiii_sub[variants_expiii_sub$passage == end_passage & variants_expiii_sub$exp_line %in% c("MT4_1", "MT4_2"),])




```


We also analyzed the effect of APOBEC3G on the accumulation of `r category` mutations. 
G-to-A mutations in the HIV-1 viral genome are mainly induced by APOBEC3G/F which is only expressed in MT-2 T-cells and not in MT-4 T-cells (Borman et al. 1995).
Expectedly, even though the absolute number of G to A `r category` mutations at the end of the experiment is slightly lower in HIV-1 growing in MT-2 T-cells than in HIV-1 growing in MT-4 T-cells (`r mt2_g2a` and `r mt4_g2a` G to A mutations, respectively),
the percentage of G-A mutations in MT2 lines are slightly higher than in MT4 (`r mt2_g2a_percent`% and `r mt4_g2a_percent`% G to A mutations, respectively) .



```{r g2a mutations in context}


mt2_context <- table(variants_expiii_sub$mut_context[variants_expiii_sub$ref_allele == "G" & variants_expiii_sub$alt_allele == "A" & variants_expiii_sub$passage == 160 & variants_expiii_sub$exp_line %in% c("MT2_1", "MT2_2")])
mt4_context <- table(variants_expiii_sub$mut_context[variants_expiii_sub$ref_allele == "G" & variants_expiii_sub$alt_allele == "A" & variants_expiii_sub$passage == 160 & variants_expiii_sub$exp_line %in% c("MT4_1", "MT4_2")])


(as.numeric(sum(mt2_context[str_sub(names(mt2_context),1,1) == "T"])))/sum(mt2_context)
(as.numeric(sum(mt4_context[str_sub(names(mt4_context),1,1) == "T"])))/sum(mt4_context)



counts <- variants_expiii_sub %>%
            filter(variants_expiii_sub$ref_allele == "G" & variants_expiii_sub$alt_allele == "A") %>%
            group_by(passage, host_line) %>%
            mutate(count_all = n()) %>%
            ungroup() %>%
            mutate_at(c("passage", "host_line"), as.factor) %>%
            filter(mut_3_context %in% c("GGG", "AAG")) %>%
            group_by(passage, host_line, count_all, .drop=FALSE) %>%
            summarise(count_g2a = n()) %>%
            mutate(percentage_g2a = count_g2a/count_all) %>%
            arrange(passage, host_line)


mt2_pct <- counts$percentage_g2a[counts$host_line == "MT2"]
mt4_pct <- counts$percentage_g2a[counts$host_line == "MT4"]

t_test_result <- t.test(mt2_pct, mt4_pct, paired = TRUE, alternative = "two.sided")


```


According to this [article](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5391955/), G-A mutations caused by APOBEC3G/F are mainly in GGG or AAG context, therefore we can expect more proportion of G-A mutations in MT2 compared to MT4
to be in those contexts. As opposed to our expectation, proportion of G-A mutations occurring in GG[G]/AA[G] context
in MT2 is not significantly higher than in MT4 (p_val of `r t_test_result$p.value`). That said,
it might be of interest to compare the frequency and context of G-A mutations that are hypermutants (2 or more G-A mutations in in the same read pair). Also based on that we might 
be able to define hotspot regions for these mutations as well. These are some extra analyses that I will do in the future to further delineate the impact of APOBEC mutagenesis
on the mutational profile of HIV-1 virus.


