---
title:  
    "Parallel Evolution Analysis"
date: 
    "Oct 27th, 2023"
author:  
    "Ali Movasati"  
output:  
    html_document:
        toc: true
        toc_depth: 3
        toc_float: true
        number_sections: true
        theme: lumen
        highlight: zenburn
        code_folding: hide
---




```{r setting global chunk parameters, include=FALSE}

# options for code chunk and output inclusion 
knitr::opts_chunk$set(
  message = FALSE, echo = TRUE, warning=FALSE, results = "hide"
)

# options for figures
knitr::opts_chunk$set(
  fig.width = 8, fig.height = 8, fig.align = "center"
)

# options(scipen=999)

# margin of plotly
m <- list(
  l = 150,
  r = 150,
  b = 100,
  t = 100,
  pad = 4
)


```




```{r loading libraries, echo=FALSE}

# loading libraries
library("plotly")
library("ggplot2")
library("DT")
library("dplyr")
library("tidyr")
library("Biostrings")
library("stringr")
library("magrittr")
library("ggvenn")
library("VennDiagram")
library("ggVennDiagram")


```


# Introduction

In this document I will try to provide an overview of parallel evolution among different lines of each LTEE experiment. I'll try to dissect the extent of parallelism in HIV-1 evolution and depict a detailed
picture of its characteristics.




```{r read-in the data}

# read in the data
variants_ann <- read.table(file = "/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/tables/2-p/all_annotated_variants.tsv.gz", 
sep = "\t", stringsAsFactors = FALSE, header = TRUE)

# set correct data types
variants_ann$exp_line <- as.integer(variants_ann$exp_line)


# set host cell line info
variants_ann$host_line <- ifelse(variants_ann$exp_line %in% c("13", "14","17", "18", "19", "20"), "MT2", "MT4")


# account for the switched experimental lines
variants_ann$exp_line[variants_ann$exp_line == 15 & variants_ann$passage > 360] <- "A"
variants_ann$exp_line[variants_ann$exp_line == 16 & variants_ann$passage > 360] <- "B"

variants_ann$exp_line[variants_ann$exp_line == "A"] <- "16"
variants_ann$exp_line[variants_ann$exp_line == "B"] <- "15"



variants_ann$exp_line[variants_ann$exp_line == 18 & variants_ann$passage > 350] <- "A"
variants_ann$exp_line[variants_ann$exp_line == 19 & variants_ann$passage > 350] <- "B"

variants_ann$exp_line[variants_ann$exp_line == "A"] <- "19"
variants_ann$exp_line[variants_ann$exp_line == "B"] <- "18"


# add a field containing pos, ref, and alt info of muts
variants_ann$mut_info <- paste(variants_ann$genomic_pos, variants_ann$ref_allele, variants_ann$alt_allele, sep = "_")

## EXPIII data

variants_ann_expiii <- variants_ann[variants_ann$exp_line %in% 13:16,]
variants_ann_expiii <- variants_ann_expiii %>%
                            mutate(exp_line = case_when(exp_line==13 ~ "MT2_1", exp_line==14 ~ "MT2_2", exp_line==15 ~ "MT4_1", exp_line==16 ~ "MT4_2"))

variants_expiii <- variants_ann_expiii %>% filter(!duplicated(variant_id))



## EXPIV data

variants_ann_expiv <- variants_ann[variants_ann$exp_line %in% 17:20,]
variants_ann_expiv <- variants_ann_expiv %>%
                            mutate(exp_line = case_when(exp_line==17 ~ "MT2_1", exp_line==18 ~ "MT2_2", exp_line==19 ~ "MT2_3", exp_line==20 ~ "MT2_4"))

variants_expiv <- variants_ann_expiv %>% filter(!duplicated(variant_id))


```

# Overview

## EXPIII

```{r}

exp <- "EXPIII"
category <- "Majority"
threshold <- 0.5
end_passage <- 560

# keep majority mutations
variants_expiii %<>% 
                        filter(allele_freq >= threshold)


# uniqueness of majority mutations observed throughout the experiment
throughout_maj <- variants_expiii %>%
                        select(c(exp_line,mut_info)) %>%
                        distinct()


#table(as.numeric(table(par_whole$mut_info)))

# create a named vector from the df
muts <- throughout_maj$mut_info
names(muts) <- throughout_maj$exp_line

# convert named vector to list
muts <- lapply(split(muts, names(muts)), unname)


# visualization
ggvenn(muts) +
        labs(title = paste0(category, " Mutations throughout the ", exp))+
        theme(plot.title = element_text(size = 20, hjust = 0.5),
        plot.margin = unit(c(1,1,1,1), "cm"))




# uniqueness of majority mutations present at the end passage
end_maj <- variants_expiii %>%
                        filter(passage == end_passage) %>%
                        select(c(exp_line,mut_info)) %>%
                        distinct()



#table(as.numeric(table(par$mut_info)))


# create a named vector from the df
muts <- end_maj$mut_info
names(muts) <- end_maj$exp_line

# convert named vector to list
muts <- lapply(split(muts, names(muts)), unname)


# visualization
ggvenn(muts) +
        labs(title = paste0(category, " Mutations at passage #", end_passage))+
        theme(plot.title = element_text(size = 20, hjust = 0.5),
        plot.margin = unit(c(1,1,1,1), "cm"))

```


```{r number of maj mutations}


# get numbers for in-text reporting

a <- nrow(variants_expiii[variants_expiii$passage == end_passage,])
b <- length(unique(variants_expiii$mut_info[variants_expiii$passage == end_passage]))



```

We observe a total of `r a` `r category` mutations across the four evolution lines of EXP III (`r b` unique mutations) at passage `r end_passage`.


## EXPIV

```{r }
exp <- "EXPIV"
category <- "Majority"
threshold <- 0.5
end_passage <- 510

# keep majority mutations
variants_expiv %<>% 
                        filter(allele_freq >= threshold)


# uniqueness of majority mutations observed throughout the experiment
throughout_maj <- variants_expiv %>%
                        select(c(exp_line,mut_info)) %>%
                        distinct()

nrow(throughout_maj)
#table(as.numeric(table(par_whole$mut_info)))

# create a named vector from the df
muts <- throughout_maj$mut_info
names(muts) <- throughout_maj$exp_line

# convert named vector to list
muts <- lapply(split(muts, names(muts)), unname)


# visualization
ggvenn(muts) +
        labs(title = paste0(category, " Mutations throughout the ", exp))+
        theme(plot.title = element_text(size = 20, hjust = 0.5),
        plot.margin = unit(c(1,1,1,1), "cm"))




# uniqueness of majority mutations present at the end passage
end_maj <- variants_expiv %>%
                        filter(passage == end_passage) %>%
                        select(c(exp_line,mut_info)) %>%
                        distinct()



#table(as.numeric(table(par$mut_info)))


# create a named vector from the df
muts <- end_maj$mut_info
names(muts) <- end_maj$exp_line

# convert named vector to list
muts <- lapply(split(muts, names(muts)), unname)


# visualization
ggvenn(muts) +
        labs(title = paste0(category, " Mutations at passage #", end_passage))+
        theme(plot.title = element_text(size = 20, hjust = 0.5),
        plot.margin = unit(c(1,1,1,1), "cm"))



```





## MT2 of EXPIII and EXPIV

```{r , results = T}
exp <- "EXPIII/IV"
category <- "Majority"
threshold <- 0.5
end_passage <- 510



variants_ann_all <- variants_ann %>%
                            mutate(exp_line = case_when(exp_line==13 ~ "III_MT2_1", exp_line==14 ~ "III_MT2_2", exp_line==15 ~ "III_MT4_1", exp_line==16 ~ "III_MT4_2",
                            exp_line==17 ~ "IV_MT2_1", exp_line==18 ~ "IV_MT2_2", exp_line==19 ~ "IV_MT2_3", exp_line==20 ~ "IV_MT2_4"))

variants_all <- variants_ann_all %>% filter(allele_freq >= threshold & str_detect(exp_line, pattern = "MT2"))



# uniqueness of majority mutations observed throughout the experiment
throughout_maj <- variants_all %>%
                        select(c(exp_line,mut_info)) %>%
                        distinct()


#table(as.numeric(table(par_whole$mut_info)))

# create a named vector from the df
muts <- throughout_maj$mut_info
names(muts) <- throughout_maj$exp_line

# convert named vector to list
muts <- lapply(split(muts, names(muts)), unname)
# visualization
ggVennDiagram(muts) +
        labs(title = paste0(category, " Mutations throughout the ", exp))+
        theme(plot.title = element_text(size = 20, hjust = 0.5),
        plot.margin = unit(c(1,1,1,1), "cm"))




# uniqueness of majority mutations present at the end passage
end_maj <- variants_all %>%
                        filter(passage == end_passage) %>%
                        select(c(exp_line,mut_info)) %>%
                        distinct()



#table(as.numeric(table(par$mut_info)))


# create a named vector from the df
muts <- end_maj$mut_info
names(muts) <- end_maj$exp_line

# convert named vector to list
muts <- lapply(split(muts, names(muts)), unname)


# visualization
ggVennDiagram(muts) +
        labs(title = paste0(category, " Mutations at passage #", end_passage))+
        theme(plot.title = element_text(size = 20, hjust = 0.5),
        plot.margin = unit(c(1,1,1,1), "cm"))



```