---
title:  
    "Parallel Evolution Analysis"
date: 
    "Oct 27th, 2023"
author:  
    "Ali Movasati"  
output:  
    html_document:
        toc: true
        toc_depth: 3
        toc_float: true
        number_sections: true
        theme: lumen
        highlight: zenburn
        code_folding: hide
---




```{r setting global chunk parameters, include=FALSE}

# options for code chunk and output inclusion 
knitr::opts_chunk$set(
  message = FALSE, echo = TRUE, warning=FALSE, results = "hide"
)

# options for figures
knitr::opts_chunk$set(
  fig.width = 8, fig.height = 8, fig.align = "center"
)

# options(scipen=999)

# margin of plotly
m <- list(
  l = 150,
  r = 150,
  b = 100,
  t = 100,
  pad = 4
)


```




```{r loading libraries, echo=FALSE}

# loading libraries
library("plotly")
library("ggplot2")
library("DT")
library("dplyr")
library("tidyr")
library("Biostrings")
library("stringr")
library("magrittr")
library("ggvenn")
library("VennDiagram")
library("ggVennDiagram")


```


# Introduction

In this document I will try to provide an overview of parallel evolution among different lines of each LTEE experiment. I'll try to dissect the extent of parallelism in HIV-1 evolution and depict a detailed
picture of its characteristics.




```{r read-in the data}

# read in the data
variants_ann <- read.table(file = "/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/tables/2-p/all_annotated_variants.tsv.gz", 
sep = "\t", stringsAsFactors = FALSE, header = TRUE)

# set correct data types
variants_ann$exp_line <- as.integer(variants_ann$exp_line)


# set host cell line info
variants_ann$host_line <- ifelse(variants_ann$exp_line %in% c("13", "14","17", "18", "19", "20"), "MT2", "MT4")


# account for the switched experimental lines
variants_ann$exp_line[variants_ann$exp_line == 15 & variants_ann$passage > 360] <- "A"
variants_ann$exp_line[variants_ann$exp_line == 16 & variants_ann$passage > 360] <- "B"

variants_ann$exp_line[variants_ann$exp_line == "A"] <- "16"
variants_ann$exp_line[variants_ann$exp_line == "B"] <- "15"



variants_ann$exp_line[variants_ann$exp_line == 18 & variants_ann$passage > 350] <- "A"
variants_ann$exp_line[variants_ann$exp_line == 19 & variants_ann$passage > 350] <- "B"

variants_ann$exp_line[variants_ann$exp_line == "A"] <- "19"
variants_ann$exp_line[variants_ann$exp_line == "B"] <- "18"


# add a field containing pos, ref, and alt info of muts
variants_ann$mut_info <- paste(variants_ann$genomic_pos, variants_ann$ref_allele, variants_ann$alt_allele, sep = "_")

## EXPIII data
threshold <- 0.5

variants_ann_expiii <- variants_ann[variants_ann$exp_line %in% 13:16,]
variants_ann_expiii <- variants_ann_expiii %>%
                        mutate(exp_line = case_when(exp_line==13 ~ "MT2_1", exp_line==14 ~ "MT2_2", exp_line==15 ~ "MT4_1", exp_line==16 ~ "MT4_2")) %>%
                        filter(allele_freq >= threshold)

variants_expiii <- variants_ann_expiii %>% filter(!duplicated(variant_id))



## EXPIV data

variants_ann_expiv <- variants_ann[variants_ann$exp_line %in% 17:20,]
variants_ann_expiv <- variants_ann_expiv %>%
                        mutate(exp_line = case_when(exp_line==17 ~ "MT2_1", exp_line==18 ~ "MT2_2", exp_line==19 ~ "MT2_3", exp_line==20 ~ "MT2_4")) %>%
                        filter(allele_freq >= threshold)

variants_expiv <- variants_ann_expiv %>% filter(!duplicated(variant_id))



## table option

tbl_opt_list <- list(pageLength = 10, columnDefs = list(list(targets = 1, render = JS(
    "function(data, type, row, meta) {",
    "return type === 'display' && data.length > 10 ?",
    "'<span title=\"' + data + '\">' + data.substr(0, 10) + '...</span>' : data;",
    "}"))))

```

# Overview

## EXPIII

```{r throughout expiii data (venn diagram), results=T}

exp <- "EXPIII"
category <- "Majority"
threshold <- 0.5
end_passage <- 560



# uniqueness of majority mutations observed throughout the experiment
throughout_maj <- variants_expiii %>%
                        select(c(exp_line,mut_info)) %>%
                        distinct()


#table(as.numeric(table(par_whole$mut_info)))

# create a named vector from the df
muts <- throughout_maj$mut_info
names(muts) <- throughout_maj$exp_line

# convert named vector to list
muts_list <- lapply(split(muts, names(muts)), unname)


# visualization
ggvenn(muts_list) +
        labs(title = paste0(category, " Mutations throughout the ", exp))+
        theme(plot.title = element_text(size = 20, hjust = 0.5),
        plot.margin = unit(c(1,1,1,1), "cm"))


```


```{r throughout expiii data (table), results=T}
variants_expiii_subset <- variants_ann_expiii %>% select(c("mut_type", "feature_id", "feature", "feature_range", "ORF", "effect", "impact", "aa_change", "loc_in_cds", "loc_in_protein", "loss_of_function", "nonsense_mediated_decay", "mut_context", "mut_3_context", "mut_5_context", "mut_info")) %>%
                                                distinct()


counts_sep_df <- as.data.frame(table(muts, names(muts)))
counts_sep_df %<>% spread(Var2, Freq)
counts_sum_df <- as.data.frame(table(muts))


counts_df <- merge(counts_sep_df, counts_sum_df, by = "muts")


counts_df_complete <- merge(counts_df, variants_expiii_subset, by.x = "muts", by.y = "mut_info")


counts_df_complete <- counts_df_complete[order(counts_df_complete$Freq, decreasing = T),]
rownames(counts_df_complete) <- 1:nrow(counts_df_complete)


datatable(counts_df_complete)


```



```{r number of maj mutations throughout}


# get numbers for in-text reporting

a <- nrow(variants_expiii)
b <- length(unique(variants_expiii$mut_info))



```


> We observe a total of `r a` `r category` mutations across the four evolution lines of `r exp` (`r b` unique mutations) throughout the experiment.






```{r end passage expiii data (venn diagram), results=T}
# uniqueness of majority mutations present at the end passage
end_maj <- variants_expiii %>%
                        filter(passage == end_passage) %>%
                        select(c(exp_line,mut_info)) %>%
                        distinct()



#table(as.numeric(table(par$mut_info)))


# create a named vector from the df
muts <- end_maj$mut_info
names(muts) <- end_maj$exp_line

# convert named vector to list
muts_list <- lapply(split(muts, names(muts)), unname)


# visualization
ggvenn(muts_list) +
        labs(title = paste0(category, " Mutations at passage #", end_passage))+
        theme(plot.title = element_text(size = 20, hjust = 0.5),
        plot.margin = unit(c(1,1,1,1), "cm"))



```


```{r end passage expiii data (table), results=T}
variants_expiii_subset <- variants_ann_expiii %>% select(c("mut_type", "feature_id", "feature", "feature_range", "ORF", "effect", "impact", "aa_change", "loc_in_cds", "loc_in_protein", "loss_of_function", "nonsense_mediated_decay", "mut_context", "mut_3_context", "mut_5_context", "mut_info")) %>%
                                                distinct()


counts_sep_df <- as.data.frame(table(muts, names(muts)))
counts_sep_df %<>% spread(Var2, Freq)
counts_sum_df <- as.data.frame(table(muts))


counts_df <- merge(counts_sep_df, counts_sum_df, by = "muts")


counts_df_complete <- merge(counts_df, variants_expiii_subset, by.x = "muts", by.y = "mut_info")


counts_df_complete <- counts_df_complete[order(counts_df_complete$Freq, decreasing = T),]
rownames(counts_df_complete) <- 1:nrow(counts_df_complete)


datatable(counts_df_complete)


```


```{r number of maj mutations end passage}


# get numbers for in-text reporting

a <- nrow(variants_expiii[variants_expiii$passage == end_passage,])
b <- length(unique(variants_expiii$mut_info[variants_expiii$passage == end_passage]))



```


> We observe a total of `r a` `r category` mutations across the four evolution lines of `r exp` (`r b` unique mutations) at passage `r end_passage`.


### Evolution of single mutations

```{r define plotting function}

plot_mut_af <- function(data, mutations){


        # check whether the inputted data is from expiii or expiv

        if (any(data$host_line == "MT4")){
                exp <- "EXPIII"
                end_passage <- 570
        } else {exp <- "EXPIV"
                end_passage <- 510}

        final_df <- data.frame()
        for (mut in mutations){
                
                df <- data %>% filter(data$mut_info == mut)
                
                for (exp_line in unique(df$exp_line)){
                        for (psg in seq(0,end_passage,10)){
                                if (nrow(df[df$exp_line == exp_line & df$passage == psg,]) == 0){
                                addition <- df[1,]
                                addition[5] <- 0
                                addition[7] <- psg
                                addition[8] <- exp_line
                                df <- rbind(df, addition)
                        }
                }
                }
                df <- df[order(df$exp_line, df$passage),]
                
                final_df <- rbind(final_df, df)
        }

        plt <- ggplot(final_df, aes(x=passage, y=allele_freq, color = exp_line)) +
        facet_wrap(~ mut_info, ncol = 2) +
        geom_line() +
        labs(title = exp) +
        theme_bw() +
        theme(plot.title = element_text(size = 20, hjust = 0.5))
        
        ggplotly(plt)
}

```


```{r plot af evolution for certain mutations in EXPIII, results = T}



variants_ann_expiii <- variants_ann[variants_ann$exp_line %in% 13:16,]
variants_ann_expiii <- variants_ann_expiii %>%
                        mutate(exp_line = case_when(exp_line==13 ~ "MT2_1", exp_line==14 ~ "MT2_2", exp_line==15 ~ "MT4_1", exp_line==16 ~ "MT4_2"))

variants_expiii <- variants_ann_expiii %>% filter(!duplicated(variant_id))



plot_mut_af(variants_expiii, c("6238_G_A", "9412_G_A", "9528_T_G"))



```



9412_G_A abruptly is removed from the population in line MT2_1. To see if possibly the rise of another mutation in the population
might have caused this phenomena or not I looked at the allele frequency dynamics of majority mutations absent in passage 330 
and present in 340 in line MT2_1.


```{r additional, results = T}

plot_mut_af(variants_expiii, c("5436_G_A", "8270_A_G", "9292_G_A", "9293_C_G"))

```



Another abnormal observation is the absence of 9528_T_G in line MT4_2 (line number 15) in only two passages 490 and 500. After 
checking the mapping files I realized the reason of these mutations is not being detected is the low coverage towards the end of the virus genome.
The coverage of the sequencing in these two samples fall between 900-1000 from position 9520. 

> What should we do to address this issue?


## EXPIV

```{r throughout expiv data (venn diagram)}
exp <- "EXPIV"
category <- "Majority"
threshold <- 0.5
end_passage <- 510


# uniqueness of majority mutations observed throughout the experiment
throughout_maj <- variants_expiv %>%
                        select(c(exp_line,mut_info)) %>%
                        distinct()


# table(as.numeric(table(par_whole$mut_info)))

# create a named vector from the df
muts <- throughout_maj$mut_info
names(muts) <- throughout_maj$exp_line

# convert named vector to list
muts_list <- lapply(split(muts, names(muts)), unname)


# visualization
ggvenn(muts_list) +
        labs(title = paste0(category, " Mutations throughout the ", exp))+
        theme(plot.title = element_text(size = 20, hjust = 0.5),
        plot.margin = unit(c(1,1,1,1), "cm"))
```


```{r throughout expiv data (table), results=T}
variants_expiv_subset <- variants_ann_expiv %>% select(c("mut_type", "feature_id", "feature", "feature_range", "ORF", "effect", "impact", "aa_change", "loc_in_cds", "loc_in_protein", "loss_of_function", "nonsense_mediated_decay", "mut_context", "mut_3_context", "mut_5_context", "mut_info")) %>%
                                                distinct()


counts_sep_df <- as.data.frame(table(muts, names(muts)))
counts_sep_df %<>% spread(Var2, Freq)
counts_sum_df <- as.data.frame(table(muts))


counts_df <- merge(counts_sep_df, counts_sum_df, by = "muts")


counts_df_complete <- merge(counts_df, variants_expiv_subset, by.x = "muts", by.y = "mut_info")


counts_df_complete <- counts_df_complete[order(counts_df_complete$Freq, decreasing = T),]
rownames(counts_df_complete) <- 1:nrow(counts_df_complete)


datatable(counts_df_complete)


```



```{r number of maj mutations expiii throughout}


# get numbers for in-text reporting

a <- nrow(variants_expiv)
b <- length(unique(variants_expiv$mut_info))



```


> We observe a total of `r a` `r category` mutations across the four evolution lines of `r exp` (`r b` unique mutations) throughout the experiment.






```{r end passage expiv data (venn diagram), results=T}




# uniqueness of majority mutations present at the end passage
end_maj <- variants_expiv %>%
                        filter(passage == end_passage) %>%
                        select(c(exp_line,mut_info)) %>%
                        distinct()



#table(as.numeric(table(par$mut_info)))


# create a named vector from the df
muts <- end_maj$mut_info
names(muts) <- end_maj$exp_line

# convert named vector to list
muts_list <- lapply(split(muts, names(muts)), unname)


# visualization
ggvenn(muts_list) +
        labs(title = paste0(category, " Mutations at passage #", end_passage))+
        theme(plot.title = element_text(size = 20, hjust = 0.5),
        plot.margin = unit(c(1,1,1,1), "cm"))



```


```{r end passage expiv data (table), results=T}

counts_sep_df <- as.data.frame(table(muts, names(muts)))
counts_sep_df %<>% spread(Var2, Freq)
counts_sum_df <- as.data.frame(table(muts))


counts_df <- merge(counts_sep_df, counts_sum_df, by = "muts")


counts_df_complete <- merge(counts_df, variants_expiv_subset, by.x = "muts", by.y = "mut_info")


counts_df_complete <- counts_df_complete[order(counts_df_complete$Freq, decreasing = T),]
rownames(counts_df_complete) <- 1:nrow(counts_df_complete)


datatable(counts_df_complete)


```



```{r number of maj mutations expiv end passage}


# get numbers for in-text reporting

a <- nrow(variants_expiv[variants_expiv$passage == end_passage,])
b <- length(unique(variants_expiv$mut_info[variants_expiv$passage == end_passage]))



```


> We observe a total of `r a` `r category` mutations across the four evolution lines of `r exp` (`r b` unique mutations) at passage `r end_passage`.








### Evolution of single mutations


```{r plot af evolution for certain mutations in EXPIV, results = T}



variants_ann_expiv <- variants_ann[variants_ann$exp_line %in% 17:20,]
variants_ann_expiv <- variants_ann_expiv %>%
                        mutate(exp_line = case_when(exp_line==17 ~ "MT2_1", exp_line==18 ~ "MT2_2", exp_line==19 ~ "MT2_3", exp_line==20 ~ "MT2_4"))

variants_expiv <- variants_ann_expiv %>% filter(!duplicated(variant_id))


plot_mut_af(variants_expiv, c("7700_G_A", "8910_G_A", "9412_G_A"))



```









## MT2 of EXPIII and EXPIV

```{r throughout expiii/iv data (venn diagram), results = T}
exp <- "EXPIII/IV"
category <- "Majority"
threshold <- 0.5
end_passage <- 510



variants_ann_all <- variants_ann %>%
                            mutate(exp_line = case_when(exp_line==13 ~ "III_MT2_1", exp_line==14 ~ "III_MT2_2", exp_line==15 ~ "III_MT4_1", exp_line==16 ~ "III_MT4_2",
                            exp_line==17 ~ "IV_MT2_1", exp_line==18 ~ "IV_MT2_2", exp_line==19 ~ "IV_MT2_3", exp_line==20 ~ "IV_MT2_4"))

variants_all <- variants_ann_all %>% filter(allele_freq >= threshold & str_detect(exp_line, pattern = "MT2"))



# uniqueness of majority mutations observed throughout the experiment
throughout_maj <- variants_all %>%
                        select(c(exp_line,mut_info)) %>%
                        distinct()


#table(as.numeric(table(par_whole$mut_info)))

# create a named vector from the df
muts <- throughout_maj$mut_info
names(muts) <- throughout_maj$exp_line

# convert named vector to list
muts_list <- lapply(split(muts, names(muts)), unname)
# visualization
ggVennDiagram(muts_list) +
        labs(title = paste0(category, " Mutations throughout the ", exp))+
        theme(plot.title = element_text(size = 20, hjust = 0.5),
        plot.margin = unit(c(1,1,1,1), "cm"))

```

```{r throughout expiii/iv data (table), results=T}

counts_sep_df <- as.data.frame(table(muts, names(muts)))
counts_sep_df %<>% spread(Var2, Freq)
counts_sum_df <- as.data.frame(table(muts))


counts_df <- merge(counts_sep_df, counts_sum_df, by = "muts")


counts_df_complete <- merge(counts_df, variants_expiv_subset, by.x = "muts", by.y = "mut_info")


counts_df_complete <- counts_df_complete[order(counts_df_complete$Freq, decreasing = T),]
rownames(counts_df_complete) <- 1:nrow(counts_df_complete)


datatable(counts_df_complete)


```


```{r end passage expiii/iv data (venn diagram), results = T}

# uniqueness of majority mutations present at the end passage
end_maj <- variants_all %>%
                        filter(passage == end_passage) %>%
                        select(c(exp_line,mut_info)) %>%
                        distinct()



#table(as.numeric(table(par$mut_info)))


# create a named vector from the df
muts <- end_maj$mut_info
names(muts) <- end_maj$exp_line

# convert named vector to list
muts_list <- lapply(split(muts, names(muts)), unname)


# visualization
ggVennDiagram(muts_list) +
        labs(title = paste0(category, " Mutations at passage #", end_passage))+
        theme(plot.title = element_text(size = 20, hjust = 0.5),
        plot.margin = unit(c(1,1,1,1), "cm"))



```


```{r end passage expiii/iv data (table), results=T}

counts_sep_df <- as.data.frame(table(muts, names(muts)))
counts_sep_df %<>% spread(Var2, Freq)
counts_sum_df <- as.data.frame(table(muts))


counts_df <- merge(counts_sep_df, counts_sum_df, by = "muts")


counts_df_complete <- merge(counts_df, variants_expiv_subset, by.x = "muts", by.y = "mut_info")


counts_df_complete <- counts_df_complete[order(counts_df_complete$Freq, decreasing = T),]
rownames(counts_df_complete) <- 1:nrow(counts_df_complete)


datatable(counts_df_complete)


```



## Summary heatmap





```{r throughout summary heatmap, results = T}
exp <- "EXPIII/IV"
category <- "Majority"
threshold <- 0.5
end_passage <- 510



variants_ann_all <- variants_ann %>%
                            mutate(exp_line = case_when(exp_line==13 ~ "III_MT2_1", exp_line==14 ~ "III_MT2_2", exp_line==15 ~ "III_MT4_1", exp_line==16 ~ "III_MT4_2",
                            exp_line==17 ~ "IV_MT2_1", exp_line==18 ~ "IV_MT2_2", exp_line==19 ~ "IV_MT2_3", exp_line==20 ~ "IV_MT2_4"))

variants_all <- variants_ann_all %>% filter(allele_freq >= threshold)


# uniqueness of majority mutations observed throughout the experiment
end_maj <- variants_all %>%
                        select(c(exp_line,mut_info)) %>%
                        distinct()




# create a named vector from the df
muts <- end_maj$mut_info
names(muts) <- end_maj$exp_line

# convert named vector to list
muts_list <- lapply(split(muts, names(muts)), unname)


shared_percent <- c()

for (i in 1:8){
        dd <- muts_list[[i]]
        for (j in 1:8){
                shared_percent <- append(shared_percent,100*(sum(dd %in% muts_list[[j]])/(length(muts_list[[i]])+length(muts_list[[j]])-sum(dd %in% muts_list[[j]]))))
        }
        
}


shared_mat <- matrix(shared_percent, ncol = 8)
shared_df <- as.data.frame(shared_mat) %>% 
                        `colnames<-`(names(muts_list)) %>%
                        mutate(line1 = names(muts_list)) %>%
                        gather(key = "line2", value = "val", 1:8) %>%
                        mutate(val = round(val, 2))




plt <- ggplot(data = as.data.frame(shared_df), aes(x=line1, y=line2, fill=val)) + 
  geom_tile() +
  geom_text(aes(line1, line2, label = val), color = "black", size = 4) +
  labs(title = paste0("Majority Mutation Parallelism until Passage #", end_passage)) +
  scale_fill_gradient2(low = "white", high = "red", 
  midpoint = 1, limit = c(0,100), space = "Lab", 
  name="Shared %") +
  theme(
        axis.title = element_blank(),
        plot.title = element_text(size = 20, hjust = 0.5)
  )

ggplotly(plt)

```



```{r end passage summary heatmap, results = T}
exp <- "EXPIII/IV"
category <- "Majority"
threshold <- 0.5
end_passage <- 510



variants_ann_all <- variants_ann %>%
                            mutate(exp_line = case_when(exp_line==13 ~ "III_MT2_1", exp_line==14 ~ "III_MT2_2", exp_line==15 ~ "III_MT4_1", exp_line==16 ~ "III_MT4_2",
                            exp_line==17 ~ "IV_MT2_1", exp_line==18 ~ "IV_MT2_2", exp_line==19 ~ "IV_MT2_3", exp_line==20 ~ "IV_MT2_4"))

variants_all <- variants_ann_all %>% filter(allele_freq >= threshold)


# uniqueness of majority mutations observed throughout the experiment
end_maj <- variants_all %>%
                        filter(passage == end_passage) %>%
                        select(c(exp_line,mut_info)) %>%
                        distinct()




# create a named vector from the df
muts <- end_maj$mut_info
names(muts) <- end_maj$exp_line

# convert named vector to list
muts_list <- lapply(split(muts, names(muts)), unname)


shared_percent <- c()


for (i in 1:8){
        dd <- muts_list[[i]]
        for (j in 1:8){
                shared_percent <- append(shared_percent,100*(sum(dd %in% muts_list[[j]])/(length(muts_list[[i]])+length(muts_list[[j]])-sum(dd %in% muts_list[[j]]))))
        }
        
}


shared_mat <- matrix(shared_percent, ncol = 8)
shared_df <- as.data.frame(shared_mat) %>% 
                        `colnames<-`(names(muts_list)) %>%
                        mutate(line1 = names(muts_list)) %>%
                        gather(key = "line2", value = "val", 1:8) %>%
                        mutate(val = round(val, 2))




plt <- ggplot(data = as.data.frame(shared_df), aes(x=line1, y=line2, fill=val)) + 
  geom_tile() +
  geom_text(aes(line1, line2, label = val), color = "black", size = 4) +
  labs(title = paste0("Majority Mutation Parallelism at Passage #", end_passage)) +
  scale_fill_gradient2(low = "white", high = "red", 
  midpoint = 1, limit = c(0,100), space = "Lab", 
  name="Shared %") +
  theme(
        axis.title = element_blank(),
        plot.title = element_text(size = 20, hjust = 0.5)
  )

ggplotly(plt)

```