---
title:  
    "Overview of HIV-1 LTEE genomic data (Fredrick's version)"
date: 
    "Sep 6th, 2023"
author:  
    "Ali Movasati"  
output:  
    html_document:
        toc: true
        toc_depth: 3
        toc_float: true
        number_sections: true
        theme: lumen
        highlight: zenburn
        code_folding: hide
---



```{r setting global chunk parameters, include=F}

# options for code chunk and output inclusion 
knitr::opts_chunk$set(
  message = FALSE, echo = FALSE, warning=FALSE, results = "hide"
)

# options for figures
knitr::opts_chunk$set(
  fig.width = 10, fig.height = 10, fig.align = "center"
)


```



# Introduction

In this document I have re-created some of the main analyses of Fredrick's paper that is relevant to our study.  
> Here's the link to the paper: https://academic.oup.com/mbe/article/36/11/2400/5524887?login=false

Our dataset includes genome sequencing of two **LTEE** experiments, namely *EXPIII* & *EXPIV*, including passages up to 570 and 510, respectively.

___
___


# Overview of the Dataset


```{r loading libraries}
# loading libraries
library(plotly)
library(DT)


```


```{r read-in the data}

# read in the data


variants_ann <- read.table(file = "/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/tables/3-p/all_annotated_variants.tsv", 
sep = "\t", stringsAsFactors = FALSE, header = TRUE)

variants_ann$exp_line <- as.integer(variants_ann$exp_line)


# set host cell line info
variants_ann$host_line <- ifelse(variants_ann$exp_line %in% c("13", "14","17", "18", "19", "20"), "MT2", "MT4")

# account for the switched experimental lines

variants_ann$exp_line[variants_ann$exp_line == 15 & variants_ann$passage > 360] <- "A"
variants_ann$exp_line[variants_ann$exp_line == 16 & variants_ann$passage > 360] <- "B"

variants_ann$exp_line[variants_ann$exp_line == "A"] <- "16"
variants_ann$exp_line[variants_ann$exp_line == "B"] <- "15"

variants_ann$mut_info <- paste(variants_ann$genomic_pos, variants_ann$ref_allele, variants_ann$alt_allele, sep = "_")

head(variants_ann)
nrow(variants_ann)
summary(variants_ann)

```

## EXPIII


```{r}
# number of muts above allele frequency of 0.5

a <- nrow(variants_ann[variants_ann$exp_line %in% 13:16 & variants_ann$allele_freq >= 0.5 & variants_ann$passage == 570,])
b <- length(unique(variants_ann$mut_info[variants_ann$exp_line %in% 13:16 & variants_ann$allele_freq >= 0.5 & variants_ann$passage == 570]))

```

We observed a total of `r a` mutations across the four evolution lines (`r b` unique mutations) 
of EXPIII that became more frequent in the population than the wild-type nucleotide (a.k.a, allele frequency above 0.5, majority mutations)!

we do not observe deceleration of mutation accumulation during the 570 passages plotted below.

```{r, results = TRUE}
# plot accumulation of majority muts in EXPIII

variants_ann_expiii <- variants_ann[variants_ann$exp_line %in% 13:16 & variants_ann$allele_freq >= 0.5,]


i <- 0
df_processed <- data.frame()

for (line in 13:16){
  for (passage in seq(10, 570, by = 10)){
    i <- i +1
    nr_subset <- nrow(variants_ann_expiii[variants_ann_expiii$exp_line == line & variants_ann_expiii$passage == passage,])

    df_processed <- rbind(df_processed, c(line, passage, nr_subset))
  }
}

colnames(df_processed) <- c("exp_line", "passage", "nr_majority")
df_processed$host_line <- ifelse(df_processed$exp_line %in% c(13:14,17:20), "MT2", "MT4")


df_processed[,2:3] <- apply(df_processed[,2:3], 2, as.numeric)
df_processed$exp_line <- factor(df_processed$exp_line)
df_processed$host_line <- factor(df_processed$host_line)


plot_ly(df_processed, x = ~passage, y = ~nr_majority, 
        #text = ~paste("Car: ", car),
        type = 'scatter', mode = "lines", color = ~exp_line, linetype= ~host_line)


```


If we plot the accumulation of majority mutations based on the effect type we can see
the trends for S (synonymous), N (non-synonomous), and U (untranslated) mutations.

```{r, results = TRUE}

variants_ann_expiii$effect_simplified[variants_ann_expiii$effect == "synonymous_variant"] <- "S"
variants_ann_expiii$effect_simplified[variants_ann_expiii$effect == "untranslated_region"] <- "U"
variants_ann_expiii$effect_simplified[is.na(variants_ann_expiii$effect_simplified)] <- "N"


i <- 0
df_processed <- data.frame()

for (line in 13:16){
  for (effect_simplified in c("S", "N", "U")){
    for (passage in seq(10, 570, by = 10)){
      i <- i +1
      nr_subset <- nrow(variants_ann_expiii[variants_ann_expiii$exp_line == line & variants_ann_expiii$passage == passage & variants_ann_expiii$effect_simplified == effect_simplified,])

      df_processed <- rbind(df_processed, c(line, effect_simplified, passage, nr_subset))
    } 
  }
}

colnames(df_processed) <- c("exp_line", "effect_type", "passage", "nr_majority")


df_processed[,3:4] <- apply(df_processed[,3:4], 2, as.numeric)
df_processed$effect_type <- factor(df_processed$effect_type)
df_processed$exp_line <- factor(df_processed$exp_line)

plot_ly(df_processed, x = ~passage, y = ~nr_majority, 
        #text = ~paste("Car: ", car),
        type = 'scatter', mode = "lines", color = ~effect_type, linetype= ~exp_line)





```


We can also plot the same accumulation for different effect types but corrected for the 
genomic length that each mutation occur. As you can see eventhough the absolute number of 
synanymous and non-synanymous mutations are higher than mutations in untranslated regions,
the density of mutations in untranslated regions is initially higher than protein-coding regions,
while the difference decreases towards the end of evolution.


```{r, results = TRUE}


i <- 0
df_processed <- data.frame()

for (line in 13:16){
  for (effect_simplified in c("S/N", "U")){
    for (passage in seq(10, 570, by = 10)){
      i <- i +1
      if (effect_simplified == "U"){
        nr_subset <- nrow(variants_ann_expiii[variants_ann_expiii$exp_line == line & variants_ann_expiii$passage == passage & variants_ann_expiii$effect_simplified == effect_simplified,])
      } else{
        nr_subset <- nrow(variants_ann_expiii[variants_ann_expiii$exp_line == line & variants_ann_expiii$passage == passage & variants_ann_expiii$effect_simplified != "U",])
      }
      

      df_processed <- rbind(df_processed, c(line, effect_simplified, passage, nr_subset))
    } 
  }
}

colnames(df_processed) <- c("exp_line", "effect_type", "passage", "nr_majority")


df_processed[,3:4] <- apply(df_processed[,3:4], 2, as.numeric)
df_processed$effect_type <- factor(df_processed$effect_type)
df_processed$exp_line <- factor(df_processed$exp_line)



df_processed$nr_majority[df_processed$effect_type == "S/N"] <- 
df_processed$nr_majority[df_processed$effect_type == "S/N"]*100/(9407-790)


df_processed$nr_majority[df_processed$effect_type == "U"] <- 
df_processed$nr_majority[df_processed$effect_type == "U"]*100/(9709-9407+790)

plot_ly(df_processed, x = ~passage, y = ~nr_majority, 
        #text = ~paste("Car: ", car),
        type = 'scatter', mode = "lines", color = ~effect_type, linetype= ~exp_line,
        colors = "Set1")

```



We can also look at number of mutations in different genes and regions of the genome:



```{r}

table(variants_ann_expiii$effect[variants_ann_expiii$exp_line == 15 & variants_ann_expiii$passage == 570], useNA = "always")

table(variants_ann_expiii$feature[variants_ann_expiii$exp_line == 15 & variants_ann_expiii$passage == 570],
variants_ann_expiii$effect[variants_ann_expiii$exp_line == 15 & variants_ann_expiii$passage == 570]
, useNA = "always")

```
