---
title:  
    "Mutation Category Thresholds"
date: 
    "Oct 5th, 2023"
author:  
    "Ali Movasati"  
output:  
    html_document:
        #css: style.css
        toc: true
        toc_depth: 3
        toc_float: true
        number_sections: true
        theme: lumen
        highlight: zenburn
        code_folding: hide
---







```{r setting global chunk parameters, include=FALSE}

# options for code chunk and output inclusion 
knitr::opts_chunk$set(
  message = FALSE, echo = TRUE, warning=FALSE, results = "hide"
)

# options for figures
knitr::opts_chunk$set(
  fig.width = 10, fig.height = 12, fig.align = "center"
)

# options(scipen=999)

# margin of plotly
m <- list(
  l = 150,
  r = 150,
  b = 100,
  t = 100,
  pad = 4
)


```




```{r loading libraries, echo=FALSE}

# loading libraries
library("ggplot2")
library("reshape2")
library("plotly")
library("magrittr")
library("dplyr")

```




```{r read-in the data}

# read in the data
variants_ann <- read.table(file = "/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/tables/2-p/all_annotated_variants.tsv.gz", 
sep = "\t", stringsAsFactors = FALSE, header = TRUE)

# set correct data types
variants_ann$exp_line <- as.integer(variants_ann$exp_line)


# set host cell line info
variants_ann$host_line <- ifelse(variants_ann$exp_line %in% c("13", "14","17", "18", "19", "20"), "MT2", "MT4")


# account for the switched experimental lines
variants_ann$exp_line[variants_ann$exp_line == 15 & variants_ann$passage > 360] <- "A"
variants_ann$exp_line[variants_ann$exp_line == 16 & variants_ann$passage > 360] <- "B"

variants_ann$exp_line[variants_ann$exp_line == "A"] <- "16"
variants_ann$exp_line[variants_ann$exp_line == "B"] <- "15"



variants_ann$exp_line[variants_ann$exp_line == 18 & variants_ann$passage > 350] <- "A"
variants_ann$exp_line[variants_ann$exp_line == 19 & variants_ann$passage > 350] <- "B"

variants_ann$exp_line[variants_ann$exp_line == "A"] <- "19"
variants_ann$exp_line[variants_ann$exp_line == "B"] <- "18"

# add a field containing pos, ref, and alt info of muts
variants_ann$mut_info <- paste(variants_ann$genomic_pos, variants_ann$ref_allele, variants_ann$alt_allele, sep = "_")


## label experimental line numbers


variants_ann <- variants_ann %>%
                            mutate(exp = case_when(exp_line%in%13:16 ~ "EXPIII", exp_line%in%17:20 ~ "EXPIV")) %>%
                            mutate(exp_line = case_when(exp_line==13 ~ "III_MT2_1", exp_line==14 ~ "III_MT2_2", exp_line==15 ~ "III_MT4_1", exp_line==16 ~ "III_MT4_2",
                            exp_line==17 ~ "IV_MT2_1", exp_line==18 ~ "IV_MT2_2", exp_line==19 ~ "IV_MT2_3", exp_line==20 ~ "IV_MT2_4"))

variants <- variants_ann %>% filter(!duplicated(variant_id))


```




# Introduction

In this document I will use different threshold cut-offs for defining mutation categories. The thresholds used by Eva in her work are as follow:

> Minority mutations: AF >= 0.05 & AF < 0.5  
Majority mutations: AF >= 0.5 & AF < 0.99  
Fixated mutations: AF >= 0.99


I will plot the mutation accumulation with different cut-off sets and provide some basic statistics for EXPIII and EXPIV data separately.


## Allele Frequency distribution

Let's first take a look at the allele frequency distribution of all mutations recorded


```{r, results=TRUE}


gg_fig <- ggplot(variants, aes(x=allele_freq)) + 
    geom_histogram(aes(y=..count..),      # Histogram with density instead of count on y-axis
                   binwidth=.01,
                   colour="black", fill="white") +
    facet_wrap(~exp) +
    geom_density(aes(y = ..count.. * .01), alpha=.3, fill="#FF6666") +
    labs(title = "Allele Freq. Distribution", x = "Allele Freq.", y = "Count") +
    theme_bw() +
    theme(
    plot.title = element_text(size = 30, hjust = 0.5),
    axis.title = element_text(size = 20),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 10),
    strip.text.x = element_text(size = 25),
    legend.title = element_text(size=15),
    legend.text = element_text(size = 10),
    plot.margin = unit(c(1,1,1,1), "cm"))


ply_fig <- ggplotly(gg_fig)

ply_fig




gg_fig <- ggplot(variants[variants$allele_freq >= 0.01,], aes(x=allele_freq)) + 
    geom_histogram(aes(y=..count..),      # Histogram with density instead of count on y-axis
                   binwidth=.01,
                   colour="black", fill="white") +
    facet_wrap(~exp) +
    geom_density(aes(y = ..count.. * .01), alpha=.3, fill="#FF6666") +
    labs(title = "Allele Freq. Distribution (AF>=0.01)", x = "Allele Freq.", y = "Count") +
    theme_bw() +
    theme(
    plot.title = element_text(size = 30, hjust = 0.5),
    axis.title = element_text(size = 20),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 10),
    strip.text.x = element_text(size = 25),
    legend.title = element_text(size=15),
    legend.text = element_text(size = 10),
    plot.margin = unit(c(1,1,1,1), "cm"))


ply_fig <- ggplotly(gg_fig)

ply_fig


```



The distribution of AF seems to be following a bi-modal distribution in both experiments.



## Minority Mutations

### Definition A

```{r}

category <- "Minority"
min_threshold <- 0.0001
max_threshold <- 0.50


```

> `r category` mutations: AF >= `r min_threshold` & AF < `r max_threshold`  


```{r, results=TRUE}
# prepare data for visualization
counts <- variants %>%
              filter(allele_freq >= min_threshold & allele_freq < max_threshold) %>%
              group_by(exp, exp_line, passage) %>%
              count() %>%
              dplyr::rename(mut_count = n) %>%
              mutate_at(c(1,2), as.factor) %>%
              mutate_at(3:4, as.numeric)



# visualize using plotly
gg_fig <- counts %>% ggplot(aes(x = passage, y = mut_count, color = exp_line)) +
                      geom_line() +
                      facet_wrap(~exp, scale = "free", ncol = 1) +
                      labs(title = paste0(category, " Mutation Accumulation"), x = "Passage", y = "Counts") +
                      theme_bw() +
                      theme(
                      plot.title = element_text(size = 30, hjust = 0.5),
                      axis.title = element_text(size = 20),
                      axis.text.x = element_text(size = 10),
                      axis.text.y = element_text(size = 10),
                      strip.text.x = element_text(size = 25),
                      legend.title = element_text(size=15),
                      legend.text = element_text(size = 10),
                      plot.margin = unit(c(1,1,1,1), "cm"))


ply_fig <- ggplotly(gg_fig)

ply_fig
```



### Definition A*


```{r, results=T}

category <- "Minority"
min_threshold_tests <- c(0.001,0.0001,0.00001,0.000001)
max_threshold <- 0.50


counts <- data.frame()
for (min_threshold_test in min_threshold_tests){
  count <- variants %>%
              filter(allele_freq >= min_threshold_test & allele_freq < max_threshold) %>%
              group_by(exp, exp_line, passage) %>%
              count() %>%
              dplyr::rename(mut_count = n) %>%
              mutate_at(c(1,2), as.factor) %>%
              mutate_at(3:4, as.numeric) %>%
              mutate(min_threshold = min_threshold_test)

  counts <- rbind(counts, count)
}
head(counts)
counts_iii <- counts[counts$exp == "EXPIII",]
counts_iv <- counts[counts$exp == "EXPIV",]

# visualize using plotly
gg_fig <- counts_iii %>% ggplot(aes(x = passage, y = mut_count, color = exp_line)) +
                      geom_line() +
                      facet_wrap(~min_threshold, scale = "free", ncol = 1) +
                      labs(title = paste0(category, " Mutation Accumulation of EXPIII"), x = "Passage", y = "Counts") +
                      theme_bw() +
                      theme(
                      plot.title = element_text(size = 30, hjust = 0.5),
                      axis.title = element_text(size = 20),
                      axis.text.x = element_text(size = 10),
                      axis.text.y = element_text(size = 10),
                      strip.text.x = element_text(size = 25),
                      legend.title = element_text(size=15),
                      legend.text = element_text(size = 10),
                      plot.margin = unit(c(1,1,1,1), "cm"))


ply_fig <- ggplotly(gg_fig)

ply_fig


# visualize using plotly
gg_fig <- counts_iv %>% ggplot(aes(x = passage, y = mut_count, color = exp_line)) +
                      geom_line() +
                      facet_wrap(~min_threshold, scale = "free", ncol = 1) +
                      labs(title = paste0(category, " Mutation Accumulation of EXPIV"), x = "Passage", y = "Counts") +
                      theme_bw() +
                      theme(
                      plot.title = element_text(size = 30, hjust = 0.5),
                      axis.title = element_text(size = 20),
                      axis.text.x = element_text(size = 10),
                      axis.text.y = element_text(size = 10),
                      strip.text.x = element_text(size = 25),
                      legend.title = element_text(size=15),
                      legend.text = element_text(size = 10),
                      plot.margin = unit(c(1,1,1,1), "cm"))


ply_fig <- ggplotly(gg_fig)

ply_fig


```
### Definition B

```{r}

category <- "Minority"
min_threshold <- 0.025
max_threshold <- 0.50


```

> `r category` mutations: AF >= `r min_threshold` & AF < `r max_threshold`  


```{r, results=TRUE}
# prepare data for visualization
counts <- variants %>%
              filter(allele_freq >= min_threshold & allele_freq < max_threshold) %>%
              group_by(exp, exp_line, passage) %>%
              count() %>%
              dplyr::rename(mut_count = n) %>%
              mutate_at(c(1,2), as.factor) %>%
              mutate_at(3:4, as.numeric)



# visualize using plotly
gg_fig <- counts %>% ggplot(aes(x = passage, y = mut_count, color = exp_line)) +
                      geom_line() +
                      facet_wrap(~exp, scale = "free", ncol = 1) +
                      labs(title = paste0(category, " Mutation Accumulation"), x = "Passage", y = "Counts") +
                      theme_bw() +
                      theme(
                      plot.title = element_text(size = 30, hjust = 0.5),
                      axis.title = element_text(size = 20),
                      axis.text.x = element_text(size = 10),
                      axis.text.y = element_text(size = 10),
                      strip.text.x = element_text(size = 25),
                      legend.title = element_text(size=15),
                      legend.text = element_text(size = 10),
                      plot.margin = unit(c(1,1,1,1), "cm"))


ply_fig <- ggplotly(gg_fig)

ply_fig
```



### Definition C

```{r}

category <- "Minority"
min_threshold <- 0.05
max_threshold <- 0.50


```

> `r category` mutations: AF >= `r min_threshold` & AF < `r max_threshold`  


```{r, results=TRUE}
# prepare data for visualization
counts <- variants %>%
              filter(allele_freq >= min_threshold & allele_freq < max_threshold) %>%
              group_by(exp, exp_line, passage) %>%
              count() %>%
              dplyr::rename(mut_count = n) %>%
              mutate_at(c(1,2), as.factor) %>%
              mutate_at(3:4, as.numeric)



# visualize using plotly
gg_fig <- counts %>% ggplot(aes(x = passage, y = mut_count, color = exp_line)) +
                      geom_line() +
                      facet_wrap(~exp, scale = "free", ncol = 1) +
                      labs(title = paste0(category, " Mutation Accumulation"), x = "Passage", y = "Counts") +
                      theme_bw() +
                      theme(
                      plot.title = element_text(size = 30, hjust = 0.5),
                      axis.title = element_text(size = 20),
                      axis.text.x = element_text(size = 10),
                      axis.text.y = element_text(size = 10),
                      strip.text.x = element_text(size = 25),
                      legend.title = element_text(size=15),
                      legend.text = element_text(size = 10),
                      plot.margin = unit(c(1,1,1,1), "cm"))


ply_fig <- ggplotly(gg_fig)

ply_fig
```


### Summary of Definitions A,B,C



```{r, results=TRUE}


# prepare data for visualization

all_counts <- data.frame()


defs <- c("A", "B", "C")
min_thresholds <- c(0.01, 0.025, 0.05)


for (i in 1:3){
  category <- "Minority"
  def <- defs[i]
  min_threshold <- min_thresholds[i]
  max_threshold <- 0.50

  counts <- variants %>%
            filter(allele_freq >= min_threshold & allele_freq < max_threshold) %>%
            group_by(exp, exp_line, passage) %>%
            count() %>%
            mutate(definition = def) %>%
            dplyr::rename(mut_count = n) %>%
            mutate_at(c(1,2,5), as.factor) %>%
            mutate_at(3:4, as.numeric)

  all_counts <- rbind(all_counts, counts)

}

# visualize using plotly
gg_fig <- all_counts %>% ggplot(aes(x = passage, y = mut_count, color = exp_line, linetype = definition)) +
                      geom_line() +
                      facet_wrap(~exp, scale = "free", ncol = 1) +
                      labs(title = paste0(category, " Mutation Accumulation (Summary)"), x = "Passage", y = "Counts") +
                      theme_bw() +
                      theme(
                      plot.title = element_text(size = 30, hjust = 0.5),
                      axis.title = element_text(size = 20),
                      axis.text.x = element_text(size = 10),
                      axis.text.y = element_text(size = 10),
                      strip.text.x = element_text(size = 25),
                      legend.title = element_text(size=15),
                      legend.text = element_text(size = 10),
                      plot.margin = unit(c(1,1,1,1), "cm"))


ply_fig <- ggplotly(gg_fig)

ply_fig
```


#### Summary Statistics

Below you can see summary statistics of range of mutation count difference between different definitions for each experimental line. 
```{r, results = TRUE}


summary_stats <- all_counts %>%
  group_by(exp_line, passage) %>%
  summarise(
          count = n(),
          mean_count = mean(mut_count),
          max_count = max(mut_count),
          min_count = min(mut_count),
          range_count = max_count - min_count
            )

tapply(summary_stats$range_count, summary_stats$exp_line, summary)



```

Standard deviation of the same metric: 

```{r, results = TRUE}

tapply(summary_stats$range_count, summary_stats$exp_line, sd)

```





## Majority Mutations



### Definition A

```{r}

category <- "Majority"
min_threshold <- 0.5
max_threshold <- 0.9


```

> `r category` mutations: AF >= `r min_threshold` & AF < `r max_threshold`  


```{r, results=TRUE}
# prepare data for visualization
counts <- variants %>%
              filter(allele_freq >= min_threshold & allele_freq < max_threshold) %>%
              group_by(exp, exp_line, passage) %>%
              count() %>%
              dplyr::rename(mut_count = n) %>%
              mutate_at(c(1,2), as.factor) %>%
              mutate_at(3:4, as.numeric)



# visualize using plotly
gg_fig <- counts %>% ggplot(aes(x = passage, y = mut_count, color = exp_line)) +
                      geom_line() +
                      facet_wrap(~exp, scale = "free", ncol = 1) +
                      labs(title = paste0(category, " Mutation Accumulation"), x = "Passage", y = "Counts") +
                      theme_bw() +
                      theme(
                      plot.title = element_text(size = 30, hjust = 0.5),
                      axis.title = element_text(size = 20),
                      axis.text.x = element_text(size = 10),
                      axis.text.y = element_text(size = 10),
                      strip.text.x = element_text(size = 25),
                      legend.title = element_text(size=15),
                      legend.text = element_text(size = 10),
                      plot.margin = unit(c(1,1,1,1), "cm"))


ply_fig <- ggplotly(gg_fig)

ply_fig
```


### Definition B

```{r}

category <- "Majority"
min_threshold <- 0.5
max_threshold <- 0.95


```

> `r category` mutations: AF >= `r min_threshold` & AF < `r max_threshold`  


```{r, results=TRUE}
# prepare data for visualization
counts <- variants %>%
              filter(allele_freq >= min_threshold & allele_freq < max_threshold) %>%
              group_by(exp, exp_line, passage) %>%
              count() %>%
              dplyr::rename(mut_count = n) %>%
              mutate_at(c(1,2), as.factor) %>%
              mutate_at(3:4, as.numeric)



# visualize using plotly
gg_fig <- counts %>% ggplot(aes(x = passage, y = mut_count, color = exp_line)) +
                      geom_line() +
                      facet_wrap(~exp, scale = "free", ncol = 1) +
                      labs(title = paste0(category, " Mutation Accumulation"), x = "Passage", y = "Counts") +
                      theme_bw() +
                      theme(
                      plot.title = element_text(size = 30, hjust = 0.5),
                      axis.title = element_text(size = 20),
                      axis.text.x = element_text(size = 10),
                      axis.text.y = element_text(size = 10),
                      strip.text.x = element_text(size = 25),
                      legend.title = element_text(size=15),
                      legend.text = element_text(size = 10),
                      plot.margin = unit(c(1,1,1,1), "cm"))


ply_fig <- ggplotly(gg_fig)

ply_fig
```


### Definition C

```{r}

category <- "Majority"
min_threshold <- 0.5
max_threshold <- 0.99


```

> `r category` mutations: AF >= `r min_threshold` & AF < `r max_threshold`  


```{r, results=TRUE}
# prepare data for visualization
counts <- variants %>%
              filter(allele_freq >= min_threshold & allele_freq < max_threshold) %>%
              group_by(exp, exp_line, passage) %>%
              count() %>%
              dplyr::rename(mut_count = n) %>%
              mutate_at(c(1,2), as.factor) %>%
              mutate_at(3:4, as.numeric)



# visualize using plotly
gg_fig <- counts %>% ggplot(aes(x = passage, y = mut_count, color = exp_line)) +
                      geom_line() +
                      facet_wrap(~exp, scale = "free", ncol = 1) +
                      labs(title = paste0(category, " Mutation Accumulation"), x = "Passage", y = "Counts") +
                      theme_bw() +
                      theme(
                      plot.title = element_text(size = 30, hjust = 0.5),
                      axis.title = element_text(size = 20),
                      axis.text.x = element_text(size = 10),
                      axis.text.y = element_text(size = 10),
                      strip.text.x = element_text(size = 25),
                      legend.title = element_text(size=15),
                      legend.text = element_text(size = 10),
                      plot.margin = unit(c(1,1,1,1), "cm"))


ply_fig <- ggplotly(gg_fig)

ply_fig
```


### Summary of Definitions A,B,C



```{r, results=TRUE}


# prepare data for visualization

all_counts <- data.frame()


defs <- c("A", "B", "C")
max_thresholds <- c(0.90, 0.95, 0.99)


for (i in 1:3){
  category <- "Majority"
  def <- defs[i]
  min_threshold <- 0.50
  max_threshold <- max_thresholds[i]
  

  counts <- variants %>%
            filter(allele_freq >= min_threshold & allele_freq < max_threshold) %>%
            group_by(exp, exp_line, passage) %>%
            count() %>%
            mutate(definition = def) %>%
            dplyr::rename(mut_count = n) %>%
            mutate_at(c(1,2,5), as.factor) %>%
            mutate_at(3:4, as.numeric)

  all_counts <- rbind(all_counts, counts)

}

# visualize using plotly
gg_fig <- all_counts %>% ggplot(aes(x = passage, y = mut_count, color = exp_line, linetype = definition)) +
                      geom_line() +
                      facet_wrap(~exp, scale = "free", ncol = 1) +
                      labs(title = paste0(category, " Mutation Accumulation (Summary)"), x = "Passage", y = "Counts") +
                      theme_bw() +
                      theme(
                      plot.title = element_text(size = 30, hjust = 0.5),
                      axis.title = element_text(size = 20),
                      axis.text.x = element_text(size = 10),
                      axis.text.y = element_text(size = 10),
                      strip.text.x = element_text(size = 25),
                      legend.title = element_text(size=15),
                      legend.text = element_text(size = 10),
                      plot.margin = unit(c(1,1,1,1), "cm"))


ply_fig <- ggplotly(gg_fig)

ply_fig
```


#### Summary Statistics

Below you can see summary statistics of range of mutation count difference between different definitions for each experimental line. 
```{r, results = TRUE}


summary_stats <- all_counts %>%
  group_by(exp_line, passage) %>%
  summarise(
          count = n(),
          mean_count = mean(mut_count),
          max_count = max(mut_count),
          min_count = min(mut_count),
          range_count = max_count - min_count
            )

tapply(summary_stats$range_count, summary_stats$exp_line, summary)


```

Standard deviation of the same metric: 

```{r, results = TRUE}

tapply(summary_stats$range_count, summary_stats$exp_line, sd)

```




## Fixated mutations


### Definition A

```{r}

category <- "Fixated"
min_threshold <- 0.9



```

> `r category` mutations: AF >= `r min_threshold`  


```{r, results=TRUE}
# prepare data for visualization
counts <- variants %>%
              filter(allele_freq >= min_threshold) %>%
              group_by(exp, exp_line, passage) %>%
              count() %>%
              dplyr::rename(mut_count = n) %>%
              mutate_at(c(1,2), as.factor) %>%
              mutate_at(3:4, as.numeric)



# visualize using plotly
gg_fig <- counts %>% ggplot(aes(x = passage, y = mut_count, color = exp_line)) +
                      geom_line() +
                      facet_wrap(~exp, scale = "free", ncol = 1) +
                      labs(title = paste0(category, " Mutation Accumulation"), x = "Passage", y = "Counts") +
                      theme_bw() +
                      theme(
                      plot.title = element_text(size = 30, hjust = 0.5),
                      axis.title = element_text(size = 20),
                      axis.text.x = element_text(size = 10),
                      axis.text.y = element_text(size = 10),
                      strip.text.x = element_text(size = 25),
                      legend.title = element_text(size=15),
                      legend.text = element_text(size = 10),
                      plot.margin = unit(c(1,1,1,1), "cm"))


ply_fig <- ggplotly(gg_fig)

ply_fig
```


### Definition B

```{r}

category <- "Fixated"
min_threshold <- 0.95



```

> `r category` mutations: AF >= `r min_threshold`  


```{r, results=TRUE}
# prepare data for visualization
counts <- variants %>%
              filter(allele_freq >= min_threshold) %>%
              group_by(exp, exp_line, passage) %>%
              count() %>%
              dplyr::rename(mut_count = n) %>%
              mutate_at(c(1,2), as.factor) %>%
              mutate_at(3:4, as.numeric)



# visualize using plotly
gg_fig <- counts %>% ggplot(aes(x = passage, y = mut_count, color = exp_line)) +
                      geom_line() +
                      facet_wrap(~exp, scale = "free", ncol = 1) +
                      labs(title = paste0(category, " Mutation Accumulation"), x = "Passage", y = "Counts") +
                      theme_bw() +
                      theme(
                      plot.title = element_text(size = 30, hjust = 0.5),
                      axis.title = element_text(size = 20),
                      axis.text.x = element_text(size = 10),
                      axis.text.y = element_text(size = 10),
                      strip.text.x = element_text(size = 25),
                      legend.title = element_text(size=15),
                      legend.text = element_text(size = 10),
                      plot.margin = unit(c(1,1,1,1), "cm"))


ply_fig <- ggplotly(gg_fig)

ply_fig
```


### Definition C

```{r}

category <- "Fixated"
min_threshold <- 0.99



```

> `r category` mutations: AF >= `r min_threshold`  


```{r, results=TRUE}
# prepare data for visualization
counts <- variants %>%
              filter(allele_freq >= min_threshold) %>%
              group_by(exp, exp_line, passage) %>%
              count() %>%
              dplyr::rename(mut_count = n) %>%
              mutate_at(c(1,2), as.factor) %>%
              mutate_at(3:4, as.numeric)



# visualize using plotly
gg_fig <- counts %>% ggplot(aes(x = passage, y = mut_count, color = exp_line)) +
                      geom_line() +
                      facet_wrap(~exp, scale = "free", ncol = 1) +
                      labs(title = paste0(category, " Mutation Accumulation"), x = "Passage", y = "Counts") +
                      theme_bw() +
                      theme(
                      plot.title = element_text(size = 30, hjust = 0.5),
                      axis.title = element_text(size = 20),
                      axis.text.x = element_text(size = 10),
                      axis.text.y = element_text(size = 10),
                      strip.text.x = element_text(size = 25),
                      legend.title = element_text(size=15),
                      legend.text = element_text(size = 10),
                      plot.margin = unit(c(1,1,1,1), "cm"))


ply_fig <- ggplotly(gg_fig)

ply_fig
```



### Summary of Definitions A,B,C



```{r, results=TRUE}


# prepare data for visualization

all_counts <- data.frame()


defs <- c("A", "B", "C")
min_thresholds <- c(0.90, 0.95, 0.99)


for (i in 1:3){
  category <- "Fixated"
  def <- defs[i]
  min_threshold <- min_thresholds[i]
  

  counts <- variants %>%
            filter(allele_freq >= min_threshold) %>%
            group_by(exp, exp_line, passage) %>%
            count() %>%
            mutate(definition = def) %>%
            dplyr::rename(mut_count = n) %>%
            mutate_at(c(1,2,5), as.factor) %>%
            mutate_at(3:4, as.numeric)

  all_counts <- rbind(all_counts, counts)

}

# visualize using plotly
gg_fig <- all_counts %>% ggplot(aes(x = passage, y = mut_count, color = exp_line, linetype = definition)) +
                      geom_line() +
                      facet_wrap(~exp, scale = "free", ncol = 1) +
                      labs(title = paste0(category, " Mutation Accumulation (Summary)"), x = "Passage", y = "Counts") +
                      theme_bw() +
                      theme(
                      plot.title = element_text(size = 30, hjust = 0.5),
                      axis.title = element_text(size = 20),
                      axis.text.x = element_text(size = 10),
                      axis.text.y = element_text(size = 10),
                      strip.text.x = element_text(size = 25),
                      legend.title = element_text(size=15),
                      legend.text = element_text(size = 10),
                      plot.margin = unit(c(1,1,1,1), "cm"))


ply_fig <- ggplotly(gg_fig)


ply_fig
```


#### Summary Statistics

Below you can see summary statistics of range of mutation count difference between different definitions for each experimental line. 

```{r, results = TRUE}


summary_stats <- all_counts %>%
  group_by(exp_line, passage) %>%
  summarise(
          count = n(),
          mean_count = mean(mut_count),
          max_count = max(mut_count),
          min_count = min(mut_count),
          range_count = max_count - min_count
            )

tapply(summary_stats$range_count, summary_stats$exp_line, summary)


```


Standard deviation of the same metric: 

```{r, results = TRUE}

tapply(summary_stats$range_count, summary_stats$exp_line, sd)

```



# Conclusion

Overall we see more difference between the number of mutation counts based on different threshold definitions in EXPIII compared to EXPIV.

