---
title:  
    "Parallel evolution of over time"
date: 
    "May 26th, 2025"
author:  
    "Ali Movasati"  
output:  
    html_document:
        #css: style.css
        toc: true
        toc_depth: 2
        toc_float: true
        number_sections: true
        theme: lumen
        highlight: zenburn
        code_folding: hide
        code_download: true # to make the Rmd source code available
---




```{r loading libraries, echo=FALSE}

# loading libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(lme4)
library(knitr)
library(kableExtra)
library(gridExtra)
library(grid)
library(magrittr)
library(ggh4x)


```

```{r set paths and load data}

# Set working directory
if (file.exists("/home/amovas/")) {
    # Running on HPC
    message("Remote HPC Connection!")
    wd <- "/home/amovas/data/genome-evo-proj/"
} else {
    # Running locally
    message("Local PC Connection!")
    wd <- "/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/"
}

# Load project-specific functions and data readers
source(file.path(wd, "src/analysis/decelerating-adaptation/readin_data.R"))
source(file.path(wd, "src/analysis/decelerating-adaptation/plot_custom.R"))

```

```{r category thresholds}

# set parameters
category <- "Fixed"

if (category == "Minority"){
    min_threshold <- 0.01
    max_threshold <- 0.5
} else if (category == "Majority"){
    min_threshold <- 0.5
    max_threshold <- 1.01
} else if (category == "Fixed"){
    min_threshold <- 0.99
    max_threshold <- 1.01
}
```



```{r simulation of parallel evolution over time}

effect_list <- list(Syn = "S", Non_Syn = "N", Untranslated = "U", All = c("S", "N", "U"))


# for (k in 4){
#     df <- data.frame()
#     effect <- effect_list[k]
#     print(effect)

#     # Nr of unique mutations with af >= 0.05
#     variants_expiii$passage <- as.integer(as.character(variants_expiii$passage))
#     variants_expiii$effect_simplified <- as.character(variants_expiii$effect_simplified)

#     # filter for relevant mutations
#     variants_expiii_sub <- variants_expiii[variants_expiii$effect_simplified %in% effect[[1]] & variants_expiii$mut_type == "P",]


#     for (end_psg in seq(10,500,10)){
#         print(end_psg)

#         for (line1 in exp_line_factor){
#             for (line2 in exp_line_factor){
#                 exp_all <- data.frame()
#                 for (base in c("A", "C", "G", "T")){
#                     variants_expiii_sub2 <- variants_expiii_sub[variants_expiii_sub$ref_allele == base,]
#                     a <- length(unique(variants_expiii_sub2$mut_info[variants_expiii_sub2$allele_freq >= 0.05 & variants_expiii_sub2$exp_line %in% c(line1) & variants_expiii_sub2$passage <= end_psg]))
#                     aa <- length(unique(variants_expiii_sub2$mut_info[variants_expiii_sub2$allele_freq >= 0.05 & variants_expiii_sub2$exp_line %in% c(line2) & variants_expiii_sub2$passage <= end_psg]))
#                     b <- length(unique(variants_expiii_sub2$mut_info[variants_expiii_sub2$allele_freq >= 0.99 & variants_expiii_sub2$exp_line %in% c(line1) & variants_expiii_sub2$passage == end_psg]))
#                     bb <- length(unique(variants_expiii_sub2$mut_info[variants_expiii_sub2$allele_freq >= 0.99 & variants_expiii_sub2$exp_line %in% c(line2) & variants_expiii_sub2$passage == end_psg]))
                    
#                     c <- length(intersect(variants_expiii_sub2$mut_info[variants_expiii_sub2$allele_freq >= 0.05 & variants_expiii_sub2$exp_line %in% c(line1) & variants_expiii_sub2$passage <= end_psg],variants_expiii_sub2$mut_info[variants_expiii_sub2$allele_freq >= 0.05 & variants_expiii_sub2$exp_line %in% c(line2) & variants_expiii_sub2$passage <= end_psg]))
#                     d <- length(intersect(unique(variants_expiii_sub2$mut_info[variants_expiii_sub2$allele_freq >= 0.99 & variants_expiii_sub2$exp_line %in% c(line1) & variants_expiii_sub2$passage == end_psg]),
#                     unique(variants_expiii_sub2$mut_info[variants_expiii_sub2$allele_freq >= 0.99 & variants_expiii_sub2$exp_line %in% c(line2) & variants_expiii_sub2$passage == end_psg])))

#                     i <- 0
                    
#                     exp_base <- integer()
#                     while (i <= 1000){
#                         i <- i + 1
#                         j <- sample(1:a, size = b, replace = F)
#                         k <- sample(1:aa, size = bb, replace = F)
#                         exp_base <- c(exp_base,sum(intersect(j,k)<=c))
#                     }

#                     exp_all <- rbind(exp_all, exp_base)
#                 }
                
#                 obs_all_vec <- c()
#                 obs_shared_vec <- c()

#                 exp_shared <- colSums(exp_all)
#                 obs_all_vec <- unique(append(obs_all_vec, unique(variants_expiii_sub$mut_info[variants_expiii_sub$allele_freq >= 0.99 & variants_expiii_sub$exp_line %in% c(line1,line2) & variants_expiii_sub$passage == end_psg])))
#                 obs_all <- length(obs_all_vec)
#                 obs_shared_vec <- unique(append(obs_shared_vec, intersect(variants_expiii_sub$mut_info[variants_expiii_sub$allele_freq >= 0.99 & variants_expiii_sub$exp_line %in% c(line1) & variants_expiii_sub$passage == end_psg],variants_expiii_sub$mut_info[variants_expiii_sub$allele_freq >= 0.99 & variants_expiii_sub$exp_line %in% c(line2) & variants_expiii_sub$passage == end_psg])))
#                 obs_shared <- length(obs_shared_vec)
                
                
#                 # Calculate mean and standard deviation
#                 mean_data <- mean(exp_shared)
#                 sd_data <- sd(exp_shared)
#                 # get CI
#                 result <- t.test(exp_shared, conf.level = 0.95)
#                 CIs <- as.numeric(result$conf.int)
#                 # Calculate the z-score
#                 z_score <- (obs_shared - mean_data) / sd_data
#                 # Calculate the p-value
#                 p_value <- 2 * (1 - pnorm(abs(z_score)))
#                 df <- rbind(df,c(line1,line2,end_psg,obs_all,obs_shared,obs_shared/obs_all,mean_data/obs_all,p_value, CIs[1]/obs_all, CIs[2]/obs_all))
#             }
#         }

#         df %<>% set_colnames(c("Var1", "Var2", "passage", "total_unique", "shared", "pct_shared", "expected_pct_shared", "p.val", "CI_lower", "CI_upper"))
#         write.table(df, file = paste0("/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/tables/misc/",names(effect),"_parallelism_temporal_simulation.tsv"), sep = "\t")
#     }
# }

```

```{r}
k <- 4
effect <- effect_list[k]
df <- read.table(file = paste0("/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/tables/misc/",names(effect),"_parallelism_temporal_simulation.tsv"), sep = "\t", header = T)

df_long <- df %>% 
    mutate(Pair = paste0(Var1, "vs", Var2)) %>% 
    filter((Pair %in% c("MT-4_1vsMT-4_2", "MT-2_1vsMT-2_2"))) %>%
    pivot_longer(cols = c(pct_shared, expected_pct_shared), names_to = "group", values_to =  "shared_all")

df_long %<>% 
    mutate_at(c(3,7,8,11), as.numeric) %>%
    mutate(shared_all = round(100*shared_all, digits = 2),
    CI_lower = round(100*CI_lower, digits = 2),
    CI_upper = round(100*CI_upper, digits = 2)) %>%
    mutate(group2 = case_when(group == "pct_shared" ~ "Observed", group == "expected_pct_shared" ~ "Randomized")) %>%
    dplyr::select(c(Pair, passage, group2, CI_lower,CI_upper, shared_all, total_unique)) %>%
    mutate_at(c(1), as.factor) %>%
    mutate(group2 = factor(group2, levels = c("Observed", "Randomized")))

df_long$group2 <- factor(df_long$group2, levels = c("Observed", "Randomized"))

# generate figure
fig <- df_long %>% 
    ggplot(aes(x = passage, y = shared_all, group = interaction(Pair,group2), color = Pair, linetype = group2)) +
    geom_line(size = 2) +
    # geom_smooth(method = "loess") +
    geom_ribbon(
        data = df_long %>% filter(group2 == "Randomized"),  # Only apply CI for the specified group
        aes(ymin = CI_lower, ymax = CI_upper, fill = Pair),
        alpha = 0.2  # Transparency of CI ribbon
    ) +
    scale_color_manual(name = "Pair",
    labels = c("MT-2 replicates", "MT-4 replicates"), 
    values = c("#ff8b08", "#00bbff")
    ) +
    scale_fill_manual(
        values = c("#ff8b08", "#00bbff"),
        guide = "none"  # Hides the legend for `fill`
    ) +
    labs(title = "", y = "Parallelism (%)", x = "Transfer") +
    scale_x_continuous(breaks = seq(0,500,50)) +
    scale_linetype_discrete(name = "Dataset",  # Change the legend title here
                    labels = c("Observed", "Randomized"))+
    custom_plot_theme(application = "publication")



# save figure
custom_plot_save(fig, "parallel_over_time.pdf", height = 14, width = 14)

```

```{r in text numbers}


df_long <- df_long[!is.na(df_long$shared_all),]

df_long[df_long$group2 == "Observed" & df_long$shared_all == max(df_long$shared_all[df_long$Pair == "MT-2_1vsMT-2_2"], na.rm = TRUE),]
df_long[df_long$group2 == "Observed" & df_long$Pair == "MT-2_1vsMT-2_2" & df_long$passage == 500,]

df_long[df_long$group2 == "Observed" & df_long$shared_all == max(df_long$shared_all[df_long$Pair == "MT-4_1vsMT-4_2"], na.rm = TRUE),]
df_long[df_long$group2 == "Observed" & df_long$Pair == "MT-4_1vsMT-4_2" & df_long$passage == 500,]


df_long[df_long$group2 == "Randomized" & df_long$shared_all == max(df_long$shared_all[df_long$group2 == "Randomized"& df_long$Pair == "MT-4_1vsMT-4_2"], na.rm = TRUE),]

```



```{r counts per bins}


t_2_f <- readRDS(paste0(wd, "results/tables/time_to_fixation.Rds"))

t_2_f <- t_2_f %>% mutate(host = case_when(exp_line %in% c("MT-2_1", "MT-2_2") ~ "MT-2", exp_line %in% c("MT-4_1", "MT-4_2") ~ "MT-4"),
                        group = case_when(Occurrence %in% c("1 out of 4") ~ "Line-specific", Occurrence %in% c("3 out of 4", "4 out of 4") ~ "Highly-parallel")) %>%
                        filter(!is.na(group))

t_2_f %>% filter(group == "Highly-parallel") %>% dplyr::select(c(mut_info, effect_simplified, Occurrence)) %>% distinct()


bins <- seq(0,500, 100)

nrow(t_2_f[t_2_f$host == "MT-2" & t_2_f$group == "Highly-parallel" & t_2_f$fixation <= 200,])/
nrow(t_2_f[t_2_f$host == "MT-2" & t_2_f$group == "Highly-parallel",])

nrow(t_2_f[t_2_f$host == "MT-2" & t_2_f$group == "Highly-parallel" & t_2_f$fixation >= 400,])/
nrow(t_2_f[t_2_f$host == "MT-2" & t_2_f$group == "Highly-parallel",])

nrow(t_2_f[t_2_f$host == "MT-4" & t_2_f$group == "Highly-parallel" & t_2_f$fixation <= 200,])/
nrow(t_2_f[t_2_f$host == "MT-4" & t_2_f$group == "Highly-parallel",])

nrow(t_2_f[t_2_f$host == "MT-4" & t_2_f$group == "Highly-parallel" & t_2_f$fixation >= 400,])/
nrow(t_2_f[t_2_f$host == "MT-4" & t_2_f$group == "Highly-parallel",])


t_2_f_summary <- t_2_f %>%
        mutate(group = factor(group, levels = c("Line-specific", "Highly-parallel")), 
            bin = cut(fixation, breaks = bins, left = FALSE, right = TRUE)) %>%
        group_by(group, exp_line, host) %>%
        count(bin, .drop = FALSE)


t_2_f_summary <- t_2_f_summary[!is.na(t_2_f_summary$bin),]
```


# per host
Do we observe a significantly different time trend between line-specific and highly-parallel mutations in each host?
- choice #1: time trend -> treat time (bin) as continuous variable (bin midpoint, scaled)
- choice #2: time trend BETWEEN mutation groups -> interaction term between time and mutation group
- choice #3: in each host -> perform two separate tests for each host
- choice #4: do not treat each replicate as a random effect




```{r making choices and fitting models}



# # generalized linear mixed effect model with exp_line as random effect with poisson distribution

# model_trend1 <- glmer(
#     n ~ bin_mid_scaled * group + (1 | exp_line),
#     data = t_2_f_summary[t_2_f_summary$host == "MT-2", ],
#     family = poisson(link = "log")
# )

# summary(model_trend1)
# Anova(model_trend1, type = 3)


# model_trend2 <- glmer(
#     n ~ bin_mid_scaled * group + (1 | exp_line),
#     data = t_2_f_summary[t_2_f_summary$host == "MT-4", ],
#     family = poisson(link = "log")
# )

# summary(model_trend2)
# Anova(model_trend2, type = 3)

# # generalized linear mixed effect model with host as random effect with NB distribution
# model_trend_nb1 <- glmer.nb(
#     n ~ bin_mid_scaled * group + (1 | exp_line),
#     data = t_2_f_summary[t_2_f_summary$host == "MT-2", ],
# )

# summary(model_trend_nb1)
# Anova(model_trend_nb1, type = 3)


# model_trend_nb2 <- glmer.nb(
#     n ~ bin_mid_scaled * group + (1 | exp_line),
#     data = t_2_f_summary[t_2_f_summary$host == "MT-4", ],
# )

# summary(model_trend_nb2)
# Anova(model_trend_nb2, type = 3)

# ## choice between poisson and NB

# overdisp_fun <- function(model) {
#   rdf <- df.residual(model)
#   rp <- residuals(model, type = "pearson")
#   Pearson.chisq <- sum(rp^2)
#   prat <- Pearson.chisq / rdf
#   pval <- pchisq(Pearson.chisq, df = rdf, lower.tail = FALSE)
#   c(chisq = Pearson.chisq, ratio = prat, rdf = rdf, p = pval)
# }

# overdisp_fun(model_trend1)
```


## Conclusions from comparing different models

for MT-2 model: negative binomial model better than Poisson (less AIC 112 compared to 122, finite dispersion parameter)
for MT-4 model: negative binomial model worse than Poisson (higher AIC 86 compared to 84, infinite dispersion parameter)
Verdict: use negative binomial model because it performs better in one model and is less stric in its assumptions compared to Poisson

Treating replicate as a random effect does not affect the model as there are 
only 2 levels and variance cannot be estimated (approximated to 0)!
AIC slightly better in model without random effect but not significantly (110 compared to 112)
Regard replicate as a random effect since it conceptually makes more sense!

```{r fitting the final model}

# treat bin as continuous (time trend test) and scale it for numerical stability improves
t_2_f_summary <- t_2_f_summary %>%
    mutate(bin_mid = as.numeric(sub("\\D*(\\d+),.*", "\\1", bin)) + 50) %>%
    mutate(bin_mid_scaled = scale(bin_mid, center = TRUE, scale = TRUE))


model_trend1 <- glmer.nb(
    n ~ bin_mid_scaled * group + (1 | exp_line),
    data = t_2_f_summary[t_2_f_summary$host == "MT-2", ]
)

summary(model_trend1)



model_trend2 <- glmer.nb(
    n ~ bin_mid_scaled * group + (1 | exp_line),
    data = t_2_f_summary[t_2_f_summary$host == "MT-4", ]
)

summary(model_trend2)


```


```{r summary table of model result and performance}


# Close any open graphics devices first
while (!is.null(dev.list())) {
  dev.off()
}
# Function to extract model information
extract_model_info <- function(model, model_name) {
  
  # Get fixed effects
  fixed_effects <- summary(model)$coefficients
  
  # Create data frame for fixed effects
  fixed_df <- data.frame(
    Model = model_name,
    Term = rownames(fixed_effects),
    Estimate = sprintf("%.4f", fixed_effects[, "Estimate"]),
    SE = sprintf("%.4f", fixed_effects[, "Std. Error"]),
    z_value = sprintf("%.3f", fixed_effects[, "z value"]),
    p_value = format.pval(fixed_effects[, "Pr(>|z|)"], digits = 3, eps = 0.001),
    Sig = case_when(
      fixed_effects[, "Pr(>|z|)"] < 0.001 ~ "***",
      fixed_effects[, "Pr(>|z|)"] < 0.01 ~ "**",
      fixed_effects[, "Pr(>|z|)"] < 0.05 ~ "*",
      fixed_effects[, "Pr(>|z|)"] < 0.1 ~ ".",
      TRUE ~ ""
    ),
    stringsAsFactors = FALSE
  )
  
  return(fixed_df)
}

# Function to extract model fit statistics
extract_fit_stats <- function(model, model_name) {
  summ <- summary(model)
  
  data.frame(
    Model = model_name,
    AIC = sprintf("%.1f", AIC(model)),
    BIC = sprintf("%.1f", BIC(model)),
    logLik = sprintf("%.1f", logLik(model)),
    Deviance = sprintf("%.1f", deviance(model)),
    N_obs = nobs(model),
    N_groups = summary(model)$ngrps,
    Theta = sprintf("%.4f", getME(model, "glmer.nb.theta")),
    stringsAsFactors = FALSE
  )
}

# Assuming your models are named model_trend1 and model_trend2
# Extract information from both models
fixed_effects_comparison <- bind_rows(
  extract_model_info(model_trend1, "MT-2"),
  extract_model_info(model_trend2, "MT-4")
)

fit_stats_comparison <- bind_rows(
  extract_fit_stats(model_trend1, "MT-2"),
  extract_fit_stats(model_trend2, "MT-4")
)

# Create the comparison tables
cat("\n=== FIXED EFFECTS COMPARISON ===\n\n")

fixed_table <- fixed_effects_comparison %>%
  kable("latex", 
        booktabs = TRUE,
        caption = "Fixed Effects Comparison: MT-2 vs MT-4 Models",
        col.names = c("Model", "Term", "Estimate", "SE", "z-value", "p-value", "Sig."),
        align = c("l", "l", "r", "r", "r", "r", "c")) %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"),
                font_size = 10) %>%
  column_spec(1, bold = TRUE) %>%
  pack_rows("MT-2 Model", 1, 4) %>%
  pack_rows("MT-4 Model", 5, 8) %>%
  footnote(general = "Significance codes: *** p<0.001, ** p<0.01, * p<0.05, . p<0.1",
          general_title = "",
          footnote_as_chunk = TRUE)

cat("\n=== MODEL FIT STATISTICS ===\n\n")

fit_table <- fit_stats_comparison %>%
  kable("latex",
        booktabs = TRUE,
        caption = "Model Fit Statistics Comparison",
        col.names = c("Model", "AIC", "BIC", "logLik", "Deviance", "N obs", "N groups", "Theta"),
        align = c("l", "r", "r", "r", "r", "r", "r", "r")) %>%
  kable_styling(latex_options = c("striped", "hold_position"),
                font_size = 10) %>%
  column_spec(1, bold = TRUE)

# Save to PDF
pdf("/Users/alimos313/Documents/studies/phd/research/manuscripts/decelerating-adaptation/reports/glmer_model_comparison.pdf", width = 11, height = 8.5)

# Create a more detailed comparison table
par(mar = c(1, 1, 1, 1))
plot.new()
text(0.5, 0.95, "GLMER Model Comparison: MT-2 vs MT-4", 
    cex = 1.8, font = 2)
text(0.5, 0.90, "Negative Binomial Mixed Effects Models", 
    cex = 1.2, font = 3)

# Create text summary
mt2_text <- paste0(
  "MT-2 Model:\n",
  "Formula: n ~ bin_mid_scaled * group + (1 | exp_line)\n",
  "Family: Negative Binomial (θ = 7.16)\n",
  "AIC: 112.0, BIC: 118.0\n",
  "Observations: 20, Groups: 2\n",
  "Convergence: OK (boundary/singular fit)\n\n",
  "Fixed Effects:\n",
  "  Intercept: 2.79 (SE=0.14, p<0.001***)\n",
  "  bin_mid_scaled: 0.45 (SE=0.18, p=0.013*)\n",
  "  groupHighly-parallel: -2.22 (SE=0.31, p<0.001***)\n",
  "  Interaction: -1.00 (SE=0.36, p=0.005**)"
)

mt4_text <- paste0(
  "MT-4 Model:\n",
  "Formula: n ~ bin_mid_scaled * group + (1 | exp_line)\n",
  "Family: Negative Binomial (θ = 106169.4)\n",
  "AIC: 86.1, BIC: 92.1\n",
  "Observations: 20, Groups: 2\n",
  "Convergence: OK (boundary/singular fit, iteration limit)\n\n",
  "Fixed Effects:\n",
  "  Intercept: 1.43 (SE=0.17, p<0.001***)\n",
  "  bin_mid_scaled: 0.64 (SE=0.18, p<0.001***)\n",
  "  groupHighly-parallel: -0.75 (SE=0.29, p=0.009**)\n",
  "  Interaction: -1.16 (SE=0.31, p<0.001***)"
)

text(0.05, 0.75, mt2_text, adj = c(0, 1), cex = 0.9, family = "mono")
text(0.05, 0.35, mt4_text, adj = c(0, 1), cex = 0.9, family = "mono")

text(0.5, 0.05, "Note: Both models show boundary/singular fits (random effect variance near zero)", 
    cex = 0.8, font = 3, col = "red")

dev.off()

cat("\nPDF saved as: glmer_model_comparison.pdf\n")


pdf("/Users/alimos313/Documents/studies/phd/research/manuscripts/decelerating-adaptation/reports/glmer_comparison_detailed.pdf", width = 11, height = 8.5)

# Title
grid.text("GLMER Model Summary for MT-2 and MT-4", 
          x = 0.5, y = 0.95, 
          gp = gpar(fontsize = 18, fontface = "bold"))

grid.text("Negative Binomial Mixed Effects Models", 
          x = 0.5, y = 0.91, 
          gp = gpar(fontsize = 12, fontface = "italic"))

# Fixed effects table
pushViewport(viewport(x = 0.5, y = 0.60, width = 0.9, height = 0.5))
grid.table(fixed_effects_comparison, rows = NULL)
popViewport()

# Fit statistics table
pushViewport(viewport(x = 0.5, y = 0.15, width = 0.9, height = 0.2))
grid.table(fit_stats_comparison, rows = NULL)
popViewport()

# Footer note
grid.text("Note: Both models exhibit boundary/singular fits (random effect variance ≈ 0)", 
          x = 0.5, y = 0.02, 
          gp = gpar(fontsize = 9, fontface = "italic", col = "red"))

dev.off()

cat("\nDetailed PDF saved as: glmer_comparison_detailed.pdf\n")

# Print tables to console
print("=== FIXED EFFECTS ===")
print(fixed_effects_comparison)
print("\n=== FIT STATISTICS ===")
print(fit_stats_comparison)

```



```{r barplot with predictions}

# Get predictions with standard errors
pred_mt2 <- predict(model_trend1, 
                    newdata = t_2_f_summary[t_2_f_summary$host == "MT-2", ],
                    type = "link", se.fit = TRUE)

pred_mt4 <- predict(model_trend2, 
                    newdata = t_2_f_summary[t_2_f_summary$host == "MT-4", ],
                    type = "link", se.fit = TRUE)

# Add to dataframe
t_2_f_summary$fit_link <- NA
t_2_f_summary$se_link <- NA

t_2_f_summary$fit_link[t_2_f_summary$host == "MT-2"] <- pred_mt2$fit
t_2_f_summary$se_link[t_2_f_summary$host == "MT-2"] <- pred_mt2$se.fit

t_2_f_summary$fit_link[t_2_f_summary$host == "MT-4"] <- pred_mt4$fit
t_2_f_summary$se_link[t_2_f_summary$host == "MT-4"] <- pred_mt4$se.fit

# Compute 95% CI on the response scale
t_2_f_summary <- t_2_f_summary %>%
    mutate(pred = exp(fit_link),
         pred_low = exp(fit_link - 1.96 * se_link),
         pred_high = exp(fit_link + 1.96 * se_link))

# Plot observed bars + predicted trends
fig_pred <- ggplot(t_2_f_summary, aes(x = bin, y = n, fill = exp_line)) +
    facet_grid(group ~ host, scales = "free_y") +
    geom_bar(stat = "identity", color = "white") +
    geom_ribbon(aes(ymin = pred_low, ymax = pred_high, fill = "grey", group = group),
            alpha = 0.5) +
    geom_line(aes(y = pred, color = host, group = group), size = 3) +
    # Dashed lines for confidence intervals
    geom_line(aes(y = pred_low, color = host, group = group),
                linetype = "dashed", size = 1.5) +
    geom_line(aes(y = pred_high, color = host, group = group),
                linetype = "dashed", size = 1.5) +
    facetted_pos_scales(
        y = list(
        group == "Highly-parallel" ~ scale_y_continuous(breaks = seq(0, 7, 1)),
        group == "Line-specific"   ~ scale_y_continuous(breaks = seq(0, 60, 5))
        )
    ) +
    scale_fill_manual(labels = exp_line_factor, values = line_col_palette) +
    scale_color_manual(values = c("MT-2" = "#E69F00", "MT-4" = "#56B4E9")) +
    labs(x = "Fixation period", y = "Count", title = "Observed and predicted fixation counts") +
    custom_plot_theme(scale = 1.5) +
    theme(axis.text.x = element_text(angle = 90),
            legend.position = "none")

# save figure
custom_plot_save(fig_pred, "highly-parallel_vs_line-specific_time-trend_glm.pdf", height = 31.5, width = 21)

```


####!!!!
Extra analysis
####!!!!

# per whole experiment

Do we observe a significantly different time trend between line-specific and highly-parallel mutations in the experiment?
- choice #1: ditto
- choice #2: ditto
- choice #3: in the experiment -> perform a single test for whole experiment
- choice #4: treat each experiment line as a random effect 

```{r}


fig <- ggplot(t_2_f_summary, aes(x = bin, y = n, fill = exp_line)) +
    facet_grid(~group, scales = "free_y") +
    geom_bar(stat = "identity", position = "stack", color = "white") +
    scale_fill_manual(labels = exp_line_factor, values = line_col_palette) +
    labs(title = "", x = "\nFixation period", y = "Count") +
    custom_plot_theme() +
    theme(axis.text.x = element_text(angle = 90),
        legend.position = "none")



```

```{r}



# generalized linear mixed effect model with host as random effect with NB distribution
model_trend_nb22 <- glmer.nb(
    n ~ bin_mid_scaled * group + (1 | exp_line),
    data = t_2_f_summary)

summary(model_trend_nb22)

ggplot(t_2_f_summary,
        aes(x = bin_mid, y = n, color = group)) +
    geom_point() +
    geom_smooth(method = "glm", method.args = list(family = "poisson"), se = TRUE) +
    theme_bw()

```





