---
title:  
    "Overview of HIV-1 LTEE genomic data (EXPIV)"
date: 
    "Oct 3th, 2023"
author:  
    "Ali Movasati"  
output:  
    html_document:
        #css: style.css
        toc: true
        toc_depth: 3
        toc_float: true
        number_sections: true
        theme: lumen
        highlight: zenburn
        code_folding: hide
---



```{r setting global chunk parameters, include=FALSE}

# options for code chunk and output inclusion 
knitr::opts_chunk$set(
  message = FALSE, echo = TRUE, warning=FALSE, results = "hide"
)

# options for figures
knitr::opts_chunk$set(
  fig.width = 8, fig.height = 8, fig.align = "center"
)

# options(scipen=999)
# margin of plotly
m <- list(
  l = 50,
  r = 50,
  b = 100,
  t = 100,
  pad = 4
)


```



# Introduction

In this document I will analyze the data for **EXPIV** of HIV-1 LTEE dataset. The data of this experiment has not been previously analyzed. This document will only 
cover the basic overview of the dataset, and in a new document I will compare the results from EXPIV to that of expIII.

EXPIV dataset includes genomic sequencing data of HIV-1 NL4-3 passages up to 510 (lines 17-20). There are four independent lines that have been passaged in MT2 T-cell lines.

In this experiment each viral transfer was done through a titer bottlenecking of the virus.



```{r loading libraries, echo=FALSE}

# loading libraries
library("ggplot2")
library("plotly")
library("DT")
library("dplyr")
library("tidyr")
library("Biostrings")
library("stringr")
library("magrittr")

```



```{r read-in the data}


variants_ann <- read.table(file = "/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/tables/pipeline-outputs/all_annotated_variants.tsv.gz", 
sep = "\t", stringsAsFactors = FALSE, header = TRUE)


# set correct data types
variants_ann$exp_line <- as.integer(variants_ann$exp_line)


# set host cell line info
variants_ann$host_line <- ifelse(variants_ann$exp_line %in% c("13", "14","17", "18", "19", "20"), "MT2", "MT4")


# account for the switched experimental lines


variants_ann$exp_line[variants_ann$exp_line == 18 & variants_ann$passage > 350] <- "A"
variants_ann$exp_line[variants_ann$exp_line == 19 & variants_ann$passage > 350] <- "B"

variants_ann$exp_line[variants_ann$exp_line == "A"] <- "19"
variants_ann$exp_line[variants_ann$exp_line == "B"] <- "18"

# add a field containing pos, ref, and alt info of muts
variants_ann$mut_info <- paste(variants_ann$genomic_pos, variants_ann$ref_allele, variants_ann$alt_allele, sep = "_")


## only keep the EXPIV data

variants_ann_expiv <- variants_ann[variants_ann$exp_line %in% 17:20,]
end_passage <- 630
variants_ann_expiv <- variants_ann_expiv %>%
                              mutate(exp_line = case_when(exp_line==17 ~ "MT2_1", exp_line==18 ~ "MT2_2", exp_line==19 ~ "MT2_3", exp_line==20 ~ "MT2_4"))

variants_expiv <- variants_ann_expiv %>% filter(!duplicated(variant_id))

```







# Overview

## Minority Mutations

```{r number of min mutations}

exp <- "EXPIV"
category <- "Minority"
min_threshold <- 0.05
max_threshold <- 0.50

# get numbers for in-text reporting
variants_expiv_subset <- variants_expiv %>%
                                filter(allele_freq >= min_threshold & allele_freq <= max_threshold)

a <- nrow(variants_expiv_subset)
b <- length(unique(variants_expiv_subset$mut_info))


```

We observe a total of `r a` `r category` mutations across the four evolution lines of EXP III (`r b` unique mutations).

> `r category` mutations are the mutations that became reliably detectable in the population (a.k.a, allele frequency above `r threshold`)!



```{r timewise min mut accumulation, results=TRUE}

# prepare data for visualization
counts <- variants_expiv %>%
              filter(allele_freq >= min_threshold & allele_freq < max_threshold) %>%
              group_by(exp_line, passage) %>%
              count() %>%
              dplyr::rename(mut_count = n) %>%
              mutate_at(1, as.factor) %>%
              mutate_at(2:3, as.numeric)

# visualize using plotly


fig <- plot_ly(as.data.frame(counts), x = ~passage, y = ~mut_count, 
            type = 'scatter', mode = "lines", color = ~exp_line)
fig <- fig %>% layout(title = paste0(category, " Mutations Accumulation"),
                      xaxis = list(title = 'Passage'),
                      yaxis = list(title = 'Count'))

fig


```


The mutations spectra of all 4 lines look similar for minority mutations!


```{r min mutation spectrum, results=TRUE}

# prepare data for visualization
counts <- variants_expiv %>%
            filter(allele_freq >= threshold) %>%
            filter(mut_type=="P") %>%
            mutate(mut_subtype=paste0(ref_allele, ">", alt_allele)) %>%
            group_by(exp_line, passage, mut_subtype) %>%
            count() %>%
            dplyr::rename(count=n) %>%
            group_by(exp_line,passage) %>%
            mutate(pct =  100 *count/sum(count)) %>%
            ungroup() %>%
            mutate_at(2:3, as.factor)


# visualize using ggplot
fig <- as.data.frame(counts) %>% ggplot(aes(x=passage, y = pct, fill = mut_subtype)) +
                                  geom_bar(position="stack", stat="identity") + 
                                  facet_wrap(~exp_line) +
                                  labs(x="Passage", y= "Percentage", fill = "Base Substitution")+
                                  theme_bw() +
                                  theme(axis.title = element_text(size = 20),
                                  axis.text.x = element_blank(),
                                  axis.text.y = element_text(size = 10),
                                  strip.text.x = element_text(size = 25),
                                  legend.text=element_text(size=10),
                                  legend.title = element_text(size=15))


fig


png("/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/figs/presentation/270923/spectrum-iv.png"
    , width = 1600, height = 1200)
fig
dev.off()


```



## Majority Mutations

```{r number of maj mutations}


category <- "Majority"
threshold <- 0.5

# get numbers for in-text reporting
variants_expiv_subset <- variants_expiv %>%
                                filter(allele_freq >= threshold)


a <- nrow(variants_expiv_subset[variants_expiv_subset$passage == end_passage,])
b <- length(unique(variants_expiv_subset$mut_info[variants_expiv_subset$passage == end_passage]))

```

We observe a total of `r a` `r category` mutations across the four evolution lines of EXP III (`r b` unique mutations).

> `r category` mutations are the mutations that became more frequent in the population than the wild-type nucleotide (a.k.a, allele frequency above `r threshold`)!



```{r time-wise maj mut accumulation, results=TRUE}


# prepare data for visualization
counts <- variants_expiv %>%
              filter(allele_freq >= threshold) %>%
              group_by(exp_line, passage) %>%
              count() %>%
              dplyr::rename(mut_count = n) %>%
              mutate_at(1, as.factor) %>%
              mutate_at(2:3, as.numeric)

# visualize using plotly
fig <- plot_ly(as.data.frame(counts), x = ~passage, y = ~mut_count, 
            type = 'scatter', mode = "lines", color = ~exp_line)
fig <- fig %>% layout(title = paste0(category, " Mutations Accumulation") ,
                      xaxis = list(title = 'Passage'),
                      yaxis = list(title = 'Count'))

fig


```


___
___

We can also plot the number of new `r category` mutations that appear or disappear compared to the immediate previous passage.

```{r prepare difference dataframe}


passages <- seq(10,510, 10)
diff_df <- data.frame()


for (i in 1:50){
    for (exp_line in c("MT2_1", "MT2_2", "MT2_3", "MT2_4")){
        prev_in <- passages[i]
        pres_in <- passages[i+1]

        prev <- variants_expiv$mut_info[variants_expiv$exp_line == exp_line & variants_expiv$passage == prev_in & variants_expiv$allele_freq >= 0.5]

        pres <- variants_expiv$mut_info[variants_expiv$exp_line == exp_line & variants_expiv$passage == pres_in & variants_expiv$allele_freq >= 0.5]

        addition <- length(setdiff(pres, prev))
        removal <- length(setdiff(prev,pres))

        diff_df <- rbind(diff_df, c(exp_line, passages[i+1], addition, removal))

    }
    
}

diff_df %<>% set_colnames(c("exp_line", "passage", "addition", "removal")) %>%
              arrange(exp_line) %>%
              mutate_at(1, as.factor) %>%
              mutate_at(2:4, as.numeric)


```



In this experiment the average rate of mutation emergence across all four lines is `r mean(diff_df$addition[!(diff_df$exp_line == "MT2_3" & diff_df$passage == 440)])` while the average rate of 
mutation disappearance is `r mean(diff_df$removal[!(diff_df$exp_line == "MT2_3" & diff_df$passage == 430)])`.

```{r plot addition plot, results=TRUE}



# visualize using plotly
fig <- plot_ly(diff_df, x = ~passage, y = ~addition, 
            type = 'scatter', mode = "lines", color = ~exp_line)
fig <- fig %>% layout(title = paste0(category, " Mutations Emergence") ,
                      xaxis = list(title = 'Passage'),
                      yaxis = list(title = 'Count'))

fig


```



```{r plot removal plot, results=TRUE}



# visualize using plotly
fig <- plot_ly(as.data.frame(diff_df), x = ~passage, y = ~removal, 
            type = 'scatter', mode = "lines", color = ~exp_line)
fig <- fig %>% layout(title = paste0(category, " Mutations Disappearance") ,
                      xaxis = list(title = 'Passage'),
                      yaxis = list(title = 'Count'))

fig


```





___
___


If we plot the accumulation of `r category` mutations based on the effect type, we can visualize
the trends for S (synonymous), N (non-synonymous), and U (untranslated) mutations.

```{r mut effect plot, results=TRUE}


# prepare data for visualization
counts <- variants_ann_expiv%>% 
              filter(allele_freq >= threshold) %>%
              mutate(effect_simplified = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
              mutate(effect_simplified = replace_na(effect_simplified, "N")) %>%
              group_by(exp_line, passage, effect_simplified) %>%
              summarise(nr_majority = n()) %>%
              mutate_at(c(2,4), as.numeric) %>%
              mutate_at(c(1,3), as.factor)


# visualize using plotly
fig <- plot_ly(as.data.frame(counts), x = ~passage, y = ~nr_majority, 
              #text = ~paste("Car: ", car),
              type = 'scatter', mode = "lines", color = ~effect_simplified, linetype= ~exp_line)
fig <- fig %>% layout(title = paste0(category, " Mutation Accumulation Based on Translational Effect") ,
                      xaxis = list(title = 'Passage'),
                      yaxis = list(title = 'Count'))

fig


```


Alternatively, we can plot the same accumulation for different effect types but corrected for the 
genomic length that each mutation can occur in. Unlike EXPIII, in EXPIV HIV-1 genomes we see that density of
synonymous and non-synonymous Majority mutations are higher than mutations in untranslated regions, especially towards the end of experiment.



```{r statistics of maj muts in genomic features}


# prepare data for visualization
counts_df <- variants_ann_expiv%>% 
              filter(allele_freq >= threshold) %>%
              mutate(effect_simplified = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
              mutate(effect_simplified = replace_na(effect_simplified, "N"))


u <- c()
s_n <- c()

for (passage in seq(10, 510, by=10)){
counts <- table(counts_df$effect_simplified[counts_df$passage == passage])
u <- append(u, as.numeric(counts["U"]))
s_n <- append(s_n, sum(as.numeric(counts["S"]),as.numeric(counts["N"])))
}
# normalize for length
u <- u*100/(9626-9076+789-454)
s_n <- s_n*100/(9407-790)
t_test_result <- t.test(u, s_n, paired = TRUE, alternative = "two.sided")



```



Of note, the higher frequency of `r category` mutations in 
protein-coding regions compared to LTRs of the HIV-1 genome is statistically significant 
(p.value of `r format(t_test_result$p.val, scientific = TRUE)`).


```{r, results=TRUE}

# prepare data for visualization
counts <- variants_ann_expiv %>% 
            filter(allele_freq >= threshold) %>%
            mutate(effect_simplified = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
            mutate(effect_simplified = replace_na(effect_simplified, "N")) %>%
            mutate(effect_simplified2 = ifelse(effect_simplified %in% c("S","N"), "S/N", "U")) %>%
            group_by(exp_line, passage, effect_simplified2) %>%
            summarise(nr_majority = n()) %>%
            mutate_at(c(2,4), as.numeric) %>%
            mutate_at(c(1,3), as.factor) %>%
            mutate(nr_majority_density = ifelse(effect_simplified2 == "S/N", nr_majority*100/(9407-790),nr_majority*100/(9626-9076+789-454)))

# make the visualization
fig <- plot_ly(as.data.frame(counts), x = ~passage, y = ~nr_majority_density, 
              #text = ~paste("Car: ", car),
              type = 'scatter', mode = "lines", color = ~effect_simplified2, linetype= ~exp_line,
              colors = "Set1")
fig <- fig %>% layout(title = paste0("Normalized ", category, " Mutation Accumulation Based on Translational Effect") ,
                      xaxis = list(title = 'Passage'),
                      yaxis = list(title = 'Normalized Count'))

fig
```



___
___


Next we take a look at the number of majority mutations in different features of the HIV-1 genome:


```{r plot majority mut counts per feature, results=TRUE}


# prepare data for visualization
counts <- variants_ann_expiv %>% 
            filter(allele_freq >= threshold) %>%
            mutate(mut_type = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
            mutate(mut_type = replace_na(mut_type, "N")) %>%
            filter(passage == end_passage) %>%
            group_by(exp_line, feature, mut_type) %>%
            summarise(count = n()) %>%
            mutate_at(c(1,2,3), as.factor)



# make the visualization
fig <- ggplot(counts) +
  geom_bar(aes(x = mut_type, y = count, fill = exp_line),
           position = "stack",
           stat = "identity") +
  labs(x = "Genomic Features", y = "Count", fill = "Line")+
  facet_grid(~ feature, switch = "x") +
  theme(strip.placement = "outside",
        strip.background = element_rect(fill = NA, color = "white"),
        panel.spacing = unit(-.01,"cm")) +
  theme(axis.title = element_text(size = 20),
  axis.text.x = element_text(size = 10),
  axis.text.y = element_text(size = 10),
  strip.text.x = element_text(size = 15, angle=90),
  legend.text=element_text(size=10),
  legend.title = element_text(size=15))

fig
```


Looking at the frequency of different mutations types in the genomic regions (normalized
for the length of the genome):


```{r plot normalized majority mut counts per feature, results=TRUE}


# based on feature range
feature_length <- data.frame(feature = c("5UTR", "gag", "pol", "vif","vpr", "tat","rev","vpu","envelope","nef","3UTR"),
                            feature_length = c(335,1503,3012,579,291,261,351,246,2565,621,550))


# prepare data for visualization
counts <- variants_ann_expiv %>% 
            filter(allele_freq >= threshold) %>%
            mutate(mut_type = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
            mutate(mut_type = replace_na(mut_type, "N")) %>%
            filter(passage == end_passage) %>%
            group_by(exp_line, feature, mut_type) %>%
            summarise(count = n()) %>%
            mutate_at(c(1,2,3), as.factor) %>%
            left_join(feature_length, by = join_by(feature == feature)) %>%
            mutate(normalized_count = count*100/feature_length)


# make the visualization
fig <- ggplot(counts) +
  geom_bar(aes(x = mut_type, y = normalized_count, fill = exp_line),
           position = "stack",
           stat = "identity") +
  labs(x = "Genomic Features", y = "Normalized Count", fill = "Line")+
  facet_grid(~ feature, switch = "x") +
  theme(strip.placement = "outside",
        strip.background = element_rect(fill = NA, color = "white"),
        panel.spacing = unit(-.01,"cm")) +
  theme(axis.title = element_text(size = 20),
  axis.text.x = element_text(size = 10),
  axis.text.y = element_text(size = 10),
  strip.text.x = element_text(size = 15, angle=90),
  legend.text=element_text(size=10),
  legend.title = element_text(size=15))

fig
```

```{r, include = FALSE}

png("/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/figs/presentation/270923/fearure-iv.png"
    , width = 1600, height = 1200)
fig
dev.off()


```



Additionally we can look at the normalized count of mutations per genomic features for previous passages
to get a sense for whether and how they are changing over time. In the following you can view the plots for
passages 100, 200, 300, and 400. (for a better visualization view the pdf version)



```{r plot normalized majority mut counts per feature for past passages, results=TRUE}


# based on feature range
feature_length <- data.frame(feature = c("5UTR", "gag", "pol", "vif","vpr", "tat","rev","vpu","envelope","nef","3UTR"),
                            feature_length = c(335,1503,3012,579,291,261,351,246,2565,621,550))


# prepare data for visualization
counts <- variants_ann_expiv %>% 
            filter(allele_freq >= threshold) %>%
            mutate(mut_type = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
            mutate(mut_type = replace_na(mut_type, "N")) %>%
            filter(passage %in% c(100,200,300,400, 510)) %>%
            group_by(exp_line, passage, feature, mut_type) %>%
            summarise(count = n()) %>%
            mutate_at(1:4, as.factor) %>%
            left_join(feature_length, by = join_by(feature == feature)) %>%
            mutate(normalized_count = count*100/feature_length)


# make the visualization
fig <- ggplot(counts) +
  geom_bar(aes(x = mut_type, y = normalized_count, fill = exp_line),
           position = "stack",
           stat = "identity") +
  labs(x = "Genomic Features", y = "Normalized Count", fill = "Line")+
  facet_grid(~ feature + passage, switch = "x") +
  theme(strip.placement = "outside",
        strip.background = element_rect(fill = NA, color = "white"),
        panel.spacing = unit(-.01,"cm"))+
  theme(axis.title = element_text(size = 20),
  axis.text.x = element_text(size = 10),
  axis.text.y = element_text(size = 10),
  strip.text.x = element_text(size = 15, angle=90),
  legend.text=element_text(size=15),
  legend.title = element_text(size=10))

fig
```



```{r, include = FALSE}

pdf("/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/results/figs/pdf/fearure-count-history-iv.pdf"
    , width = 24.5, height = 14)
fig
dev.off()


```







## Mutational profile



In the charts below you can see the mutation subtypes of all/unique majority and 
minority mutations of all 4 lines of EXPIV.


```{r mut_profile pie-chart, results=TRUE}


# add subtype info
variants_expiv_sub <-  variants_expiv %>% mutate(mut_type_sub = ifelse(mut_type == "P", paste0(ref_allele,">",alt_allele), mut_type))


# minority

subset_name <- "Minority"
threshold <- 0.05


# prepare data for visualization (unique mutations)
mut_profile_min_unique <- variants_expiv_sub %>% filter(allele_freq >= threshold) %>%
                            select(mut_info, mut_type_sub) %>%
                            distinct() %>%
                            group_by(mut_type_sub) %>%
                            summarise(count = n())



# prepare data for visualization (all mutations)
mut_profile_min_all <- variants_expiv_sub %>% filter(allele_freq >= threshold) %>%
                            group_by(mut_type_sub) %>%
                            summarise(count = n())



# make the visualization
fig <- plot_ly()
fig <- fig %>% add_pie(data = mut_profile_min_unique, labels = ~mut_type_sub, values = ~count, title = paste0("Unique Mutations", " (", sum(mut_profile_min_unique$count),")"),
                       name = "Cut", domain = list(x = c(0, 0.4), y = c(0.4, 1.4)))
fig <- fig %>% add_pie(data = mut_profile_min_all, labels = ~mut_type_sub, values = ~count, title = paste0("All Mutations", " (", sum(mut_profile_min_all$count),")"),
                       name = "Color", domain = list(x = c(0.6, 1), y = c(0.4, 1.4))) %>%
                       layout(title=paste0('Mutation Sub-Types _ ', subset_name))
fig



# majority

subset_name <- "Majority"
threshold <- 0.5

# prepare data for visualization
mut_profile_maj_unique <- variants_expiv_sub %>% filter(allele_freq >= threshold) %>%
                            select(mut_info, mut_type_sub) %>%
                            distinct() %>%
                            group_by(mut_type_sub) %>%
                            summarise(count = n())
#                            rename(mut_sub_type = mut_type_sub)


# prepare data for visualization
mut_profile_maj_all <- variants_expiv_sub %>% filter(allele_freq >= threshold) %>%
                            group_by(mut_type_sub) %>%
                            summarise(count = n())
#                            rename(mut_sub_type = mut_type_sub)

# make the visualization
fig <- plot_ly()
fig <- fig %>% add_pie(data = mut_profile_maj_unique, labels = ~mut_type_sub, values = ~count, title = paste0("All Mutations", " (", sum(mut_profile_maj_all$count),")"),
                       name = "Cut", domain = list(x = c(0, 0.4), y = c(0.4, 1)))
fig <- fig %>% add_pie(data = mut_profile_maj_all, labels = ~mut_type_sub, values = ~count, title = paste0("Unique Mutations", " (", sum(mut_profile_maj_unique$count),")"),
                       name = "Color", domain = list(x = c(0.6, 1), y = c(0.4, 1))) %>%
                       layout(title=paste0('Mutation Sub-Types _ ', subset_name))
fig

```


```{r genome composition}

# read in ancestral reference sequence 
fastaFile <- readDNAStringSet("/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/data/reference/plasmid/hiv_plasmid_ref_genome.fasta")
seq_name = names(fastaFile)
sequence = paste(fastaFile)

# get the guanine composition of the reference genome
g_percent <- str_count(sequence, pattern = "G")/nchar(sequence)
```
<!--
As it is evident almost `r #mut_profile_min$count[mut_profile_min$mut_sub_type == "G>A"]/nrow(variants_ann_expiii_min)`% of minority and 
`r #mut_profile_maj$count[mut_profile_maj$mut_sub_type == "G>A"]/nrow(variants_ann_expiii_maj)`% of majority mutations are "G>A" mutations. This is while Guanine constitutes 
`r #g_percent`% of HIV-1 genome.

-->