---
title:  
    "Overview of mutation composition of EXPIV"
date: 
    "May 16th, 2024"
author:  
    "Ali Movasati"  
output:  
    html_document:
        #css: style.css
        toc: true
        toc_depth: 2
        toc_float: true
        number_sections: true
        theme: lumen
        highlight: zenburn
        code_folding: hide
        code_download: true # to make the Rmd source code available
---


```{css, echo=F}

#TOC {
  position: fixed;
  left: 0;
  top: 0;
  width: 250px;
  height: 100%;
  overflow:auto;
}

body, td {
  font-family: sans-serif;
  background-color: white;
  font-size: 16px;
}


h1, .h1, h2, .h2, h3, .h3 {
    margin-top: 30px;
}


p.caption {
  font-size: 0.9em;
  font-style: italic;
  color: grey;
  margin-right: 10%;
  margin-left: 10%;  
  text-align: justify;
}


/* scrollable code chunks, limit code chunk window height */

pre {
  max-height: 600px;
  overflow-y: auto;
}

pre[class] {
  max-height: 300px;
}


/* scrollable output of code chunks */
/* add class.output="scroll-100" to code chunks */

.scroll-100 {
  max-height: 300px;
  overflow-y: auto;
  background-color: inherit;
}
```



```{r setting global code chunk parameters, include=FALSE}

# options for code chunk and output display 
knitr::opts_chunk$set(
  message = FALSE, echo = TRUE, warning=FALSE, results = "hide"
)

# options for figure display
knitr::opts_chunk$set(
  fig.width = 8, fig.height = 8, fig.align = "center"
)


# options(scipen=999)

# margin of plotly
m <- list(
  l = 150,
  r = 150,
  b = 100,
  t = 100,
  pad = 4
)



```




# Introduction

In this document I will characterize the mutational profile that we observe in EXPIV evolutionary lines. One part of the result section of the manuscript will be dedicated to mutation statistics. I will try
to compose those paragraphs in this document using in-text coding feature of R Markdown. My goal is to later copy-paste the whole paragraphs from this document to the manuscript.


```{r loading libraries, echo=FALSE}

# loading libraries
library("ggplot2")
library("plotly")
library("gganimate")
library("DT")
library("dplyr")
library("tidyr")
library("Biostrings")
library("stringr")
library("magrittr")
library("hrbrthemes")

```


```{r functions and operators}
`%notin%` <- Negate(`%in%`)
```


```{r read-in the data}

# determine the working directory path based on server

if (file.exists("/home/amovas/")){
  print("Remote HPC Connection!")
  wd <- "/home/amovas/data/genome-evo-proj/"
} else {
  print("Local PC Connection!")
  wd <- "/Users/alimos313/Documents/studies/phd/hpc-research/genome-evo-proj/"
}

variants_ann <- read.table(file = paste0(wd, "results/tables/pipeline-outputs/all_annotated_variants.tsv.gz"), 
sep = "\t", stringsAsFactors = FALSE, header = TRUE)

head(variants_ann)
# only keep the EXPIII data
exp <- "EXPIV"
variants_ann_expiv <- variants_ann[variants_ann$exp_line %in% 17:20,]


# account for the switched experimental lines

variants_ann_expiv$exp_line[variants_ann_expiv$exp_line == 18 & variants_ann_expiv$passage > 350] <- "A"
variants_ann_expiv$exp_line[variants_ann_expiv$exp_line == 19 & variants_ann_expiv$passage > 350] <- "B"

variants_ann_expiv$exp_line[variants_ann_expiv$exp_line == "A"] <- "19"
variants_ann_expiv$exp_line[variants_ann_expiv$exp_line == "B"] <- "18"




# facors

host_factor <- "MT2"
exp_line_factor <- c("MT-2_i", "MT-2_ii", "MT-2_iii", "MT-2_iv")

translational_impact_factor <- c("U", "S", "N")
gene_factor <- c("gag", "pol", "vif", "vpr", "tat", "rev", "vpu", "envelope", "nef")
feature_factor <- c("5utr", "gag", "pol", "vif", "vpr", "tat1", "rev1", "vpu", "envelope", "tat2", "rev2", "nef", "3utr")


# format the dataframe
variants_ann_expiv <- variants_ann_expiv %>%
                            mutate(host_line = "MT2", # set host cell line info
                                exp_line = case_when(exp_line==17 ~ "MT-2_i", exp_line==18 ~ "MT-2_ii", exp_line==19 ~ "MT-2_iii", exp_line==20 ~ "MT-2_iv"), # update exp line labels
                                mut_info = paste(exp_line, genomic_pos, ref_allele, alt_allele, sep = "_")) %>% 
                            mutate(effect_simplified = case_when(effect == "synonymous_variant" ~ "S", effect == "untranslated_region" ~ "U")) %>%
                            mutate(effect_simplified = replace_na(effect_simplified, "N")) %>%
                            mutate(exp_line = factor(exp_line, levels = exp_line_factor),
                            host_line = factor(host_line, levels = host_factor),
                            effect_simplified = factor(effect_simplified, levels = translational_impact_factor)) 

# remove duplicate records for the same variants
variants_expiv <- variants_ann_expiv %>% filter(!duplicated(variant_id))


# last passage 

end_passage <- 510


# Visual settings


if (exp == "EXPIII"){
    line_palette <- c("#ff00ff", "#ff2400", "#6600cc", "#0000ff")
    breaks_step <- 50
    line_labels <- c("MT-2_1", "MT-2_2", "MT-4_1", "MT-4_2")
} else if ((exp == "EXPIV")) {
    line_palette <- c("#66ff66", "#009900", "#00cc99", "#558000")
    breaks_step <- 5
    line_labels <- c("MT-2_i", "MT-2_ii", "MT-2_iii", "MT-2_iv")
}


```


## Allele Frequency distribution

Let's first take a look at the allele frequency distribution of all mutations.



```{r, results=TRUE}


gg_fig <- ggplot(variants_expiv[variants_expiv$allele_freq >= 0.01,], aes(x=allele_freq)) + 
    geom_histogram(aes(y=..count..),      # Histogram with density instead of count on y-axis
                binwidth=.01,
                colour="black", fill="white") +
    geom_density(aes(y = ..count.. * .01), alpha=.3, fill="#FF6666") +
    facet_wrap(~exp_line, ncol = 2) +
    transition_states(passage, transition_length = 1, state_length = 1, wrap = TRUE) + # to create animation 
    labs(title = "Allele Freq. Distribution \n(Passage : {closest_state})", x = "Allele Freq.", y = "Count") +
    ylim(y = c(0,200))+
    theme_bw() +
    theme(
    plot.title = element_text(size = 25, hjust = 0.5),
    axis.title = element_text(size = 20),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    strip.text.x = element_text(size = 25),
    legend.title = element_text(size=15),
    legend.text = element_text(size = 10),
    plot.margin = unit(c(1,1,1,1), "cm"))



gg_fig

if (!file.exists(paste0(wd, "results/figs/gif/", exp, "-allele-freq-accumulation.gif"))){
  anim_save(gg_fig, file = paste0(wd, "results/figs/gif/", exp, "-allele-freq-accumulation.gif"))
}


# in-line variables
a <- 100*nrow(variants_expiii[variants_expiii$allele_freq >= 0.01 & variants_expiii$allele_freq < 0.05 & variants_expiii$passage == end_passage,])/
nrow(variants_expiii[variants_expiii$allele_freq >= 0.01 & variants_expiii$passage == end_passage,])

b <- 100*nrow(variants_expiii[variants_expiii$allele_freq >= 0.95 & variants_expiii$passage == end_passage,])/
nrow(variants_expiii[variants_expiii$allele_freq >= 0.01 & variants_expiii$passage == end_passage,])


```
